### 2014-01-15 text1.tex
ms text source

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or epsБ•√ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}% http://ctan.org/pkg/multirow
\usepackage{tablefootnote}
\usepackage{import}
\usepackage{subcaption}
\usepackage{caption}

\title{Hadoop Or Not Hadoop}
\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth}
%\date{}							% Activate to display a given date or no date

\graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}

\begin{document}
\maketitle

\abstract{New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field. With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging. 
Hadoop, emerged in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes. We here propose and report the quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases. We modified and wrote new pieces of code in order to substantially increase the performance of existing software. Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches.
Our results show that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines  become performance-competitive with a canonical high-performance cluster solutions.

}

\section{Introduction}
\label{sectionI}
%The problem: dealing with BIG data
Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}. Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.  As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 day~\cite{illumina}.  Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} HPC center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.

%Current practices: HPC systems
A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample. Most of the bioinformatic programs (tools) are written for the Linux operating system. Some of the most widespread software tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000}, like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.

%Current practices: How to parallelize

Most common  way to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.  For the tools which do not support the OMP natively, i.e.  the Samtools, a variant calling can be parallelised specifying the option, creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
The fact of great importance, however, is that the multi-core approach does not improve the performance of operations that are limited by {\it disk} or {\it network } throughputs, motivating to split  the dataset and  use the  multi-node parallelisation.

The Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.

Another common way to introduce the parallelisation in Linux systems is to use Bash scripting e.g. involve  already existing utilities and cluster tools, split the data into chunks to deliver and proceed them  on the separate nodes, merging the results afterwards. This kind of solution benefiting  from both MPI-like and OMP, is involved in our work as well. It provides great performance, but usually it is tightly binded to the local computational cluster and network architectures.

%Introduce Hadoop, MR, and HDFS
The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running task in a {\it massively} parallel way. This paradigm, however, shifts the focus from the best performance to scalability, suited for managing  huge datasets of size  of terabytes~\cite{lin2010}.
The prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. 
For the purpose of this work, we focus on the Cloudera Hadoop MR with the Hadoop Distributed File System~(HDFS).

%Properties of Hadoop and HDFS- data localization and programming model etc
The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}. HDFS is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto each node’s local disks, therefore effectively moving the computation to the data and hence reducing network traffic. The HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.

%Manuscript focus: When is Hadoop currently an appealing alternative
In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. Since Hadoop is written in Java, which is slower than the standard~HPC programming languages, like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.

%Introduce what we did in the manuscript.
We use five datasets with short reads of different size to align them against the reference genome, and call the  variances.  For one dataset we made a check to identify the actual mutation to make sure that HPC and Hadoop give the same answers. 
From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
For the classical HPC DNA-seq analysis programs we developed a general-use bash scripts, utilizing multiple nodes for short reads alignment and exploiting the network fully, speeding up the calculations.
The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.

The manuscript is structured as follows: In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used. Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources. The results are discussed in Section~IV, and conclusions in Section~V. Supplementary materials are provided in the corresponding section.

\section{Methods}
\label{sectionII}

\subsection{Datasets}
We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals, Table~1. Data for datasets I-III,V were generated using Illumina/(HiSeq) sequencing platforms. Further information about the datasets is provided in the Supplementary material section.

\begin{table}[htdp]
\small
\caption{Datasets used}
\begin{center}
\begin{tabular}{|l|l|l|}
dataset &	organism &	size in Gbases\\
\hline
 I		&	{\it A.thaliana}	&	1.4	\\
 II	&	{\it A.thaliana}	&	7.0\\
  III	&	{\it A.thaliana}	&	30.0	\\
 IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
 V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\

\end{tabular}
\end{center}
\label{table:datasets}
\normalsize
\end{table}%

\subsection{Computational resources}
To run the HPC analysis pipeline nodes of  the computational clusters at UPPMAX, equipped with dual 4-core~(HPC1) and 8-core processors~(HPC2) were used. Data and reference genomes were read and written to a parallel shared storage system. The main Hadoop test platform (Hadoop-in-the-Cloud) was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} as a cloud manager. Each node in this deployment was equipped with dual 4-core CPUs. The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}. For details on computational resources, see Supplementary material.

\subsection{Analysis pipelines}
We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on HPC. Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage. All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform. Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e. GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:

\begin{enumerate}
\item HPC approach
\subitem short read alignment: Bowtie~ver.~0.12.8
\subitem SNP calling: Samtool~ver.~0.1.19
\item Hadoop approach: Crossbow~\cite{crossbow}
\subitem short read alignment: Bowtie~ver.~0.12.8
\subitem SNP calling: SOAPsnp~1.02~\cite{soapsnp}
\end{enumerate}

In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools. The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained in~8 times faster than on a single core. Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions. Where available these features were used to improve the analysis speed. For the exact workflow used one can refer to the code repository created for this work~\cite{code_repo}.

The equivalent Hadoop-based pipeline was implemented with the Crossbow.  The input data and the indexed genome reference were copied to Hadoop’s storage system~(HDFS) before starting the experiments.  Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately Crossbow's preprocessor is not written  in the MR manner, and thus cannot be run in a massively parallel way. Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted the Crossbow's preprocessor with our own, written in the MR-style, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by the Hadoop.}. To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing stage time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster. The scripts for preprocessing stage are publicly available~\cite{code_repo}.

\section{Results}
\label{sectionIII}

\subsection{Accuracy of pipelines}
Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, still we expect to capture and correctly identify the mutation. We tested the correctness just for the smallest dataset~I. The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized\cite{schneeberger}.

\subsection{Scalability of HPC and Hadoop approaches}
To demonstrate the scalability of  the HPC approach we collected running time for dataset~I, as a function of a number of cores used, see Table~2.
\begin{table}[htdp]
\small
\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC1 cluster. }
\begin{center}
\begin{tabular}{|c|c|c|c|}
$N$ cores	&\multicolumn{3}{c|}{timing, minutes}\\
\hline
	& aligning 	&	SNP calling\tablefootnote{We do not consider here special tricks how to parallelize the Samtools analysis by chromosome, as could be done, as exemplified here \url{http://www.biostars.org/p/48781/
}}	&	total \\
\hline
1	&	38	&	37		&	75 \\
2	&	19	&	36		&	55\\
4	&	10	&	36		&	46\\
8	&	5	&	36		&	41\\
\end{tabular}
\end{center}
\label{table:hpctiming}
\normalsize
\end{table}
As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, revealing almost no scaling at all.
For the given datasets (I--V) the timings for alignment against the corresponding genomes, and SNP calling for  the HPC and Hadoop approaches were collected, Table~3.

\begin{table}[htdp]
\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.    Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}

data, Gbases&	1.4&	3.5&	7.0 	&15.0 	&30.0 	&100.0 	&250 	\\
			&	(I)			& (II, f.r.)	&	(II)		&(III, f.r.) &(III)&(IV) &(V)\\
\hline
Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
\hline
%Hadoop, 56 cores, Shared~TM&	25	&	24	&40		&	70&	150	&410&\\
%\hline
%Hadoop, 56 cores, ``bare-metal''  &	36	&	41	&53  	&63	&	92& 380	&\\
%\hline
HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
%\hline
%HPC2, 8 cores&72	&	&205	&&413&1765&\\
%\hline
%$F_{Hadoop}/F_{HPC1}$, Shared TM &	4.27		&1.75	&1.78	&	1.60	&	1.75	&	1.93&\\
%\hline
%$F_{Hadoop}/F_{HPC1}$, SSH TM &&		&1.74	&1.41	&1.28	&1.17	&\\
%\hline
%$F_{Hadoop}/F_{HPC2}$, ``bare-metal'' &&&&&&&\\

%data, 	& \multicolumn{2}{c|}{Hadoop, 56 cores}		  	&  HPC			&	\multicolumn{2}{c|}{$F_{Hadoop}/F_{HPC}$}\\
%Gbases 	&	 SSH~TM,	&	Shared~TM,	&	(8 cores), 	&	Shared TM	& 	SSH TM\\
%&  minutes& minutes&  minutes& &\\

\end{tabular}
\end{center}
\label{table:2}
\end{table}%

One of the attractive sides of using Hadoop is its almost linear scalability,~i.e. calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, {\it regardless} the scaling nature of the underlying program (Samtools, Bowtie etc.). 
Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster. Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
To stress the linear nature of scaling, both sets of points were fit into linear polynomial using least squares method.

\begin{figure}
	\input{./figures/fig1/fig1.tex}
	\caption{Calculation time depending on the dataset size for fixed size~($p=56$ cores) of Hadoop cluster on the private Cloud. The linear fit based on least squares method are provided, revealing almost linear scaling.}
	\label{fig:fig1}
\end{figure}

\subsection{Comparing Hadoop and HPC efficiency for different dataset sizes}
Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}. One can propose then,  that the larger the dataset is, more suitable Hadoop becomes compare to HPC. In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
$$F=T_{p}\times p,$$ 
where~$T_{p}$ is a calculation time on~$p$ cores.
Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC1}$, keeping in mind, the closer the ratio to the unity, the closer Hadoop’s efficiency approaches to that of HPC.

\begin{figure}
	\input{./figures/fig2/fig2.tex}
	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC1 correspondingly.
The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.}
	\label{fig:fig2}
\end{figure}

The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop-in-the-Cloud cluster, and a HPC1 node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC1}$ as a function of the {\it reciprocal} dataset size. The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical {\it infinite} dataset. As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data becomes larger. With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the {\it virtualized} environment is competitive with  the HPC approach, which is being run in ``bare metal'', for the datasets greater than $100$\,Gbases (Dataset IV), which is a usual size for human genome sequencing, giving the average sequencing depth of  about~$30x$.

 \subsection{Comparing the network communication efficiency for Hadoop and HPC approaches }
Network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with the NAS or NFS attached storages. Namely, the effective bandwidth of the Hadoop network increases with the cluster size to increase~\cite{Sammer:2012}, opposite to that of the HPC network, where the cluster grows results in network saturation and performance depletion.
We provide a comparison how HPC and Hadoop  network communication costs depend on the number of nodes involved, for a fixed dataset size, dataset~IV.

Due to the fact of trivial parallelisation of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computation resources, i.e. split the initial data into chunks to proceed them independently.
Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost {\it simultaneously} have to travel through the network, potentially causing traffic jams, therefore increasing the communication costs,~$T_{comm}$.

There are several program packages for short reads alignment designed with the MPI support~\cite{pmap, gnumap}. Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}. However,  i.e. Pmap package had been proved to function poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising  memory exceptions.
We wrote custom state-of-art bash script involving existing Unix utilities~\cite{repo} to use the  HPC2 cluster network as efficiently as possible,  and compared the network performance with the {\it standard} Hadoop HDFS approach.

We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$. 
Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases. For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to  able to be run in the MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS. 
We rewrote the code for the preprocessing stage for the Crossbow to make it suitable for MR-style parallelisation.
For the HPC approach, the~$T_{comm}$ involves the chunks travel from the sequencer delivery location to the local node scratches, where the actual mapping happens, and the travel of the aligned SAM files back to the delivery location over the network.

Figure~\ref{fig:fig3} shows the~$T_{mapping}/T_{comm}$ ratio as a function of the reciprocal number of nodes~$1/N$ for Hadoop and HPC approaches. In its turn, the HPC approach is presented in two versions, which are based on a bit different strategies of the resource allocation.

Lets start from description of the Hadoop results, filled circles. One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40. It is known  that the Bowtie, used for mapping, provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure 1. Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication in the preprocessing stage is.

Lets now consider the curves for HPC2. The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in {\it parallel}, which are  {\it simultaneously} being pushed  to the local scratches of the nodes, allocated by SLURM  (open circles). One can see two distinct linear stretches, each with different tangent. One stretch is for the range from 4 to about 12 nodes the, and the another is from 12 up to 60. The former, horizontal stretch, is explained as for Hadoop~-- the more nodes is being involved the faster the chunks are being distributed. The latter stretch with the positive slope could be explained as follows: in the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX   is a set of RAIDs (Redundant Array of Inexpensive Disks) with data {\it striping}, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx const, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e. a linear dependency, which one can observe on the plot. 
The transition area between two modes~-- saturated and unsaturated~-- has the next origin: each HDD on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, what is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch. The nodes are being allocated  on the same rack, what is the default SLURM behaviour, therefore the caption reads~``HPC SLURM''.

The scalability can be improved by overriding the default behaviour of the SLURM and allocating the nodes not from the same rack, but randomly from all available racks, ``HPC random'', (open squares in Figure~\ref{fig:fig3}). Allocating the nodes on random racks allows one to engage more nodes without network saturation. For our cluster we could go up to 30-35 nodes with perfect linear scaling. For the most resources been used (58 nodes) the deviation from a linear speed-up is~$\approx 7\%$ i.e. $5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}  for the data.

\begin{figure}
	\small
	\input{./figures/fig9/fig3.tex}
	\normalsize
	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC2 and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$. Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one, where nodes are being allocated from random racks.
%One  can see the differences in the Hadoop and HPC communication patterns: for Hadoop the ratio is almost constant, while for HPC cluster there are two defined regions.
%The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
%The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
%linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
Linear fit  done with the least-squares method.}
	\label{fig:fig3}
\end{figure}

The threshold number of nodes in this strategy~($\approx35$) is driven by the saturating the uplink cable with the  throughput of~50Gbit/sec. 
The proposed strategies aimed to gain the maximum from the existing resources show, that even properly adjusted and tuned, HPC approach sooner of later starts to suffer from the network saturation.

\begin{table}[htdp]
\caption{Timings  for mapping, and the ratio~$T_{mapping}/T_{comm}$  for HPC2 and  Hadoop clusters for Dataset IV.
For ``HPC random'' approach data chunks have to be copied to the local scratches first, and  the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS. Hadoop needs to preprocess reads before the actual alignment stage, in order  to be able to operate in a MR manner, also resulting in  ``communication costs''. Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
 \multicolumn{3}{|c|}{Hadoop} & \multicolumn{3}{c|}{ HPC random} \\
 \hline

Number nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$	&Number  nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$\\
(cores)					&minutes		&							&(cores)			&minutes&\\
\hline
4(28)	&293.5	&2.33	&4(64)	&74.4	&3.89\\
6(42)	&189.8	&2.19	&10(160)	&32.4	&3.76\\
8(56)	&136.0	&2.23	&14(224)	&22.7	&3.77\\
16(112)	&70.3	&1.96	&18(288)	&17.9	&3.78\\
32(224)	&39.3	&2.20	&22(352)	&14.5	&3.79\\
40(280)	&32.5	&2.15	&26(416)	&12.3	&3.77\\
			&&&30(480)	&10.7	&3.73\\
			&&&34(544)	&9.5	&3.45\\
			&&&38(608)	&8.5	&3.16\\
			&&&42(672)	&7.6	&2.96\\
			&&&46(736)	&7.0	&2.55\\
			&&&50(800)	&6.4	&2.65\\
			&&&54(864)	&5.9	&2.34\\
			&&&58(928)	&5.5	&2.12\\

\end{tabular}
\end{center}
\label{table:4}
\end{table}%
 At the same time, the HDFS keeps data locality, aiming to reduce the amount of communications, resulting less data move and, therefore, better scalability.
Our Hadoop-in-the-Cloud cluster has no more~($\approx 40$) free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the exposed behaviour is a generic for Hadoop with HDFS.

\subsection{Usability aspects}
\label{subsectionIV_2}

At the present moment a popular way to steer the bioinformatic pipelines on the HPC resources  is to use Galaxy\cite{galaxy}, which provides a Web-based graphical user interface~(GUI) to the bioinformatic programs, simplifying an experience for the end user. 
One of the alternatives for Hadoop could be the publically available Cloudgene\cite{cloudgene}, which seems to be a very light-weight, and flexible Web-based solution for serve~GUI for both public and private cloud, which we involved in our work.
For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow, Figure~\ref{fig:fig4}. The most of data managing job is done automatically, and the results can be downloaded to the client machine. Modular structure allows one easy to modify the source code to integrate to the existing computing center’s architecture, Figure~\ref{fig:fig4}(a). For example the UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will stay locally, reducing amount of unnecessary risks.

\begin{figure}
	\begin{subfigure}[b]{0.6\textwidth}
	        	\includegraphics[width=\textwidth]{c.png}
		\subcaption{UPPMAX-adapted Cloudgene: browsing the users home folder}
	\end{subfigure}
 	\begin{subfigure}[b]{0.4\textwidth}
        	\includegraphics[width=\textwidth]{a.png}
		\subcaption{Cloudgene: selecting a job to run}
	\end{subfigure}
	\begin{subfigure}[b]{0.6\textwidth}
		\includegraphics[width=\textwidth]{b.png}
		\subcaption{Cloudgene: specifying job parameters}
	\end{subfigure}
	\caption{An example of a job setup with the Cloudgene, a web-based GUI wrapper, providing a smooth user experience even for novice users.}
	\label{fig:fig4}
\end{figure}

\section{Conclusions}
\label{sectionV}

In this communication we brought together different approaches to analyze  the DNA-seq data in the bioinformatic study, in order to answer the question: which conditions a task should meet in order to make Hadoop a proper tool to use in favour of using HPC resources.

To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow software were suggested.  A state-of-art bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.

We concentrated ourselves on the Hadoop  in the private cloud installation, Hadoop-in-the-Cloud, orchestrated by the OpenNebula. 
Picking the appropriate program settings we found out that DNA-seq dataset size larger than 100\,Gbases is suitable for Hadoop, with competitive  execution time compare to  that of HPC.
The scaling graphs (Figure~\ref{fig:fig1}) confirms the known fact for almost linear scalability of Hadoop.
The extrapolation to the infinite dataset size (Figure~\ref{fig:fig2}) revealed, that HPC, however, provides the results faster, given the same amount of resources, than the Hadoop-in-the-Cloud.

Exploiting  the embarrassingly parallel nature of the short reads mapping we used custom state-of-art bash script to engage up to 58 nodes (928 cores) of the HPC2 cluster to  see the scaling relations between the ratio of the mapping time to the communication time as a function of a reciprocal number of nodes, Figure~\ref{fig:fig3}. 
Our results show that Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
In addition we show  an example, how one can improve the performance of the HPC approach redefining the SLURM's default behaviour.

Finally, existing and easily transformed publically available web-based GUI, like Cloudgene~(Figure~\ref{fig:fig4}),  gives an opportunity to perform bioinformatic analysis on Hadoop for those who are not experts in Linux world.  Modular structure of Cloudgene gives cluster staff an easy way to adapt it for the specific needs.

\section{Acknowledge}
The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (UPPMAX) under Project p2013023.
We also greatly appreciate   Pontus Freyhult and Peter Ankerst{\aa}l at  UPPMAX for explaining the effective storage usage. Great job was done by Jonas Hagberg at BILS, Stockholm, Sweden, by the adaptation of the Cloudgene to the local UPPMAX needs.

\section{Supplementary material}

\subsection{Used datasets}

The datasets used in the paper are publicly available at:

I: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0\\/data/reads/Schneeberger.2009/Schneeberger.2009.single\_end.gz}

II: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads1.fq.gz, http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads2.fq.gz}

III: \url{ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611084//SRR611084.sra, ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611085//SRR611085.sra}

IV: artificial pair-ended dataset for {\it A.thaliana} created with the {\tt wgsim} program from the Samtools package.

V: \url{http://www.ncbi.nlm.nih.gov/sra/SRX148888}

\subsection{Reference genomes}
\begin{itemize}
\item TAIR10 for datasets II-IV \url{ftp://ftp.arabidopsis.org/home/tair/Sequences/whole\_chromosomes/*.fas}
\item TAIR8 for dataset I \url{ftp://ftp.arabidopsis.org/home/tair/Genes/TAIR8\_genome\_release/}
\item H.sapiens, NCBI v37 \url{ftp://ftp.ccb.jhu.edu/pub/data/bowtie\_indexes/h\_sapiens\_37\_asm.ebwt.zip}
\end{itemize}

\subsection{Description of computational facilities}

\begin{enumerate}
\item  HPC1:
The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink. The data and reference genomes were read and written to a parallel shared storage system.

\item 
HPC2:
Multinode short read mapping was performed on the Milou cluster\cite{milouCluster} , equipped with dual 8cores CPUs Intel Xeon E5-2660, ( 2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache.), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.

\item Storage: 
Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600 storage boxes and an additional node for metadata handling. In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol. It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.

\item Our main Hadoop test platform was deployed on the private cloud at UPPMAX, using the OpenNebula~\cite{opennebula} cloud management system. Each node in this deployment was equipped with dual quad-core CPUs (Intel Xeon 5420; clock frequency of 2.50GHz\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet. The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.

\end{enumerate}

\bibliography{text1}
\bibliographystyle{unsrt}

\end{document}

### 2014-01-15 figures/fig2/text.tex
ms text figures

\documentclass{article}
\usepackage{graphics}
\usepackage{nopageno}
%\usepackage{txfonts}
\usepackage[usenames]{color}
%\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb} %maths
\begin{document}
\begin{center}
\input{fig.tex}
\end{center}
The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC correspondingly.
The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.
\end{document}

### 2014-01-15 figures/fig3/text.tex
ms text figures

\documentclass{article}
\usepackage{graphicx}
\usepackage{nopageno}
\usepackage{txfonts}
\usepackage[usenames]{color}
\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
%\usepackage{amsmath}

\begin{document}
\begin{center}
	\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	   \includegraphics[width=13cm]{fig3.jpg} 
	   \caption{Sketch of difference between the possible scenarios of Hadoop deployment: Shared~TM and SSH~TM. The Shared~TM deployment scenario (left) assumes that the storage, where the data is being keeped and proceeded, is a NFS-mounted Shared Storage, connected to the computing nodes by a fast cable. The SSH~TM (right) is in opposite engages the local storages of each virtual node, which are usually not slower than a fast cable. The difference emerges then computation nodes start to communicate with each other in order to share with the data. In that case, even fast cable starts to saturate, and the IO bottleneck appears.}
	   \label{fig:fig3}
	\end{figure}
\end{center}
\end{document}

### 2014-01-15 figures/fig5/text.tex
ms text figures

--- 2014-01-15 text1.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or epsБ•√ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
-\usepackage{amssymb}
-\usepackage[colorlinks]{hyperref}
-\usepackage{booktabs}
-\usepackage{amsmath}
-\usepackage{multirow}% http://ctan.org/pkg/multirow
-\usepackage{tablefootnote}
-\usepackage{import}
+\documentclass{article}
+\usepackage{graphicx}
+\usepackage{nopageno}
+\usepackage{txfonts}
+\usepackage[usenames]{color}
+%\usepackage[T2A]{fontenc}
+\usepackage[utf8]{inputenc}
 \usepackage{subcaption}
@@
 
-\title{Hadoop Or Not Hadoop}
-\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth}
-%\date{}							% Activate to display a given date or no date
-
-\graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
-
 \begin{document}
-\maketitle
-
-\abstract{New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field.
-With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
-Hadoop, emerged in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-We here propose and report the quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases.
-We modified and wrote new pieces of code in order to substantially increase the performance of existing software.
-Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches.
-Our results show that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines  become performance-competitive with a canonical high-performance cluster solutions.
-
-}
-
-\section{Introduction}
-\label{sectionI}
-%The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}.
-Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 day~\cite{illumina}.
-Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} HPC center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
-
-%Current practices: HPC systems
-A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Most of the bioinformatic programs (tools) are written for the Linux operating system.
-Some of the most widespread software tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000}, like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
-
-%Current practices: How to parallelize
-
-Most common  way to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.
-For the tools which do not support the OMP natively, i.e.
-the Samtools, a variant calling can be parallelised specifying the option, creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
-The fact of great importance, however, is that the multi-core approach does not improve the performance of operations that are limited by {\it disk} or {\it network } throughputs, motivating to split  the dataset and  use the  multi-node parallelisation.
-
-The Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
-
-Another common way to introduce the parallelisation in Linux systems is to use Bash scripting e.g.
-involve  already existing utilities and cluster tools, split the data into chunks to deliver and proceed them  on the separate nodes, merging the results afterwards.
-This kind of solution benefiting  from both MPI-like and OMP, is involved in our work as well.
-It provides great performance, but usually it is tightly binded to the local computational cluster and network architectures.
-
-%Introduce Hadoop, MR, and HDFS
-The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running task in a {\it massively} parallel way.
-This paradigm, however, shifts the focus from the best performance to scalability, suited for managing  huge datasets of size  of terabytes~\cite{lin2010}.
-The prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. 
-For the purpose of this work, we focus on the Cloudera Hadoop MR with the Hadoop Distributed File System~(HDFS).
-
-%Properties of Hadoop and HDFS- data localization and programming model etc
-The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
-HDFS is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto each node’s local disks, therefore effectively moving the computation to the data and hence reducing network traffic.
-The HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
-
-%Manuscript focus: When is Hadoop currently an appealing alternative
-In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
-Since Hadoop is written in Java, which is slower than the standard~HPC programming languages, like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
-
-%Introduce what we did in the manuscript.
-We use five datasets with short reads of different size to align them against the reference genome, and call the  variances.
-For one dataset we made a check to identify the actual mutation to make sure that HPC and Hadoop give the same answers.
-From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
-For the classical HPC DNA-seq analysis programs we developed a general-use bash scripts, utilizing multiple nodes for short reads alignment and exploiting the network fully, speeding up the calculations.
-The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.
-
-The manuscript is structured as follows: In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
-Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources.
-The results are discussed in Section~IV, and conclusions in Section~V.
-Supplementary materials are provided in the corresponding section.
-
-\section{Methods}
-\label{sectionII}
-
-\subsection{Datasets}
-We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals, Table~1.
-Data for datasets I-III,V were generated using Illumina/(HiSeq) sequencing platforms.
-Further information about the datasets is provided in the Supplementary material section.
-
-\begin{table}[htdp]
-\small
-\caption{Datasets used}
 \begin{center}
-\begin{tabular}{|l|l|l|}
-dataset &	organism &	size in Gbases\\
-\hline
- I		&	{\it A.thaliana}	&	1.4	\\
- II	&	{\it A.thaliana}	&	7.0\\
-  III	&	{\it A.thaliana}	&	30.0	\\
- IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
- V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
-
-\end{tabular}
-\end{center}
-\label{table:datasets}
-\normalsize
-\end{table}%
-
-\subsection{Computational resources}
-To run the HPC analysis pipeline nodes of  the computational clusters at UPPMAX, equipped with dual 4-core~(HPC1) and 8-core processors~(HPC2) were used.
-Data and reference genomes were read and written to a parallel shared storage system.
-The main Hadoop test platform (Hadoop-in-the-Cloud) was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} as a cloud manager.
-Each node in this deployment was equipped with dual 4-core CPUs.
-The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-For details on computational resources, see Supplementary material.
-
-\subsection{Analysis pipelines}
-We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on HPC.
-Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage.
-All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
-Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e.
-GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:
-
-\begin{enumerate}
-\item HPC approach
-\subitem short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: Samtool~ver.~0.1.19
-\item Hadoop approach: Crossbow~\cite{crossbow}
-\subitem short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: SOAPsnp~1.02~\cite{soapsnp}
-\end{enumerate}
-
-In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools.
-The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained in~8 times faster than on a single core.
-Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
-Where available these features were used to improve the analysis speed.
-For the exact workflow used one can refer to the code repository created for this work~\cite{code_repo}.
-
-The equivalent Hadoop-based pipeline was implemented with the Crossbow.
-The input data and the indexed genome reference were copied to Hadoop’s storage system~(HDFS) before starting the experiments.
-Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
-Unfortunately Crossbow's preprocessor is not written  in the MR manner, and thus cannot be run in a massively parallel way.
-Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted the Crossbow's preprocessor with our own, written in the MR-style, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by the Hadoop.}.
-To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing stage time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster.
-The scripts for preprocessing stage are publicly available~\cite{code_repo}.
-
-\section{Results}
-\label{sectionIII}
-
-\subsection{Accuracy of pipelines}
-Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, still we expect to capture and correctly identify the mutation.
-We tested the correctness just for the smallest dataset~I.
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized\cite{schneeberger}.
-
-\subsection{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of  the HPC approach we collected running time for dataset~I, as a function of a number of cores used, see Table~2.
-\begin{table}[htdp]
-\small
-\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC1 cluster. }
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}\\
-\hline
-	& aligning 	&	SNP calling\tablefootnote{We do not consider here special tricks how to parallelize the Samtools analysis by chromosome, as could be done, as exemplified here \url{http://www.biostars.org/p/48781/
-}}	&	total \\
-\hline
-1	&	38	&	37		&	75 \\
-2	&	19	&	36		&	55\\
-4	&	10	&	36		&	46\\
-8	&	5	&	36		&	41\\
-\end{tabular}
-\end{center}
-\label{table:hpctiming}
-\normalsize
-\end{table}
-As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, revealing almost no scaling at all.
-For the given datasets (I--V) the timings for alignment against the corresponding genomes, and SNP calling for  the HPC and Hadoop approaches were collected, Table~3.
-
-\begin{table}[htdp]
-\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.    Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-
-data, Gbases&	1.4&	3.5&	7.0 	&15.0 	&30.0 	&100.0 	&250 	\\
-			&	(I)			& (II, f.r.)	&	(II)		&(III, f.r.) &(III)&(IV) &(V)\\
-\hline
-Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
-\hline
-%Hadoop, 56 cores, Shared~TM&	25	&	24	&40		&	70&	150	&410&\\
-%\hline
-%Hadoop, 56 cores, ``bare-metal''  &	36	&	41	&53  	&63	&	92& 380	&\\
-%\hline
-HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
-%\hline
-%HPC2, 8 cores&72	&	&205	&&413&1765&\\
-%\hline
-%$F_{Hadoop}/F_{HPC1}$, Shared TM &	4.27		&1.75	&1.78	&	1.60	&	1.75	&	1.93&\\
-%\hline
-%$F_{Hadoop}/F_{HPC1}$, SSH TM &&		&1.74	&1.41	&1.28	&1.17	&\\
-%\hline
-%$F_{Hadoop}/F_{HPC2}$, ``bare-metal'' &&&&&&&\\
-
-%data, 	& \multicolumn{2}{c|}{Hadoop, 56 cores}		  	&  HPC			&	\multicolumn{2}{c|}{$F_{Hadoop}/F_{HPC}$}\\
-%Gbases 	&	 SSH~TM,	&	Shared~TM,	&	(8 cores), 	&	Shared TM	& 	SSH TM\\
-%&  minutes& minutes&  minutes& &\\
-
-\end{tabular}
-\end{center}
-\label{table:2}
-\end{table}%
-
-One of the attractive sides of using Hadoop is its almost linear scalability,~i.e.
-calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, {\it regardless} the scaling nature of the underlying program (Samtools, Bowtie etc.).
-Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
-Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
-To stress the linear nature of scaling, both sets of points were fit into linear polynomial using least squares method.
-
-\begin{figure}
-	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time depending on the dataset size for fixed size~($p=56$ cores) of Hadoop cluster on the private Cloud.
-The linear fit based on least squares method are provided, revealing almost linear scaling.}
-	\label{fig:fig1}
-\end{figure}
-
-\subsection{Comparing Hadoop and HPC efficiency for different dataset sizes}
-Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then,  that the larger the dataset is, more suitable Hadoop becomes compare to HPC.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
-$$F=T_{p}\times p,$$ 
-where~$T_{p}$ is a calculation time on~$p$ cores.
-Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC1}$, keeping in mind, the closer the ratio to the unity, the closer Hadoop’s efficiency approaches to that of HPC.
-
-\begin{figure}
-	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC1 correspondingly.
-The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.}
-	\label{fig:fig2}
-\end{figure}
-
-The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop-in-the-Cloud cluster, and a HPC1 node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC1}$ as a function of the {\it reciprocal} dataset size.
-The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical {\it infinite} dataset.
-As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data becomes larger.
-With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the {\it virtualized} environment is competitive with  the HPC approach, which is being run in ``bare metal'', for the datasets greater than $100$\,Gbases (Dataset IV), which is a usual size for human genome sequencing, giving the average sequencing depth of  about~$30x$.
-
- \subsection{Comparing the network communication efficiency for Hadoop and HPC approaches }
-Network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with the NAS or NFS attached storages.
-Namely, the effective bandwidth of the Hadoop network increases with the cluster size to increase~\cite{Sammer:2012}, opposite to that of the HPC network, where the cluster grows results in network saturation and performance depletion.
-We provide a comparison how HPC and Hadoop  network communication costs depend on the number of nodes involved, for a fixed dataset size, dataset~IV.
-
-Due to the fact of trivial parallelisation of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computation resources, i.e.
-split the initial data into chunks to proceed them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost {\it simultaneously} have to travel through the network, potentially causing traffic jams, therefore increasing the communication costs,~$T_{comm}$.
-
-There are several program packages for short reads alignment designed with the MPI support~\cite{pmap, gnumap}.
-Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}.
-However,  i.e.
-Pmap package had been proved to function poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising  memory exceptions.
-We wrote custom state-of-art bash script involving existing Unix utilities~\cite{repo} to use the  HPC2 cluster network as efficiently as possible,  and compared the network performance with the {\it standard} Hadoop HDFS approach.
-
-We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$. 
-Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases.
-For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to  able to be run in the MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS.
-We rewrote the code for the preprocessing stage for the Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, the~$T_{comm}$ involves the chunks travel from the sequencer delivery location to the local node scratches, where the actual mapping happens, and the travel of the aligned SAM files back to the delivery location over the network.
-
-Figure~\ref{fig:fig3} shows the~$T_{mapping}/T_{comm}$ ratio as a function of the reciprocal number of nodes~$1/N$ for Hadoop and HPC approaches.
-In its turn, the HPC approach is presented in two versions, which are based on a bit different strategies of the resource allocation.
-
-Lets start from description of the Hadoop results, filled circles.
-One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known  that the Bowtie, used for mapping, provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure 1.
-Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication in the preprocessing stage is.
-
-Lets now consider the curves for HPC2.
-The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in {\it parallel}, which are  {\it simultaneously} being pushed  to the local scratches of the nodes, allocated by SLURM  (open circles).
-One can see two distinct linear stretches, each with different tangent.
-One stretch is for the range from 4 to about 12 nodes the, and the another is from 12 up to 60.
-The former, horizontal stretch, is explained as for Hadoop~-- the more nodes is being involved the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: in the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX   is a set of RAIDs (Redundant Array of Inexpensive Disks) with data {\it striping}, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx const, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
-a linear dependency, which one can observe on the plot.
-The transition area between two modes~-- saturated and unsaturated~-- has the next origin: each HDD on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, what is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated  on the same rack, what is the default SLURM behaviour, therefore the caption reads~``HPC SLURM''.
-
-The scalability can be improved by overriding the default behaviour of the SLURM and allocating the nodes not from the same rack, but randomly from all available racks, ``HPC random'', (open squares in Figure~\ref{fig:fig3}).
-Allocating the nodes on random racks allows one to engage more nodes without network saturation.
-For our cluster we could go up to 30-35 nodes with perfect linear scaling.
-For the most resources been used (58 nodes) the deviation from a linear speed-up is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}  for the data.
-
-\begin{figure}
-	\small
-	\input{./figures/fig9/fig3.tex}
-	\normalsize
-	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC2 and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
-Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one, where nodes are being allocated from random racks.
-%One  can see the differences in the Hadoop and HPC communication patterns: for Hadoop the ratio is almost constant, while for HPC cluster there are two defined regions.
-%The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
-%The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
-%linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
-Linear fit  done with the least-squares method.}
-	\label{fig:fig3}
-\end{figure}
-
-The threshold number of nodes in this strategy~($\approx35$) is driven by the saturating the uplink cable with the  throughput of~50Gbit/sec. 
-The proposed strategies aimed to gain the maximum from the existing resources show, that even properly adjusted and tuned, HPC approach sooner of later starts to suffer from the network saturation.
-
-\begin{table}[htdp]
-\caption{Timings  for mapping, and the ratio~$T_{mapping}/T_{comm}$  for HPC2 and  Hadoop clusters for Dataset IV.
-For ``HPC random'' approach data chunks have to be copied to the local scratches first, and  the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS.
-Hadoop needs to preprocess reads before the actual alignment stage, in order  to be able to operate in a MR manner, also resulting in  ``communication costs''.
-Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
- \multicolumn{3}{|c|}{Hadoop} & \multicolumn{3}{c|}{ HPC random} \\
- \hline
-
-Number nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$	&Number  nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$\\
-(cores)					&minutes		&							&(cores)			&minutes&\\
-\hline
-4(28)	&293.5	&2.33	&4(64)	&74.4	&3.89\\
-6(42)	&189.8	&2.19	&10(160)	&32.4	&3.76\\
-8(56)	&136.0	&2.23	&14(224)	&22.7	&3.77\\
-16(112)	&70.3	&1.96	&18(288)	&17.9	&3.78\\
-32(224)	&39.3	&2.20	&22(352)	&14.5	&3.79\\
-40(280)	&32.5	&2.15	&26(416)	&12.3	&3.77\\
-			&&&30(480)	&10.7	&3.73\\
-			&&&34(544)	&9.5	&3.45\\
-			&&&38(608)	&8.5	&3.16\\
-			&&&42(672)	&7.6	&2.96\\
-			&&&46(736)	&7.0	&2.55\\
-			&&&50(800)	&6.4	&2.65\\
-			&&&54(864)	&5.9	&2.34\\
-			&&&58(928)	&5.5	&2.12\\
-
-\end{tabular}
-\end{center}
-\label{table:4}
-\end{table}%
- At the same time, the HDFS keeps data locality, aiming to reduce the amount of communications, resulting less data move and, therefore, better scalability.
-Our Hadoop-in-the-Cloud cluster has no more~($\approx 40$) free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the exposed behaviour is a generic for Hadoop with HDFS.
-
-\subsection{Usability aspects}
-\label{subsectionIV_2}
-
-At the present moment a popular way to steer the bioinformatic pipelines on the HPC resources  is to use Galaxy\cite{galaxy}, which provides a Web-based graphical user interface~(GUI) to the bioinformatic programs, simplifying an experience for the end user.
-One of the alternatives for Hadoop could be the publically available Cloudgene\cite{cloudgene}, which seems to be a very light-weight, and flexible Web-based solution for serve~GUI for both public and private cloud, which we involved in our work.
-For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow, Figure~\ref{fig:fig4}.
-The most of data managing job is done automatically, and the results can be downloaded to the client machine.
-Modular structure allows one easy to modify the source code to integrate to the existing computing center’s architecture, Figure~\ref{fig:fig4}(a).
-For example the UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will stay locally, reducing amount of unnecessary risks.
-
-\begin{figure}
+	\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
 		\begin{subfigure}[b]{0.6\textwidth}
-	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{UPPMAX-adapted Cloudgene: browsing the users home folder}
-	\end{subfigure}
- 	\begin{subfigure}[b]{0.4\textwidth}
 		   \includegraphics[width=\textwidth]{a.png} 
@@
 		\caption{An example of a job setup with the Cloudgene, a web-based GUI wrapper, providing a smooth user experience even for novice users.}
-	\label{fig:fig4}
+		\label{fig:fig5}
 	\end{figure}
-
-\section{Conclusions}
-\label{sectionV}
-
-In this communication we brought together different approaches to analyze  the DNA-seq data in the bioinformatic study, in order to answer the question: which conditions a task should meet in order to make Hadoop a proper tool to use in favour of using HPC resources.
-
-To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow software were suggested.
-A state-of-art bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
-
-We concentrated ourselves on the Hadoop  in the private cloud installation, Hadoop-in-the-Cloud, orchestrated by the OpenNebula. 
-Picking the appropriate program settings we found out that DNA-seq dataset size larger than 100\,Gbases is suitable for Hadoop, with competitive  execution time compare to  that of HPC.
-The scaling graphs (Figure~\ref{fig:fig1}) confirms the known fact for almost linear scalability of Hadoop.
-The extrapolation to the infinite dataset size (Figure~\ref{fig:fig2}) revealed, that HPC, however, provides the results faster, given the same amount of resources, than the Hadoop-in-the-Cloud.
-
-Exploiting  the embarrassingly parallel nature of the short reads mapping we used custom state-of-art bash script to engage up to 58 nodes (928 cores) of the HPC2 cluster to  see the scaling relations between the ratio of the mapping time to the communication time as a function of a reciprocal number of nodes, Figure~\ref{fig:fig3}.
-Our results show that Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show  an example, how one can improve the performance of the HPC approach redefining the SLURM's default behaviour.
-
-Finally, existing and easily transformed publically available web-based GUI, like Cloudgene~(Figure~\ref{fig:fig4}),  gives an opportunity to perform bioinformatic analysis on Hadoop for those who are not experts in Linux world.
-Modular structure of Cloudgene gives cluster staff an easy way to adapt it for the specific needs.
-
-\section{Acknowledge}
-The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (UPPMAX) under Project p2013023.
-We also greatly appreciate   Pontus Freyhult and Peter Ankerst{\aa}l at  UPPMAX for explaining the effective storage usage.
-Great job was done by Jonas Hagberg at BILS, Stockholm, Sweden, by the adaptation of the Cloudgene to the local UPPMAX needs.
-
-\section{Supplementary material}
-
-\subsection{Used datasets}
-
-The datasets used in the paper are publicly available at:
-
-I: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0\\/data/reads/Schneeberger.2009/Schneeberger.2009.single\_end.gz}
-
-II: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads1.fq.gz, http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads2.fq.gz}
-
-III: \url{ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611084//SRR611084.sra, ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611085//SRR611085.sra}
-
-IV: artificial pair-ended dataset for {\it A.thaliana} created with the {\tt wgsim} program from the Samtools package.
-
-V: \url{http://www.ncbi.nlm.nih.gov/sra/SRX148888}
-
-\subsection{Reference genomes}
-\begin{itemize}
-\item TAIR10 for datasets II-IV \url{ftp://ftp.arabidopsis.org/home/tair/Sequences/whole\_chromosomes/*.fas}
-\item TAIR8 for dataset I \url{ftp://ftp.arabidopsis.org/home/tair/Genes/TAIR8\_genome\_release/}
-\item H.sapiens, NCBI v37 \url{ftp://ftp.ccb.jhu.edu/pub/data/bowtie\_indexes/h\_sapiens\_37\_asm.ebwt.zip}
-\end{itemize}
-
-\subsection{Description of computational facilities}
-
-\begin{enumerate}
-\item  HPC1:
-The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
-The data and reference genomes were read and written to a parallel shared storage system.
-
-\item 
-HPC2:
-Multinode short read mapping was performed on the Milou cluster\cite{milouCluster} , equipped with dual 8cores CPUs Intel Xeon E5-2660, ( 2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache.), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
-
-\item Storage: 
-Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600 storage boxes and an additional node for metadata handling.
-In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
-It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
-
-\item Our main Hadoop test platform was deployed on the private cloud at UPPMAX, using the OpenNebula~\cite{opennebula} cloud management system.
-Each node in this deployment was equipped with dual quad-core CPUs (Intel Xeon 5420; clock frequency of 2.50GHz\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
-The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-
-\end{enumerate}
-
-\bibliography{text1}
-\bibliographystyle{unsrt}
-
+\end{center}
 \end{document}

### 2014-01-15 figures/fig8/text.tex
ms text figures

\documentclass{article}
\usepackage{graphics}
\usepackage{nopageno}
\usepackage{txfonts}
\usepackage[usenames]{color}
\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
%\usepackage{amsmath}

\begin{document}
\begin{center}
\input{fig.tex}
\end{center}
The effeciency ratios\ref{eq:efficiency} $E_{p}/E_{56}$ for different  Hadoop on the Cloud cluster sizes~112  and~224 cores.
The data for Shared~TM and SSH~TM deployment scenarios are shown.
The efficinency ratios are shown for different pipeline stages: reads preprocessing~(1-3), alingment~(4), and SNP calling~(5).
The values above~1 shows that the cluster of current size is more efficient than the $p=56$ cores cluster, and vice versa. 
\end{document}

### 2014-01-15 figures/fig9/text.tex
ms text figures

--- 2014-01-15 text1.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or epsБ•√ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
-\usepackage{amssymb}
-\usepackage[colorlinks]{hyperref}
-\usepackage{booktabs}
-\usepackage{amsmath}
-\usepackage{multirow}% http://ctan.org/pkg/multirow
-\usepackage{tablefootnote}
-\usepackage{import}
-\usepackage{subcaption}
-\usepackage{caption}
-
-\title{Hadoop Or Not Hadoop}
-\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth}
-%\date{}							% Activate to display a given date or no date
-
-\graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
+\documentclass{article}
+\usepackage{graphics}
+\usepackage{nopageno}
+\usepackage{txfonts}
+\usepackage[usenames]{color}
+\usepackage[T2A]{fontenc}
+\usepackage[utf8]{inputenc}
+%\usepackage{amsmath}
 
 \begin{document}
-\maketitle
-
-\abstract{New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field.
-With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
-Hadoop, emerged in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-We here propose and report the quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases.
-We modified and wrote new pieces of code in order to substantially increase the performance of existing software.
-Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches.
-Our results show that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines  become performance-competitive with a canonical high-performance cluster solutions.
-
-}
-
-\section{Introduction}
-\label{sectionI}
-%The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}.
-Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 day~\cite{illumina}.
-Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} HPC center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
-
-%Current practices: HPC systems
-A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Most of the bioinformatic programs (tools) are written for the Linux operating system.
-Some of the most widespread software tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000}, like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
-
-%Current practices: How to parallelize
-
-Most common  way to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.
-For the tools which do not support the OMP natively, i.e.
-the Samtools, a variant calling can be parallelised specifying the option, creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
-The fact of great importance, however, is that the multi-core approach does not improve the performance of operations that are limited by {\it disk} or {\it network } throughputs, motivating to split  the dataset and  use the  multi-node parallelisation.
-
-The Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
-
-Another common way to introduce the parallelisation in Linux systems is to use Bash scripting e.g.
-involve  already existing utilities and cluster tools, split the data into chunks to deliver and proceed them  on the separate nodes, merging the results afterwards.
-This kind of solution benefiting  from both MPI-like and OMP, is involved in our work as well.
-It provides great performance, but usually it is tightly binded to the local computational cluster and network architectures.
-
-%Introduce Hadoop, MR, and HDFS
-The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running task in a {\it massively} parallel way.
-This paradigm, however, shifts the focus from the best performance to scalability, suited for managing  huge datasets of size  of terabytes~\cite{lin2010}.
-The prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. 
-For the purpose of this work, we focus on the Cloudera Hadoop MR with the Hadoop Distributed File System~(HDFS).
-
-%Properties of Hadoop and HDFS- data localization and programming model etc
-The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
-HDFS is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto each node’s local disks, therefore effectively moving the computation to the data and hence reducing network traffic.
-The HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
-
-%Manuscript focus: When is Hadoop currently an appealing alternative
-In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
-Since Hadoop is written in Java, which is slower than the standard~HPC programming languages, like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
-
-%Introduce what we did in the manuscript.
-We use five datasets with short reads of different size to align them against the reference genome, and call the  variances.
-For one dataset we made a check to identify the actual mutation to make sure that HPC and Hadoop give the same answers.
-From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
-For the classical HPC DNA-seq analysis programs we developed a general-use bash scripts, utilizing multiple nodes for short reads alignment and exploiting the network fully, speeding up the calculations.
-The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.
-
-The manuscript is structured as follows: In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
-Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources.
-The results are discussed in Section~IV, and conclusions in Section~V.
-Supplementary materials are provided in the corresponding section.
-
-\section{Methods}
-\label{sectionII}
-
-\subsection{Datasets}
-We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals, Table~1.
-Data for datasets I-III,V were generated using Illumina/(HiSeq) sequencing platforms.
-Further information about the datasets is provided in the Supplementary material section.
-
-\begin{table}[htdp]
-\small
-\caption{Datasets used}
 \begin{center}
-\begin{tabular}{|l|l|l|}
-dataset &	organism &	size in Gbases\\
-\hline
- I		&	{\it A.thaliana}	&	1.4	\\
- II	&	{\it A.thaliana}	&	7.0\\
-  III	&	{\it A.thaliana}	&	30.0	\\
- IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
- V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
-
-\end{tabular}
+\input{fig.tex}
 \end{center}
-\label{table:datasets}
-\normalsize
-\end{table}%
-
-\subsection{Computational resources}
-To run the HPC analysis pipeline nodes of  the computational clusters at UPPMAX, equipped with dual 4-core~(HPC1) and 8-core processors~(HPC2) were used.
-Data and reference genomes were read and written to a parallel shared storage system.
-The main Hadoop test platform (Hadoop-in-the-Cloud) was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} as a cloud manager.
-Each node in this deployment was equipped with dual 4-core CPUs.
-The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-For details on computational resources, see Supplementary material.
-
-\subsection{Analysis pipelines}
-We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on HPC.
-Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage.
-All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
-Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e.
-GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:
-
-\begin{enumerate}
-\item HPC approach
-\subitem short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: Samtool~ver.~0.1.19
-\item Hadoop approach: Crossbow~\cite{crossbow}
-\subitem short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: SOAPsnp~1.02~\cite{soapsnp}
-\end{enumerate}
-
-In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools.
-The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained in~8 times faster than on a single core.
-Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
-Where available these features were used to improve the analysis speed.
-For the exact workflow used one can refer to the code repository created for this work~\cite{code_repo}.
-
-The equivalent Hadoop-based pipeline was implemented with the Crossbow.
-The input data and the indexed genome reference were copied to Hadoop’s storage system~(HDFS) before starting the experiments.
-Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
-Unfortunately Crossbow's preprocessor is not written  in the MR manner, and thus cannot be run in a massively parallel way.
-Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted the Crossbow's preprocessor with our own, written in the MR-style, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by the Hadoop.}.
-To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing stage time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster.
-The scripts for preprocessing stage are publicly available~\cite{code_repo}.
-
-\section{Results}
-\label{sectionIII}
-
-\subsection{Accuracy of pipelines}
-Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, still we expect to capture and correctly identify the mutation.
-We tested the correctness just for the smallest dataset~I.
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized\cite{schneeberger}.
-
-\subsection{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of  the HPC approach we collected running time for dataset~I, as a function of a number of cores used, see Table~2.
-\begin{table}[htdp]
-\small
-\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC1 cluster. }
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}\\
-\hline
-	& aligning 	&	SNP calling\tablefootnote{We do not consider here special tricks how to parallelize the Samtools analysis by chromosome, as could be done, as exemplified here \url{http://www.biostars.org/p/48781/
-}}	&	total \\
-\hline
-1	&	38	&	37		&	75 \\
-2	&	19	&	36		&	55\\
-4	&	10	&	36		&	46\\
-8	&	5	&	36		&	41\\
-\end{tabular}
-\end{center}
-\label{table:hpctiming}
-\normalsize
-\end{table}
-As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, revealing almost no scaling at all.
-For the given datasets (I--V) the timings for alignment against the corresponding genomes, and SNP calling for  the HPC and Hadoop approaches were collected, Table~3.
-
-\begin{table}[htdp]
-\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.    Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-
-data, Gbases&	1.4&	3.5&	7.0 	&15.0 	&30.0 	&100.0 	&250 	\\
-			&	(I)			& (II, f.r.)	&	(II)		&(III, f.r.) &(III)&(IV) &(V)\\
-\hline
-Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
-\hline
-%Hadoop, 56 cores, Shared~TM&	25	&	24	&40		&	70&	150	&410&\\
-%\hline
-%Hadoop, 56 cores, ``bare-metal''  &	36	&	41	&53  	&63	&	92& 380	&\\
-%\hline
-HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
-%\hline
-%HPC2, 8 cores&72	&	&205	&&413&1765&\\
-%\hline
-%$F_{Hadoop}/F_{HPC1}$, Shared TM &	4.27		&1.75	&1.78	&	1.60	&	1.75	&	1.93&\\
-%\hline
-%$F_{Hadoop}/F_{HPC1}$, SSH TM &&		&1.74	&1.41	&1.28	&1.17	&\\
-%\hline
-%$F_{Hadoop}/F_{HPC2}$, ``bare-metal'' &&&&&&&\\
-
-%data, 	& \multicolumn{2}{c|}{Hadoop, 56 cores}		  	&  HPC			&	\multicolumn{2}{c|}{$F_{Hadoop}/F_{HPC}$}\\
-%Gbases 	&	 SSH~TM,	&	Shared~TM,	&	(8 cores), 	&	Shared TM	& 	SSH TM\\
-%&  minutes& minutes&  minutes& &\\
-
-\end{tabular}
-\end{center}
-\label{table:2}
-\end{table}%
-
-One of the attractive sides of using Hadoop is its almost linear scalability,~i.e.
-calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, {\it regardless} the scaling nature of the underlying program (Samtools, Bowtie etc.).
-Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
-Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
-To stress the linear nature of scaling, both sets of points were fit into linear polynomial using least squares method.
-
-\begin{figure}
-	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time depending on the dataset size for fixed size~($p=56$ cores) of Hadoop cluster on the private Cloud.
-The linear fit based on least squares method are provided, revealing almost linear scaling.}
-	\label{fig:fig1}
-\end{figure}
-
-\subsection{Comparing Hadoop and HPC efficiency for different dataset sizes}
-Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then,  that the larger the dataset is, more suitable Hadoop becomes compare to HPC.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
-$$F=T_{p}\times p,$$ 
-where~$T_{p}$ is a calculation time on~$p$ cores.
-Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC1}$, keeping in mind, the closer the ratio to the unity, the closer Hadoop’s efficiency approaches to that of HPC.
-
-\begin{figure}
-	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC1 correspondingly.
-The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.}
-	\label{fig:fig2}
-\end{figure}
-
-The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop-in-the-Cloud cluster, and a HPC1 node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC1}$ as a function of the {\it reciprocal} dataset size.
-The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical {\it infinite} dataset.
-As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data becomes larger.
-With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the {\it virtualized} environment is competitive with  the HPC approach, which is being run in ``bare metal'', for the datasets greater than $100$\,Gbases (Dataset IV), which is a usual size for human genome sequencing, giving the average sequencing depth of  about~$30x$.
-
- \subsection{Comparing the network communication efficiency for Hadoop and HPC approaches }
-Network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with the NAS or NFS attached storages.
-Namely, the effective bandwidth of the Hadoop network increases with the cluster size to increase~\cite{Sammer:2012}, opposite to that of the HPC network, where the cluster grows results in network saturation and performance depletion.
-We provide a comparison how HPC and Hadoop  network communication costs depend on the number of nodes involved, for a fixed dataset size, dataset~IV.
-
-Due to the fact of trivial parallelisation of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computation resources, i.e.
-split the initial data into chunks to proceed them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost {\it simultaneously} have to travel through the network, potentially causing traffic jams, therefore increasing the communication costs,~$T_{comm}$.
-
-There are several program packages for short reads alignment designed with the MPI support~\cite{pmap, gnumap}.
-Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}.
-However,  i.e.
-Pmap package had been proved to function poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising  memory exceptions.
-We wrote custom state-of-art bash script involving existing Unix utilities~\cite{repo} to use the  HPC2 cluster network as efficiently as possible,  and compared the network performance with the {\it standard} Hadoop HDFS approach.
-
-We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$. 
-Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases.
-For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to  able to be run in the MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS.
-We rewrote the code for the preprocessing stage for the Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, the~$T_{comm}$ involves the chunks travel from the sequencer delivery location to the local node scratches, where the actual mapping happens, and the travel of the aligned SAM files back to the delivery location over the network.
-
-Figure~\ref{fig:fig3} shows the~$T_{mapping}/T_{comm}$ ratio as a function of the reciprocal number of nodes~$1/N$ for Hadoop and HPC approaches.
-In its turn, the HPC approach is presented in two versions, which are based on a bit different strategies of the resource allocation.
-
-Lets start from description of the Hadoop results, filled circles.
-One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known  that the Bowtie, used for mapping, provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure 1.
-Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication in the preprocessing stage is.
-
-Lets now consider the curves for HPC2.
-The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in {\it parallel}, which are  {\it simultaneously} being pushed  to the local scratches of the nodes, allocated by SLURM  (open circles).
-One can see two distinct linear stretches, each with different tangent.
-One stretch is for the range from 4 to about 12 nodes the, and the another is from 12 up to 60.
-The former, horizontal stretch, is explained as for Hadoop~-- the more nodes is being involved the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: in the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX   is a set of RAIDs (Redundant Array of Inexpensive Disks) with data {\it striping}, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx const, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
-a linear dependency, which one can observe on the plot.
-The transition area between two modes~-- saturated and unsaturated~-- has the next origin: each HDD on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, what is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated  on the same rack, what is the default SLURM behaviour, therefore the caption reads~``HPC SLURM''.
-
-The scalability can be improved by overriding the default behaviour of the SLURM and allocating the nodes not from the same rack, but randomly from all available racks, ``HPC random'', (open squares in Figure~\ref{fig:fig3}).
-Allocating the nodes on random racks allows one to engage more nodes without network saturation.
-For our cluster we could go up to 30-35 nodes with perfect linear scaling.
-For the most resources been used (58 nodes) the deviation from a linear speed-up is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}  for the data.
-
-\begin{figure}
-	\small
-	\input{./figures/fig9/fig3.tex}
-	\normalsize
-	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC2 and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
+Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC2 and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
 Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one, where nodes are being allocated from random racks.
@@
 %linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
-Linear fit  done with the least-squares method.}
-	\label{fig:fig3}
-\end{figure}
-
-The threshold number of nodes in this strategy~($\approx35$) is driven by the saturating the uplink cable with the  throughput of~50Gbit/sec. 
-The proposed strategies aimed to gain the maximum from the existing resources show, that even properly adjusted and tuned, HPC approach sooner of later starts to suffer from the network saturation.
-
-\begin{table}[htdp]
-\caption{Timings  for mapping, and the ratio~$T_{mapping}/T_{comm}$  for HPC2 and  Hadoop clusters for Dataset IV.
-For ``HPC random'' approach data chunks have to be copied to the local scratches first, and  the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS.
-Hadoop needs to preprocess reads before the actual alignment stage, in order  to be able to operate in a MR manner, also resulting in  ``communication costs''.
-Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
- \multicolumn{3}{|c|}{Hadoop} & \multicolumn{3}{c|}{ HPC random} \\
- \hline
-
-Number nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$	&Number  nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$\\
-(cores)					&minutes		&							&(cores)			&minutes&\\
-\hline
-4(28)	&293.5	&2.33	&4(64)	&74.4	&3.89\\
-6(42)	&189.8	&2.19	&10(160)	&32.4	&3.76\\
-8(56)	&136.0	&2.23	&14(224)	&22.7	&3.77\\
-16(112)	&70.3	&1.96	&18(288)	&17.9	&3.78\\
-32(224)	&39.3	&2.20	&22(352)	&14.5	&3.79\\
-40(280)	&32.5	&2.15	&26(416)	&12.3	&3.77\\
-			&&&30(480)	&10.7	&3.73\\
-			&&&34(544)	&9.5	&3.45\\
-			&&&38(608)	&8.5	&3.16\\
-			&&&42(672)	&7.6	&2.96\\
-			&&&46(736)	&7.0	&2.55\\
-			&&&50(800)	&6.4	&2.65\\
-			&&&54(864)	&5.9	&2.34\\
-			&&&58(928)	&5.5	&2.12\\
-
-\end{tabular}
-\end{center}
-\label{table:4}
-\end{table}%
- At the same time, the HDFS keeps data locality, aiming to reduce the amount of communications, resulting less data move and, therefore, better scalability.
-Our Hadoop-in-the-Cloud cluster has no more~($\approx 40$) free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the exposed behaviour is a generic for Hadoop with HDFS.
-
-\subsection{Usability aspects}
-\label{subsectionIV_2}
-
-At the present moment a popular way to steer the bioinformatic pipelines on the HPC resources  is to use Galaxy\cite{galaxy}, which provides a Web-based graphical user interface~(GUI) to the bioinformatic programs, simplifying an experience for the end user.
-One of the alternatives for Hadoop could be the publically available Cloudgene\cite{cloudgene}, which seems to be a very light-weight, and flexible Web-based solution for serve~GUI for both public and private cloud, which we involved in our work.
-For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow, Figure~\ref{fig:fig4}.
-The most of data managing job is done automatically, and the results can be downloaded to the client machine.
-Modular structure allows one easy to modify the source code to integrate to the existing computing center’s architecture, Figure~\ref{fig:fig4}(a).
-For example the UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will stay locally, reducing amount of unnecessary risks.
-
-\begin{figure}
-	\begin{subfigure}[b]{0.6\textwidth}
-	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{UPPMAX-adapted Cloudgene: browsing the users home folder}
-	\end{subfigure}
- 	\begin{subfigure}[b]{0.4\textwidth}
-        	\includegraphics[width=\textwidth]{a.png}
-		\subcaption{Cloudgene: selecting a job to run}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.6\textwidth}
-		\includegraphics[width=\textwidth]{b.png}
-		\subcaption{Cloudgene: specifying job parameters}
-	\end{subfigure}
-	\caption{An example of a job setup with the Cloudgene, a web-based GUI wrapper, providing a smooth user experience even for novice users.}
-	\label{fig:fig4}
-\end{figure}
-
-\section{Conclusions}
-\label{sectionV}
-
-In this communication we brought together different approaches to analyze  the DNA-seq data in the bioinformatic study, in order to answer the question: which conditions a task should meet in order to make Hadoop a proper tool to use in favour of using HPC resources.
-
-To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow software were suggested.
-A state-of-art bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
-
-We concentrated ourselves on the Hadoop  in the private cloud installation, Hadoop-in-the-Cloud, orchestrated by the OpenNebula. 
-Picking the appropriate program settings we found out that DNA-seq dataset size larger than 100\,Gbases is suitable for Hadoop, with competitive  execution time compare to  that of HPC.
-The scaling graphs (Figure~\ref{fig:fig1}) confirms the known fact for almost linear scalability of Hadoop.
-The extrapolation to the infinite dataset size (Figure~\ref{fig:fig2}) revealed, that HPC, however, provides the results faster, given the same amount of resources, than the Hadoop-in-the-Cloud.
-
-Exploiting  the embarrassingly parallel nature of the short reads mapping we used custom state-of-art bash script to engage up to 58 nodes (928 cores) of the HPC2 cluster to  see the scaling relations between the ratio of the mapping time to the communication time as a function of a reciprocal number of nodes, Figure~\ref{fig:fig3}.
-Our results show that Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show  an example, how one can improve the performance of the HPC approach redefining the SLURM's default behaviour.
-
-Finally, existing and easily transformed publically available web-based GUI, like Cloudgene~(Figure~\ref{fig:fig4}),  gives an opportunity to perform bioinformatic analysis on Hadoop for those who are not experts in Linux world.
-Modular structure of Cloudgene gives cluster staff an easy way to adapt it for the specific needs.
-
-\section{Acknowledge}
-The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (UPPMAX) under Project p2013023.
-We also greatly appreciate   Pontus Freyhult and Peter Ankerst{\aa}l at  UPPMAX for explaining the effective storage usage.
-Great job was done by Jonas Hagberg at BILS, Stockholm, Sweden, by the adaptation of the Cloudgene to the local UPPMAX needs.
-
-\section{Supplementary material}
-
-\subsection{Used datasets}
-
-The datasets used in the paper are publicly available at:
-
-I: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0\\/data/reads/Schneeberger.2009/Schneeberger.2009.single\_end.gz}
-
-II: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads1.fq.gz, http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads2.fq.gz}
-
-III: \url{ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611084//SRR611084.sra, ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611085//SRR611085.sra}
-
-IV: artificial pair-ended dataset for {\it A.thaliana} created with the {\tt wgsim} program from the Samtools package.
-
-V: \url{http://www.ncbi.nlm.nih.gov/sra/SRX148888}
-
-\subsection{Reference genomes}
-\begin{itemize}
-\item TAIR10 for datasets II-IV \url{ftp://ftp.arabidopsis.org/home/tair/Sequences/whole\_chromosomes/*.fas}
-\item TAIR8 for dataset I \url{ftp://ftp.arabidopsis.org/home/tair/Genes/TAIR8\_genome\_release/}
-\item H.sapiens, NCBI v37 \url{ftp://ftp.ccb.jhu.edu/pub/data/bowtie\_indexes/h\_sapiens\_37\_asm.ebwt.zip}
-\end{itemize}
-
-\subsection{Description of computational facilities}
-
-\begin{enumerate}
-\item  HPC1:
-The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
-The data and reference genomes were read and written to a parallel shared storage system.
-
-\item 
-HPC2:
-Multinode short read mapping was performed on the Milou cluster\cite{milouCluster} , equipped with dual 8cores CPUs Intel Xeon E5-2660, ( 2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache.), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
-
-\item Storage: 
-Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600 storage boxes and an additional node for metadata handling.
-In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
-It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
-
-\item Our main Hadoop test platform was deployed on the private cloud at UPPMAX, using the OpenNebula~\cite{opennebula} cloud management system.
-Each node in this deployment was equipped with dual quad-core CPUs (Intel Xeon 5420; clock frequency of 2.50GHz\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
-The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-
-\end{enumerate}
-
-\bibliography{text1}
-\bibliographystyle{unsrt}
-
+Linear fit  done with the least-squares method.
 \end{document}

### 2014-01-15 figures/fig1/makefig1.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+gnuplot script.gnuplot1 && pdflatex text.tex && open text.pdf

### 2014-01-15 figures/fig2/makefig1.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+gnuplot script.gnuplot2 && pdflatex text.tex && open text.pdf

### 2014-01-15 text1.tex
Updated title and abstract

@@
 
-\title{Hadoop Or Not Hadoop}
+\title{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)}
 \author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth}
@@
 Hadoop, emerged in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-We here propose and report the quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases.
-We modified and wrote new pieces of code in order to substantially increase the performance of existing software.
-Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches.
-Our results show that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines  become performance-competitive with a canonical high-performance cluster solutions.
-
+We here report a quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases.
+In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts.
+Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution.
+As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in future biological data analysis.
 }

### 2014-01-15 text1.tex
Brushed up text in introduction.

@@
 Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 day~\cite{illumina}.
-Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} HPC center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
+As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 days~\cite{illumina}.
+Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
 
@@
 A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Most of the bioinformatic programs (tools) are written for the Linux operating system.
-Some of the most widespread software tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000}, like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
+Most of the bioinformatic programs are written for the Linux operating system.
+Some of the most widespread tools \footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
 
@@
 
-Most common  way to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.
+The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.
 For the tools which do not support the OMP natively, i.e.
-the Samtools, a variant calling can be parallelised specifying the option, creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
-The fact of great importance, however, is that the multi-core approach does not improve the performance of operations that are limited by {\it disk} or {\it network } throughputs, motivating to split  the dataset and  use the  multi-node parallelisation.
+Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
+Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughput, motivating to split  the dataset and  use multi-node parallelisation.
+Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
-The Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
-
-Another common way to introduce the parallelisation in Linux systems is to use Bash scripting e.g.
-involve  already existing utilities and cluster tools, split the data into chunks to deliver and proceed them  on the separate nodes, merging the results afterwards.
-This kind of solution benefiting  from both MPI-like and OMP, is involved in our work as well.
-It provides great performance, but usually it is tightly binded to the local computational cluster and network architectures.
+Another common way to introduce parallelisation in Linux systems is to use Bash scripting e.g.
+involve already existing utilities and cluster tools, split the data into chunks and process them on the separate nodes, and merging the results afterwards.
+This kind of solution benefits from both MPI and OMP and provides good performance, but the developments requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.
 
 %Introduce Hadoop, MR, and HDFS
+%Properties of Hadoop and HDFS- data localization and programming model etc
 The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running task in a {\it massively} parallel way.
-This paradigm, however, shifts the focus from the best performance to scalability, suited for managing  huge datasets of size  of terabytes~\cite{lin2010}.
+This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
 The prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. 
-For the purpose of this work, we focus on the Cloudera Hadoop MR with the Hadoop Distributed File System~(HDFS).
-
-%Properties of Hadoop and HDFS- data localization and programming model etc
 The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
-HDFS is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto each node’s local disks, therefore effectively moving the computation to the data and hence reducing network traffic.
-The HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
+Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively moving the computation to the data and reducing network traffic.
+HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
 
 %Manuscript focus: When is Hadoop currently an appealing alternative
+%Introduce what we did in the manuscript.
 In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
-Since Hadoop is written in Java, which is slower than the standard~HPC programming languages, like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
-
-%Introduce what we did in the manuscript.
-We use five datasets with short reads of different size to align them against the reference genome, and call the  variances.
-For one dataset we made a check to identify the actual mutation to make sure that HPC and Hadoop give the same answers.
-From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
-For the classical HPC DNA-seq analysis programs we developed a general-use bash scripts, utilizing multiple nodes for short reads alignment and exploiting the network fully, speeding up the calculations.
+Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
+We use five datasets with short reads of different sizes, align them against the reference genome, and call the variances.
+%From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
+For the classical HPC DNA-seq analysis programs we developed a set of bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
 The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.

### 2014-01-17 text1.tex
Updated text in Methods section

@@
 \subsection{Datasets}
-We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals, Table~1.
-Data for datasets I-III,V were generated using Illumina/(HiSeq) sequencing platforms.
+We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
+Data for datasets I-III and V were obtained using Illumina/(HiSeq) sequencing platforms.
 Further information about the datasets is provided in the Supplementary material section.
@@
 \small
+\footnotesize
 \caption{Datasets used}
@@
 
-\subsection{Computational resources}
-To run the HPC analysis pipeline nodes of  the computational clusters at UPPMAX, equipped with dual 4-core~(HPC1) and 8-core processors~(HPC2) were used.
-Data and reference genomes were read and written to a parallel shared storage system.
-The main Hadoop test platform (Hadoop-in-the-Cloud) was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} as a cloud manager.
-Each node in this deployment was equipped with dual 4-core CPUs.
-The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-For details on computational resources, see Supplementary material.
-
 \subsection{Analysis pipelines}
-We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on HPC.
+We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on regular HPC with a batch system (hereafter referred to as HPC).
 Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage.
@@
 \item HPC approach
-\subitem short read alignment: Bowtie~ver.~0.12.8
+\subitem Short read alignment: Bowtie~ver.~0.12.8
 \subitem SNP calling: Samtool~ver.~0.1.19
 \item Hadoop approach: Crossbow~\cite{crossbow}
-\subitem short read alignment: Bowtie~ver.~0.12.8
+\subitem Short read alignment: Bowtie~ver.~0.12.8
 \subitem SNP calling: SOAPsnp~1.02~\cite{soapsnp}
@@
 In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools.
-The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained in~8 times faster than on a single core.
+The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core.
 Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
@@
 
-The equivalent Hadoop-based pipeline was implemented with the Crossbow.
-The input data and the indexed genome reference were copied to Hadoop’s storage system~(HDFS) before starting the experiments.
+The equivalent Hadoop-based pipeline was implemented with Crossbow.
+The input data and the indexed genome reference were copied to Hadoop's storage system~(HDFS) before starting the experiments.
 Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
-Unfortunately Crossbow's preprocessor is not written  in the MR manner, and thus cannot be run in a massively parallel way.
+Unfortunately Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way.
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted the Crossbow's preprocessor with our own, written in the MR-style, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by the Hadoop.}.
-To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing stage time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster.
-The scripts for preprocessing stage are publicly available~\cite{code_repo}.
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by Hadoop.}.
+To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster.
+The scripts for the preprocessing stage are publicly available~\cite{code_repo}.
+
+\subsection{Computational resources}
+To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 4 dual-core CPUs.
+Data and reference genomes were stored on a parallel shared storage system.
+The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager.
+The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
+For details on computational resources, see Supplementary material.
 

### 2014-01-17 text1.tex
Text massaging up to section 3.4

@@
 \usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
+%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
+\geometry{a4paper}                   		% ... or a4paper or a5paper or ... 
 %\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
@@
 \usepackage{caption}
+\usepackage[format=hang,font=small,labelfont=bf]{caption}
 
@@
 \subsection{Accuracy of pipelines}
-Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, still we expect to capture and correctly identify the mutation.
+Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, but still we expect them to capture and correctly identify the mutation.
 We tested the correctness just for the smallest dataset~I.
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized\cite{schneeberger}.
+The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized by both applications\cite{schneeberger}.
+[IS THIS REFERENCE CORRECT AND IN RIGHT PLACE?]
 
 \subsection{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of  the HPC approach we collected running time for dataset~I, as a function of a number of cores used, see Table~2.
-\begin{table}[htdp]
+To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:hpctiming}).
+As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, showing almost no scaling at all.
+For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for  the HPC and Hadoop approaches were collected (Table~\ref{table:2}).
+
+One of the attractive sides of using Hadoop is its almost linear scalability,~i.e.
+calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie etc.).
+Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
+Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
+To stress the linear nature of scaling, both sets of points were fit to linear polynomial using the least squares method.
+
+\begin{table}[b]
 \small
@@
 \end{table}
-As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, revealing almost no scaling at all.
-For the given datasets (I--V) the timings for alignment against the corresponding genomes, and SNP calling for  the HPC and Hadoop approaches were collected, Table~3.
 
@@
 
-One of the attractive sides of using Hadoop is its almost linear scalability,~i.e.
-calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, {\it regardless} the scaling nature of the underlying program (Samtools, Bowtie etc.).
-Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
-Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
-To stress the linear nature of scaling, both sets of points were fit into linear polynomial using least squares method.
-
 \begin{figure}
@@
 Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then,  that the larger the dataset is, more suitable Hadoop becomes compare to HPC.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
+One can propose then that the larger the dataset is, the more suitable Hadoop becomes compare to the HPC approach.
+In order to compare the ''suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
 $$F=T_{p}\times p,$$ 
-where~$T_{p}$ is a calculation time on~$p$ cores.
-Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC1}$, keeping in mind, the closer the ratio to the unity, the closer Hadoop’s efficiency approaches to that of HPC.
+where~$T_{p}$ is the calculation time on~$p$ cores.
+Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC}$, keeping in mind that the closer the ratio to the unity, the closer Hadoop's efficiency is to that of the HPC approach.
 
@@
 
-The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop-in-the-Cloud cluster, and a HPC1 node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC1}$ as a function of the {\it reciprocal} dataset size.
-The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical {\it infinite} dataset.
-As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data becomes larger.
-With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the {\it virtualized} environment is competitive with  the HPC approach, which is being run in ``bare metal'', for the datasets greater than $100$\,Gbases (Dataset IV), which is a usual size for human genome sequencing, giving the average sequencing depth of  about~$30x$.
+The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
+The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical infinite dataset.
+As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data is larger.
+With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the virtualized environment of a private cloud, is competitive with  the HPC approach which is being run on ''bare metal''  for the datasets greater than $100$\,Gbases (IV), which is a usual size for human genome sequencing giving the average sequencing depth of  about~$30x$.
 

### 2014-01-17 text1.tex
text edits up to Usability aspects

@@
 
-\subsection{Comparing Hadoop and HPC efficiency for different dataset sizes}
+\subsection{Comparing Hadoop and HPC runtime efficiency for different dataset sizes}
 Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}.
@@
 
- \subsection{Comparing the network communication efficiency for Hadoop and HPC approaches }
-Network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with the NAS or NFS attached storages.
-Namely, the effective bandwidth of the Hadoop network increases with the cluster size to increase~\cite{Sammer:2012}, opposite to that of the HPC network, where the cluster grows results in network saturation and performance depletion.
-We provide a comparison how HPC and Hadoop  network communication costs depend on the number of nodes involved, for a fixed dataset size, dataset~IV.
+ \subsection{Comparing the network communication efficiency for the Hadoop and HPC approaches }
+Network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
+The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
+We provide a comparison on how HPC and Hadoop network communication costs depend on the number of nodes involved, for a fixed dataset size (dataset~IV).
 
-Due to the fact of trivial parallelisation of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computation resources, i.e.
-split the initial data into chunks to proceed them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost {\it simultaneously} have to travel through the network, potentially causing traffic jams, therefore increasing the communication costs,~$T_{comm}$.
+Due to the trivially parallelisation nature of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, i.e.
+split the initial data into chunks to process them independently.
+Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network and thereby increasing the communication costs,~$T_{comm}$.
 
-There are several program packages for short reads alignment designed with the MPI support~\cite{pmap, gnumap}.
+There are several program packages for short read alignment designed with MPI support~\cite{pmap, gnumap}.
 Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}.
 However,  i.e.
-Pmap package had been proved to function poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising  memory exceptions.
-We wrote custom state-of-art bash script involving existing Unix utilities~\cite{repo} to use the  HPC2 cluster network as efficiently as possible,  and compared the network performance with the {\it standard} Hadoop HDFS approach.
-
-We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$. 
+Pmap package functions poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising memory exceptions.
+We implemented a highly optimized bash script making use of standard Unix utilities to use the  HPC cluster network as efficiently as possible~\cite{repo},  and compared the network performance with the standard Hadoop HDFS approach.
+We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
 Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases.
-For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to  able to be run in the MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS.
-We rewrote the code for the preprocessing stage for the Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, the~$T_{comm}$ involves the chunks travel from the sequencer delivery location to the local node scratches, where the actual mapping happens, and the travel of the aligned SAM files back to the delivery location over the network.
+For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS.
+We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
+For the HPC approach, the~$T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.
 
 Figure~\ref{fig:fig3} shows the~$T_{mapping}/T_{comm}$ ratio as a function of the reciprocal number of nodes~$1/N$ for Hadoop and HPC approaches.
-In its turn, the HPC approach is presented in two versions, which are based on a bit different strategies of the resource allocation.
+The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
 
-Lets start from description of the Hadoop results, filled circles.
-One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known  that the Bowtie, used for mapping, provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure 1.
-Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication in the preprocessing stage is.
+\textit{Hadoop results (filled circles)}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
+It is known that the Bowtie provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}.
+Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication is in the preprocessing stage.
 
-Lets now consider the curves for HPC2.
-The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in {\it parallel}, which are  {\it simultaneously} being pushed  to the local scratches of the nodes, allocated by SLURM  (open circles).
-One can see two distinct linear stretches, each with different tangent.
-One stretch is for the range from 4 to about 12 nodes the, and the another is from 12 up to 60.
-The former, horizontal stretch, is explained as for Hadoop~-- the more nodes is being involved the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: in the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX   is a set of RAIDs (Redundant Array of Inexpensive Disks) with data {\it striping}, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx const, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
-a linear dependency, which one can observe on the plot.
-The transition area between two modes~-- saturated and unsaturated~-- has the next origin: each HDD on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, what is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated  on the same rack, what is the default SLURM behaviour, therefore the caption reads~``HPC SLURM''.
+\textit{HPC results}: The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in parallel, which are simultaneously being pushed to the local scratches of the nodes, allocated by SLURM  (open circles).
+One can see two distinct linear stretches, each with a different slope.
+One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
+The former (horizontal stretch) is explained as for Hadoop~-- the more nodes being involved the faster the chunks are being distributed.
+The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx const, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
+a linear dependency, which one can observe in the plot.
+The transition area between the two modes~-- saturated and unsaturated~-- has the next origin: each hard disk drive on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
+The nodes are being allocated  on the same rack, which is the default SLURM behaviour.
 
-The scalability can be improved by overriding the default behaviour of the SLURM and allocating the nodes not from the same rack, but randomly from all available racks, ``HPC random'', (open squares in Figure~\ref{fig:fig3}).
+The scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (''HPC random'', open squares in Figure~\ref{fig:fig3}).
 Allocating the nodes on random racks allows one to engage more nodes without network saturation.
 For our cluster we could go up to 30-35 nodes with perfect linear scaling.
-For the most resources been used (58 nodes) the deviation from a linear speed-up is~$\approx 7\%$ i.e.
+For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
 $5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}  for the data.
+The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
+The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approach suffer from the network saturation at higher number of nodes.
 
@@
 	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC2 and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
-Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one, where nodes are being allocated from random racks.
+Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
 %One  can see the differences in the Hadoop and HPC communication patterns: for Hadoop the ratio is almost constant, while for HPC cluster there are two defined regions.
@@
 
-The threshold number of nodes in this strategy~($\approx35$) is driven by the saturating the uplink cable with the  throughput of~50Gbit/sec. 
-The proposed strategies aimed to gain the maximum from the existing resources show, that even properly adjusted and tuned, HPC approach sooner of later starts to suffer from the network saturation.
-
 \begin{table}[htdp]
-\caption{Timings  for mapping, and the ratio~$T_{mapping}/T_{comm}$  for HPC2 and  Hadoop clusters for Dataset IV.
-For ``HPC random'' approach data chunks have to be copied to the local scratches first, and  the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS.
-Hadoop needs to preprocess reads before the actual alignment stage, in order  to be able to operate in a MR manner, also resulting in  ``communication costs''.
+\caption{Timings  for mapping, and the ratio~$T_{mapping}/T_{comm}$  for HPC and  Hadoop clusters for Dataset IV.
+For ``HPC random'' approach data chunks have to be copied to the local scratch disks first, and the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS and hence do not need data staging.
+Hadoop however needs to preprocess reads before the actual alignment stage, in order  to be able to operate in MR manner, also resulting in what we term ''communication costs''.
 Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
@@
 
-At the present moment a popular way to steer the bioinformatic pipelines on the HPC resources  is to use Galaxy\cite{galaxy}, which provides a Web-based graphical user interface~(GUI) to the bioinformatic programs, simplifying an experience for the end user.
+At the present moment a popular way to construct and execute bioinformatic pipelines on HPC resources  is to install and use the Galaxy\cite{galaxy} software, which provides a Web-based graphical user interface~(GUI) to the bioinformatic programs, simplifying an experience for the end user.
 One of the alternatives for Hadoop could be the publically available Cloudgene\cite{cloudgene}, which seems to be a very light-weight, and flexible Web-based solution for serve~GUI for both public and private cloud, which we involved in our work.

### 2014-01-19 text1.tex
Updated text in Usability aspects

@@
 
-At the present moment a popular way to construct and execute bioinformatic pipelines on HPC resources  is to install and use the Galaxy\cite{galaxy} software, which provides a Web-based graphical user interface~(GUI) to the bioinformatic programs, simplifying an experience for the end user.
-One of the alternatives for Hadoop could be the publically available Cloudgene\cite{cloudgene}, which seems to be a very light-weight, and flexible Web-based solution for serve~GUI for both public and private cloud, which we involved in our work.
-For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow, Figure~\ref{fig:fig4}.
+A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying an experience for the end user.
+One of the alternatives for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private cloud.
+We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}(a)).
+For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow, Figure~.
 The most of data managing job is done automatically, and the results can be downloaded to the client machine.
@@
 \begin{figure}
+ 	\begin{subfigure}[b]{0.4\textwidth}
+        	\includegraphics[width=\textwidth]{a.png}
+		\subcaption{}
+	\end{subfigure}
+	\hspace{0.5cm}
 	\begin{subfigure}[b]{0.6\textwidth}
 	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{UPPMAX-adapted Cloudgene: browsing the users home folder}
+		\subcaption{}
 	\end{subfigure}
- 	\begin{subfigure}[b]{0.4\textwidth}
-        	\includegraphics[width=\textwidth]{a.png}
-		\subcaption{Cloudgene: selecting a job to run}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.6\textwidth}
-		\includegraphics[width=\textwidth]{b.png}
-		\subcaption{Cloudgene: specifying job parameters}
-	\end{subfigure}
-	\caption{An example of a job setup with the Cloudgene, a web-based GUI wrapper, providing a smooth user experience even for novice users.}
+%	\begin{subfigure}[b]{0.6\textwidth}
+%		\includegraphics[width=\textwidth]{b.png}
+%		\subcaption{Cloudgene: specifying job parameters}
+%	\end{subfigure}
+	\caption{An example of a job setup with the graphical Hadoop front-end Cloudgene, providing a smooth user experience even for novice users.
+\textbf{a)} Pipeline selection - in our case containing the Crossbow pipeline.
+\textbf{b)} The UPPMAX-adapted functionality to browsing and import data from the users home folder in the shared file system.}
 	\label{fig:fig4}

### 2014-01-19 text1.tex
Text updates to conclusions and acknowledgements

@@
 
-In this communication we brought together different approaches to analyze  the DNA-seq data in the bioinformatic study, in order to answer the question: which conditions a task should meet in order to make Hadoop a proper tool to use in favour of using HPC resources.
+In this report we have described two approaches for high-performance analysis of DNA sequencing data; one based on regular HPC using a batch system and one based on Hadoop.
+We show that Hadoop installed on a private cloud is an appealing solution both on terms of runtime performance and also in terms of usability.
+A key result is the data size where Hadoop is favorable to regular HPC batch systems, and we developed highly optimized pipelines for both scenarios.
+We found that dataset sizes larger than 100\,Gbases is where Hadoop excel over HPC, and also that Hadoop shows almost linear scalability (Figure~\ref{fig:fig1}).
+Extrapolation to infinite dataset size (Figure~\ref{fig:fig2}) reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
 
-To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow software were suggested.
+%To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow software were suggested.
 A state-of-art bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
 
-We concentrated ourselves on the Hadoop  in the private cloud installation, Hadoop-in-the-Cloud, orchestrated by the OpenNebula. 
-Picking the appropriate program settings we found out that DNA-seq dataset size larger than 100\,Gbases is suitable for Hadoop, with competitive  execution time compare to  that of HPC.
-The scaling graphs (Figure~\ref{fig:fig1}) confirms the known fact for almost linear scalability of Hadoop.
-The extrapolation to the infinite dataset size (Figure~\ref{fig:fig2}) revealed, that HPC, however, provides the results faster, given the same amount of resources, than the Hadoop-in-the-Cloud.
+%We concentrated ourselves on the Hadoop  in the private cloud installation, Hadoop-in-the-Cloud, orchestrated by the OpenNebula. 
+%Picking the appropriate program settings we found out that DNA-seq dataset size larger than 100\,Gbases is suitable for Hadoop, with competitive  execution time compare to  that of HPC.
+%The scaling graphs (Figure~\ref{fig:fig1}) confirms the known fact for almost linear scalability of Hadoop.
+%The extrapolation to the infinite dataset size (Figure~\ref{fig:fig2}) revealed, that HPC, however, provides the results faster, given the same amount of resources, than the Hadoop-in-the-Cloud.
 
-Exploiting  the embarrassingly parallel nature of the short reads mapping we used custom state-of-art bash script to engage up to 58 nodes (928 cores) of the HPC2 cluster to  see the scaling relations between the ratio of the mapping time to the communication time as a function of a reciprocal number of nodes, Figure~\ref{fig:fig3}.
+Exploiting the embarrassingly parallel nature of the short reads mapping we used a highly optimized bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes (Figure~\ref{fig:fig3}).
 Our results show that Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show  an example, how one can improve the performance of the HPC approach redefining the SLURM's default behaviour.
+In addition we show how performance of the HPC approach can be improved redefining the queueing system's default behaviour.
+Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene), provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
 
-Finally, existing and easily transformed publically available web-based GUI, like Cloudgene~(Figure~\ref{fig:fig4}),  gives an opportunity to perform bioinformatic analysis on Hadoop for those who are not experts in Linux world.
-Modular structure of Cloudgene gives cluster staff an easy way to adapt it for the specific needs.
-
-\section{Acknowledge}
-The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (UPPMAX) under Project p2013023.
-We also greatly appreciate   Pontus Freyhult and Peter Ankerst{\aa}l at  UPPMAX for explaining the effective storage usage.
-Great job was done by Jonas Hagberg at BILS, Stockholm, Sweden, by the adaptation of the Cloudgene to the local UPPMAX needs.
+\section{Acknowledgements}
+The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
+We thank Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage usage.
+We also thank Jonas Hagberg (BILS, Stockholm, Sweden), for implementing the Cloudgene extensions to import data from UPPMAX filesystem.
 

### 2014-02-14 text1.tex
Version on 14.02.14

@@
 Most of the bioinformatic programs are written for the Linux operating system.
-Some of the most widespread tools \footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
+Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
 
@@
 Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
-Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughput, motivating to split  the dataset and  use multi-node parallelisation.
+Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating to split  the dataset and  use multi-node parallelisation.
+
 Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
-Another common way to introduce parallelisation in Linux systems is to use Bash scripting e.g.
+Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting e.g.
 involve already existing utilities and cluster tools, split the data into chunks and process them on the separate nodes, and merging the results afterwards.
-This kind of solution benefits from both MPI and OMP and provides good performance, but the developments requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.
+This kind of solution benefits from both MPI-like  and OMP and provides good performance, but the development requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.
 
@@
 We use five datasets with short reads of different sizes, align them against the reference genome, and call the variances.
-%From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
-For the classical HPC DNA-seq analysis programs we developed a set of bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
+For the existing Hadoop  software we propose modifications, in order to benefit fully from massively parallel nature of MR computations. 
+For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
 The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.
@@
 \subsection{Analysis pipelines}
-We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on regular HPC with a batch system (hereafter referred to as HPC).
+We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and one on regular HPC with a batch processing system (hereafter referred to as HPC).
 Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage.
@@
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by Hadoop.}.
-To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster.
-The scripts for the preprocessing stage are publicly available~\cite{code_repo}.
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation.
 
 \subsection{Computational resources}
-To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 4 dual-core CPUs.
-Data and reference genomes were stored on a parallel shared storage system.
+To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 8 dual-core CPUs.
+Data and reference genomes were stored on a parallel shared storage system\cite{gulo}.
 The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager.
-The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
+The cluster was set up with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}.
 For details on computational resources, see Supplementary material.
@@
 
+\subsection{Modified preprocessing stage}
+A native preprocessing stage in the Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster.
+Data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}.
+The only way to introduce parallelisations in the native preprocessing stage~-- is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
+This, however, is lacking the massively parallel way of treating the data.
+
+Being  pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both~--  storage and  computers,  
+therefore considered the data already been downloaded to the storage Hadoop can access.
+We rewrote the preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
+The demand is that the FASTQ data are BZIP2 archived, and  accessible by Hadoop.
+For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested to the HDFS.
+The BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}.
+Also the BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e.
+for the developer there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
+
+The Hadoop streaming library offers a possibility to write  Mapper and Reducer for Hadoop jobs in any  programming language.
+Our Python scripts efficiently process  short reads, produced by  Illumina sequencing platform of different versions, as well as the FASTQ files converted from the SRA format, NCBI Sequence Read Archive.
+The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
+To ensure that the both forward and reverse read of the same pair will end up on the same Reducer, the secondary Mapper key-sorting mechanisms were involved.
+
+In order to compare the native preprocessor and the proposed one, the data was put on the HDFS in beforehand, to eliminate possible network delays i.e. while downloading data from some Amazon cloud.
+Table~\ref{table:preprocess} illustrates the benefits of our approach.
+\begin{table}[htdp]
+\small
+\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on~$p=56$ cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node, $p=16$ cores.}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|}
+dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
+\hline
+Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1$	\\
+Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
+BASH, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
+\end{tabular}
+\end{center}
+\label{table:preprocess}
+\normalsize
+\end{table}%
+To complete the comparison the same preprocessing was performed with the  BASH scripting against the data located on the HPC storage.
+The script\footnote{\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}} ran on a single HPC node, utilizes  multiple cores, with the limiting factor being the  HDD IO performance, which effectively leaves with 5-7 cores of 16.
+
 \subsection{Accuracy of pipelines}
@@
 We tested the correctness just for the smallest dataset~I.
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized by both applications\cite{schneeberger}.
-[IS THIS REFERENCE CORRECT AND IN RIGHT PLACE?]
+The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.
 
 \subsection{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:hpctiming}).
-As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, showing almost no scaling at all.
-For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for  the HPC and Hadoop approaches were collected (Table~\ref{table:2}).
+To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}).
+As one can see the aligning process with the Bowtie scales fairly good, but the SNP calling part is a bottleneck,  scaling worse than the Bowtie, and consuming progressively larger part of the total calculation time
+%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
+During its sorting phase the lack of RAM results in spilling to the HDD.
+For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM.
+For our tasks we used the default Samtool's settings.}.
+For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for  the HPC and Hadoop approaches were collected (Table~\ref{table:4}).
 
-One of the attractive sides of using Hadoop is its almost linear scalability,~i.e.
+One of the attractive sides of using Hadoop~MR is its almost linear scalability,~i.e.
 calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie etc.).
-Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
+Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
 Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
@@
 
-\begin{table}[b]
+\begin{table}[htb]
 \small
-\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC1 cluster. }
+\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC cluster. }
 \begin{center}
-\begin{tabular}{|c|c|c|c|}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}\\
+\begin{tabular}{|c|c|c|c|ccc|}
+$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c|}{speed-ups} \\
 \hline
-	& aligning 	&	SNP calling\tablefootnote{We do not consider here special tricks how to parallelize the Samtools analysis by chromosome, as could be done, as exemplified here \url{http://www.biostars.org/p/48781/
-}}	&	total \\
+	& mapping 	&	SNP calling\tablefootnote{We  used here special tricks how to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781/
+}}	&	total  &total & map& SNP call\\
 \hline
-1	&	38	&	37		&	75 \\
-2	&	19	&	36		&	55\\
-4	&	10	&	36		&	46\\
-8	&	5	&	36		&	41\\
+1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
+2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
+4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
+6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
+8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
+10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
+12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
+14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
+16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
 \end{tabular}
 \end{center}
-\label{table:hpctiming}
+\label{table:3}
 \normalsize
@@
 \begin{table}[htdp]
-\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.    Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.}
+\small
+
+\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.
+Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.
+Big deviation for the HPC is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.}
 \begin{center}
@@
 
-data, Gbases&	1.4&	3.5&	7.0 	&15.0 	&30.0 	&100.0 	&250 	\\
-			&	(I)			& (II, f.r.)	&	(II)		&(III, f.r.) &(III)&(IV) &(V)\\
+data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
+				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
 \hline
 Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
-\hline
-%Hadoop, 56 cores, Shared~TM&	25	&	24	&40		&	70&	150	&410&\\
-%\hline
-%Hadoop, 56 cores, ``bare-metal''  &	36	&	41	&53  	&63	&	92& 380	&\\
-%\hline
+Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
 HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
-%\hline
-%HPC2, 8 cores&72	&	&205	&&413&1765&\\
-%\hline
-%$F_{Hadoop}/F_{HPC1}$, Shared TM &	4.27		&1.75	&1.78	&	1.60	&	1.75	&	1.93&\\
-%\hline
-%$F_{Hadoop}/F_{HPC1}$, SSH TM &&		&1.74	&1.41	&1.28	&1.17	&\\
-%\hline
-%$F_{Hadoop}/F_{HPC2}$, ``bare-metal'' &&&&&&&\\
-
-%data, 	& \multicolumn{2}{c|}{Hadoop, 56 cores}		  	&  HPC			&	\multicolumn{2}{c|}{$F_{Hadoop}/F_{HPC}$}\\
-%Gbases 	&	 SSH~TM,	&	Shared~TM,	&	(8 cores), 	&	Shared TM	& 	SSH TM\\
-%&  minutes& minutes&  minutes& &\\
+HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\
 
@@
 \end{center}
-\label{table:2}
+\label{table:4}
+\normalsize
 \end{table}%
@@
 	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time depending on the dataset size for fixed size~($p=56$ cores) of Hadoop cluster on the private Cloud.
-The linear fit based on least squares method are provided, revealing almost linear scaling.}
+	\caption{Calculation time depending on the  size for selected datasets for~$p=56$ cores Hadoop cluster on the private Cloud.
+The linear fit based on least squares method are provided, revealing  linear scaling.
+Datasets are marked with roman numerals.}
 	\label{fig:fig1}
@@
 One can propose then that the larger the dataset is, the more suitable Hadoop becomes compare to the HPC approach.
-In order to compare the ''suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
+In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
 $$F=T_{p}\times p,$$ 
 where~$T_{p}$ is the calculation time on~$p$ cores.
-Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC}$, keeping in mind that the closer the ratio to the unity, the closer Hadoop's efficiency is to that of the HPC approach.
-
+Using the data from Table~\ref{table:4},  we plot the ratio $F_{Hadoop}/F_{HPC}$, keeping in mind that the closer the ratio to the unity, the closer Hadoop's efficiency is to that of the HPC approach.
 \begin{figure}
 	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC1 correspondingly.
-The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.}
+	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=16$ cores for Hadoop and HPC correspondingly.
+The points are fit in a curve using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.
+Datasets are marked with roman numerals.}
 	\label{fig:fig2}
 \end{figure}
-
-The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
+The curve  in~Figure~\ref{fig:fig2}, based on Table~\ref{table:4}, is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=16$), and displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
 The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical infinite dataset.
 As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data is larger.
-With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the virtualized environment of a private cloud, is competitive with  the HPC approach which is being run on ''bare metal''  for the datasets greater than $100$\,Gbases (IV), which is a usual size for human genome sequencing giving the average sequencing depth of  about~$30x$.
+The parabolic extrapolation for  the ratio reads~$1.70\pm0.01$, meaning that Hadoop running even  in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with  the HPC approach which is being run on ``bare metal''  on the modern hardware for the datasets greater than $100$\,Gbases (dataset~IV), which is a usual size for human genome sequencing giving the average sequencing depth of  about~$30x$.
+
+The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
+Taking into account  a {\it linear} scaling for the Hadoop approach for those datasets, Figure~\ref{fig:fig1}, one can conclude that HPC approach scales worse than linearly\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
+being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
+We suspect the main reason for such a behavior is the fact that for large datasets (starting from dataset~III in our case) the SNP calling routine with the Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
+If lacking  RAM, the Samtools spilling to HDD, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and  time expensive IO on a high rate.
+The Table~\ref{table:5} shows the role of the amount of  allocated RAM on the Samtools sort timing, as a most time expensive,  for the dataset~IV, starting from 8GB up to 80GB, performing on a single HPC node.
+\begin{table}[htdp]
+\small
+
+\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58Gb BAM file).
+The measurements are done on a single HPC node with $p=16$ cores.
+Averaging is done over 5 independent simulations.
+The RAM amount is given for the all  16 cores.
+}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+RAM, GB		&	8			&		16			&			32		& 		64				\\
+timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
+\end{tabular}
+\end{center}
+\label{table:5}
+\normalsize
+\end{table}%
+A large uncertainty for  small amount of RAM we believe is due to other users' load  on the cluster network.
+As the  result a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling  on much economic hardware on linear time  with just 12GB RAM per VM node.
 
@@
 split the initial data into chunks to process them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network and thereby increasing the communication costs,~$T_{comm}$.
+Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
 
@@
 Pmap package functions poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising memory exceptions.
-We implemented a highly optimized bash script making use of standard Unix utilities to use the  HPC cluster network as efficiently as possible~\cite{repo},  and compared the network performance with the standard Hadoop HDFS approach.
+We implemented a highly optimized Bash script making use of standard Unix utilities to use the  HPC cluster network as efficiently as possible~\cite{repo},  and compared the network performance with the standard Hadoop HDFS approach.
 We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
@@
 We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, the~$T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.
+For the HPC approach, the~$T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}} files back to the delivery location over the network.
 
@@
 \textit{Hadoop results (filled circles)}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that the Bowtie provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}.
-Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication is in the preprocessing stage.
+It is known that the Bowtie provides almost  linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}.
+Since the ratio~$T_{mapping}/T_{comm}\approx \mbox{const}$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication is in the preprocessing stage.
 
@@
 The former (horizontal stretch) is explained as for Hadoop~-- the more nodes being involved the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx const, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
+The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
 a linear dependency, which one can observe in the plot.
@@
 
-The scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (''HPC random'', open squares in Figure~\ref{fig:fig3}).
+The scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
 Allocating the nodes on random racks allows one to engage more nodes without network saturation.
@@
 	\normalsize
-	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC2 and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
+	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
 Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
@@
 For ``HPC random'' approach data chunks have to be copied to the local scratch disks first, and the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS and hence do not need data staging.
-Hadoop however needs to preprocess reads before the actual alignment stage, in order  to be able to operate in MR manner, also resulting in what we term ''communication costs''.
+Hadoop however needs to preprocess reads before the actual alignment stage, in order  to be able to operate in MR manner, also resulting in what we term ``communication costs''.
 Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
@@
 \end{center}
-\label{table:4}
+\label{table:6}
 \end{table}%
@@
 We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}(a)).
-For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow, Figure~.
+For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow.
 The most of data managing job is done automatically, and the results can be downloaded to the client machine.
-Modular structure allows one easy to modify the source code to integrate to the existing computing center’s architecture, Figure~\ref{fig:fig4}(a).
+Modular structure allows one easy to modify the source code to integrate to the existing computing center’s architecture, Figure~\ref{fig:fig4}(b).
 For example the UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will stay locally, reducing amount of unnecessary risks.
@@
 
-%To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow software were suggested.
-A state-of-art bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
+To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow/Myrna  were suggested.
+The calculation time reduction can be observed in the Table~\ref{table:preprocess}.
+%A state-of-art Bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
 
@@
 
-Exploiting the embarrassingly parallel nature of the short reads mapping we used a highly optimized bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes (Figure~\ref{fig:fig3}).
-Our results show that Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
+Exploiting the embarrassingly parallel nature of the short reads mapping we used a highly optimized Bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes (Figure~\ref{fig:fig3}).
+Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
 In addition we show how performance of the HPC approach can be improved redefining the queueing system's default behaviour.
@@
 The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-We thank Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage usage.
+This work was supported by the Swedish strategic research programme eSSENCE~\url{http://essenceofescience.se/home/}.
+We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage.
 We also thank Jonas Hagberg (BILS, Stockholm, Sweden), for implementing the Cloudgene extensions to import data from UPPMAX filesystem.
@@
 \begin{enumerate}
-\item  HPC1:
-The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
+%\item  HPC1:
+%The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
 The data and reference genomes were read and written to a parallel shared storage system.
@@
 \item 
-HPC2:
-Multinode short read mapping was performed on the Milou cluster\cite{milouCluster} , equipped with dual 8cores CPUs Intel Xeon E5-2660, ( 2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache.), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
+HPC:
+Multinode short read mapping was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8cores CPUs Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
 
@@
 
-\item Our main Hadoop test platform was deployed on the private cloud at UPPMAX, using the OpenNebula~\cite{opennebula} cloud management system.
-Each node in this deployment was equipped with dual quad-core CPUs (Intel Xeon 5420; clock frequency of 2.50GHz\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
+\item Our  Hadoop test platform was deployed on the private cloud at UPPMAX, using the OpenNebula~\cite{opennebula} cloud management system.
+Each node in this deployment was equipped with dual 4-core CPUs (Intel Xeon 5420; clock frequency of 2.50GHz\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
 The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
+One should notice that the used physical hardware provided with less RAM, as desired.
+Each VM node has 7 cores, each of each can use less than 2 GB of RAM, that is a little for Hadoop.
+We would expect much better performance would have twice as much memory, as recommended.
 

### 2014-02-19 text1.tex
proofread, with some open questions

@@
 								% TeX will automatically convert eps --> pdf in pdflatex		
+\usepackage[normalem]{ulem}
 \usepackage{amssymb}
@@
 
+\newcommand{\COMMENT}[1]{{\color{red} #1 }}
+
 \title{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)}
-\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth}
+\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth \COMMENT{[specify affiliation]}} 
 %\date{}							% Activate to display a given date or no date
@@
 With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
-Hadoop, emerged in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-We here report a quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases.
+Hadoop, emerging in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
+Herein we report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
 In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts.
-Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution.
-As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in future biological data analysis.
+From the observered scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100\,gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution.
+As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in the future of biological data analysis.
+\COMMENT{Brit spelling assumed since that's most common in this document}
 }
@@
 As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 days~\cite{illumina}.
-Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
+Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} \COMMENT{[unusual to use both footnotes and endnotes]} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) \COMMENT{huh?} storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
 
@@
 A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Most of the bioinformatic programs are written for the Linux operating system.
-Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.
+\sout{Most of the bioinformatic programs are written for the Linux operating system.} /COMMENT{awkward but maybe it's needed} Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made with distributed computing in mind.
+Many others do not even have the native ability to use multiple cores \sout{on the same computer}\footnote{Samtools acquired this feature in v.0.1.19}.
 
@@
 
-The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.
-For the tools which do not support the OMP natively, i.e.
+The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16.
+For the tools which do not support the OMP natively, e.g.
 Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
-Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating to split  the dataset and  use multi-node parallelisation.
+Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating \COMMENT{huh?} to split the dataset and use multi-node parallelisation.
 
-Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
+Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronisation (or load balancing) has to be woven into the software code \sout{by a programmer} and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
-Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting e.g.
-involve already existing utilities and cluster tools, split the data into chunks and process them on the separate nodes, and merging the results afterwards.
-This kind of solution benefits from both MPI-like  and OMP and provides good performance, but the development requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.
+Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting.
+This involves using existing utilities and cluster tools to split the data into chunks, process them on the separate nodes, and merge the results afterwards.
+This kind of solution benefits from both MPI-like and OMP parallelisation and provides good performance, but the development requires substantial expertise in order to be efficient.
+Since the process is tightly coupled to the local computational cluster and network architectures, it might not be possible to be re-used in other settings.
 
@@
 %Properties of Hadoop and HDFS- data localization and programming model etc
-The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running task in a {\it massively} parallel way.
+The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running tasks in a {\it massively} parallel way.
 This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
-The prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. 
+The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. \COMMENT{Fix bib error (can't have both editor and author for this ref type)} 
 The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
@@
 %Introduce what we did in the manuscript.
-In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
+In this manuscript we focus on the question {\it if} and {\it when} \COMMENT{that's two questions} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
 Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
-We use five datasets with short reads of different sizes, align them against the reference genome, and call the variances.
+We use five datasets with short reads of different sizes, align them against the appropriate reference genome, and call the variances.
 For the existing Hadoop software we propose modifications, in order to benefit fully from massively parallel nature of MR computations. 
-For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
-The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.
+For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short read alignment and exploiting the network fully, thereby speeding up calculations.
+The execution times for each dataset were collected and scaling relations were analysed to answer the question\COMMENT{s} in focus.
 
-The manuscript is structured as follows: In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
+The manuscript is structured as follows: \COMMENT{[odd to do this in a normal publication, but OK]} In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
 Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources.
@@
 \subsection{Datasets}
-We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
-Data for datasets I-III and V were obtained using Illumina/(HiSeq) sequencing platforms.
+We used publicly available DNA-seq datasets~(I--III), a synthetic dataset~(IV) of {\it A.thaliana}, the well-known model plant, and dataset (V) of two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
+Data for datasets I-III and V were obtained using Illumina/HiSeq sequencing platforms.
 Further information about the datasets is provided in the Supplementary material section.
@@
 \subsection{Analysis pipelines}
-We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and one on regular HPC with a batch processing system (hereafter referred to as HPC).
+We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as HPC).
 Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage.
-All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
-Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e.
-GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:
+All experiments were repeated several times and the results averaged \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
+Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (e.g., GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass \COMMENT{huh?} the same stages on Hadoop and HPC:
 
@@
 In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools.
-The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core.
+The Bowtie aligner natively implements~OMP, meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core.
 Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
 Where available these features were used to improve the analysis speed.
-For the exact workflow used one can refer to the code repository created for this work~\cite{code_repo}.
+The exact workflow used is available in the code repository created for this work~\cite{code_repo}.
+\COMMENT{missing ref}
 
@@
 Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
-Unfortunately Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way.
+Unfortunately, Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way.
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
@@
 \subsection{Computational resources}
-To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 8 dual-core CPUs.
-Data and reference genomes were stored on a parallel shared storage system\cite{gulo}.
+To run the HPC analysis pipeline we used a computational cluster at UPPMAX on nodes equipped with 8 dual-core CPUs.
+Data and reference genomes were stored on a parallel, shared-storage system\cite{gulo}.
 The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager.
@@
 \subsection{Modified preprocessing stage}
-A native preprocessing stage in the Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster.
+A native preprocessing stage in Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster.
 Data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}.
-The only way to introduce parallelisations in the native preprocessing stage~-- is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
+The only way to introduce parallelisations in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
 This, however, is lacking the massively parallel way of treating the data.
 
-Being  pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both~--  storage and  computers,  
-therefore considered the data already been downloaded to the storage Hadoop can access.
+Being pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both storage and computers, therefore we considered that the data already been downloaded to storage Hadoop could access.
 We rewrote the preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
-The demand is that the FASTQ data are BZIP2 archived, and  accessible by Hadoop.
-For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested to the HDFS.
-The BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}.
-Also the BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e.
-for the developer there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
+The requirement is that the FASTQ data are BZIP2 archived and accessible by Hadoop.
+For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested \COMMENT{special word?} to the HDFS.
+BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}.
+Also, BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e., \sout{for the developer} there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
 
 The Hadoop streaming library offers a possibility to write Mapper and Reducer for Hadoop jobs in any programming language.
-Our Python scripts efficiently process  short reads, produced by  Illumina sequencing platform of different versions, as well as the FASTQ files converted from the SRA format, NCBI Sequence Read Archive.
+Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive).\COMMENT{ref?}
 The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
-To ensure that the both forward and reverse read of the same pair will end up on the same Reducer, the secondary Mapper key-sorting mechanisms were involved.
+To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.
 
-In order to compare the native preprocessor and the proposed one, the data was put on the HDFS in beforehand, to eliminate possible network delays i.e. while downloading data from some Amazon cloud.
+In order to compare the native preprocessor and the proposed one, the data was put on the HDFS beforehand, to eliminate possible network delays, e.g., while downloading data from some Amazon cloud.
 Table~\ref{table:preprocess} illustrates the benefits of our approach.
@@
 \small
-\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on~$p=56$ cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node, $p=16$ cores.}
+\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on \sout{$p=56$} 56 cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node \sout{, $p=16$} with 16 cores.
+\COMMENT{A bit odd to use the vertical lines before/after table, but OK.}}
 \begin{center}
 \begin{tabular}{|l|c|c|c|c|}
-dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
+Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
 \hline
-Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1$	\\
+Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1.0$	\\
 Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
@@
 The script\footnote{\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}} ran on a single HPC node, utilizes multiple cores, with the limiting factor being the HDD IO performance, which effectively leaves with 5-7 cores of 16.
+\COMMENT{awkward}
 
 \subsection{Accuracy of pipelines}
-Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, but still we expect them to capture and correctly identify the mutation.
-We tested the correctness just for the smallest dataset~I.
+Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp, correspondingly) we should not expect them to deliver perfectly matching SNP lists, but still we expect them to capture and correctly identify the mutations.
+We tested the correctness just for the smallest dataset (I).
 The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.
@@
 \subsection{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}).
-As one can see the aligning process with the Bowtie scales fairly good, but the SNP calling part is a bottleneck,  scaling worse than the Bowtie, and consuming progressively larger part of the total calculation time
+To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
+\sout{As one can see} the aligning process with the Bowtie scales fairly well, but the SNP calling part is a bottleneck, scaling worse than the Bowtie, and consuming a progressively larger portion of the total calculation time.
 %\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
@@
 
-One of the attractive sides of using Hadoop~MR is its almost linear scalability,~i.e.
-calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie etc.).
-Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster.
+One of the attractive sides of using Hadoop~MR is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie, etc.).
+Figure~\ref{fig:fig1} \COMMENT{this figure should be closer, table and figure layout odd in general}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ \COMMENT{is this 'p=' notation necessary} cores Hadoop cluster.
 Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
-To stress the linear nature of scaling, both sets of points were fit to linear polynomial using the least squares method.
+\COMMENT{so what}
+To stress the linear nature of scaling, both sets of points were fit \sout{to linear polynomial} using the least squares method.
 
@@
 \small
-\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC cluster. }
+\caption{Calculation time for Dataset I executed on different number of cores on a node of HPC cluster. \COMMENT{awkward title, consider replacing with graph (and new title)}}
 \begin{center}
@@
 \hline
-	& mapping 	&	SNP calling\tablefootnote{We  used here special tricks how to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781/
-}}	&	total  &total & map& SNP call\\
+	& mapping 	&	SNP calling\tablefootnote{We used special tricks to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781}.}	&	total  &total & map& SNP call\\
 \hline
@@
 Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.
-Big deviation for the HPC is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.}
+Big deviation \COMMENT{what aspect of deviation, not fixable?} for the HPC is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.
+\COMMENT{No p= used here.} }
 \begin{center}
@@
 	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time depending on the  size for selected datasets for~$p=56$ cores Hadoop cluster on the private Cloud.
-The linear fit based on least squares method are provided, revealing  linear scaling.
-Datasets are marked with roman numerals.}
+	\caption{Calculation time depending on the size of selected datasets for~$p=56$ \COMMENT{p= returns} cores Hadoop cluster on the private Cloud.
+\sout{The least squares fit reveals linear scaling.} Datasets are marked with roman numerals.
+\COMMENT{Rather large graph for such a simple result.} }
 	\label{fig:fig1}
@@
 \subsection{Comparing Hadoop and HPC runtime efficiency for different dataset sizes}
-Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then that the larger the dataset is, the more suitable Hadoop becomes compare to the HPC approach.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
+\COMMENT{delete 'for different dataset sizes'} Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
+One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
+In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the following function:
 $$F=T_{p}\times p,$$ 
-where~$T_{p}$ is the calculation time on~$p$ cores.
-Using the data from Table~\ref{table:4},  we plot the ratio $F_{Hadoop}/F_{HPC}$, keeping in mind that the closer the ratio to the unity, the closer Hadoop's efficiency is to that of the HPC approach.
+where~$T_{p}$ is the calculation time on~$p$ cores. \COMMENT{This trivial/obvious formula doesn't deserve such special treatment.}
+Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ \sout{, keeping in mind that the closer the ratio is to the unity, the closer Hadoop's efficiency is to that of the HPC approach}.
+
 \begin{figure}
@@
 	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases. Calculations were carried out for~$p=56$ and $p=16$ cores for Hadoop and HPC correspondingly.
-The points are fit in a curve using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.
-Datasets are marked with roman numerals.}
+The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are marked with roman numerals.}
 	\label{fig:fig2}
 \end{figure}
-The curve  in~Figure~\ref{fig:fig2}, based on Table~\ref{table:4}, is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=16$), and displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
-The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical infinite dataset.
-As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data is larger.
-The parabolic extrapolation for  the ratio reads~$1.70\pm0.01$, meaning that Hadoop running even  in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with  the HPC approach which is being run on ``bare metal''  on the modern hardware for the datasets greater than $100$\,Gbases (dataset~IV), which is a usual size for human genome sequencing giving the average sequencing depth of  about~$30x$.
+
+The curve  in~Figure~\ref{fig:fig2}, based on Table~\ref{table:4}\sout{, is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=16$), and} displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
+Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
+As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
+The parabolic extrapolation for the ratio is $1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
 The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
-Taking into account  a {\it linear} scaling for the Hadoop approach for those datasets, Figure~\ref{fig:fig1}, one can conclude that HPC approach scales worse than linearly\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
+Considering a {\it linear} scaling for the Hadoop approach for those datasets, Figure~\ref{fig:fig1}, one can conclude that HPC approach scales worse than linearly\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
 being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
-We suspect the main reason for such a behavior is the fact that for large datasets (starting from dataset~III in our case) the SNP calling routine with the Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
-If lacking  RAM, the Samtools spilling to HDD, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and  time expensive IO on a high rate.
-The Table~\ref{table:5} shows the role of the amount of  allocated RAM on the Samtools sort timing, as a most time expensive,  for the dataset~IV, starting from 8GB up to 80GB, performing on a single HPC node.
+\COMMENT{Why jump back and forth in the discussion of figures 1 and 2?} We suspect the main reason for such a behavior is the fact that for large datasets (starting from dataset~III in our case) the SNP calling routine with the Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
+If lacking  RAM, the Samtools swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
+Table~\ref{table:5} shows the role of the amount of allocated RAM on the Samtools sort timing, as a most time expensive \COMMENT{huh?}, for the dataset~IV, starting from 8GB up to 80GB, performing on a single HPC node.
+The large uncertainty for small amounts of RAM we believe is due to other users' load on the cluster network.
+As a result, a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling on much more economic hardware in linear time with just 12GB RAM per VM node.
+
 \begin{table}[htdp]
@@
 Averaging is done over 5 independent simulations.
-The RAM amount is given for the all  16 cores.
+The amount of RAM is given for all 16 cores.
 }
@@
 \end{table}%
-A large uncertainty for  small amount of RAM we believe is due to other users' load  on the cluster network.
-As the  result a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling  on much economic hardware on linear time  with just 12GB RAM per VM node.
 
  \subsection{Comparing the network communication efficiency for the Hadoop and HPC approaches }
-Network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
+The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
 The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
-We provide a comparison on how HPC and Hadoop network communication costs depend on the number of nodes involved, for a fixed dataset size (dataset~IV).
+We provide a comparison of HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gb, dataset~IV).
 
-Due to the trivially parallelisation nature of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, i.e.
-split the initial data into chunks to process them independently.
+Due to the trivially parallelisability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
 Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
 
-There are several program packages for short read alignment designed with MPI support~\cite{pmap, gnumap}.
+There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}.
 Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}.
-However,  i.e.
-Pmap package functions poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising memory exceptions.
-We implemented a highly optimized Bash script making use of standard Unix utilities to use the  HPC cluster network as efficiently as possible~\cite{repo},  and compared the network performance with the standard Hadoop HDFS approach.
-We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
-Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases.
-For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS.
+\COMMENT{why footnote rather than endnote} However, e.g., the Pmap package functions poorly \sout{on UPPMAX cluster} for datasets larger than 20 Gbases \sout{, raising memory exceptions}.
+\COMMENT{Needs an intro like To circumvent this limitation} We implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, \COMMENT{missing reference, written as 'repo' but assumed to be same as 'code\_repo' above} and compared the network performance with the standard Hadoop HDFS approach.
+We separated the mapping time, $T_{mapping}$, and the communication time, $T_{comm}$, and plotted $T_{mapping}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
+This measure is applicable to both HPC and Hadoop, however, $T_{comm}$ has a different explanation.
+For Hadoop, the short reads in FASTQ format have to be preprocessed (involving node communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data ingestion into the HDFS.
 We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, the~$T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}} files back to the delivery location over the network.
+For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}} files back to the delivery location over the network.
 
 Figure~\ref{fig:fig3} shows the~$T_{mapping}/T_{comm}$ ratio as a function of the reciprocal number of nodes~$1/N$ for Hadoop and HPC approaches.
-The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
+\COMMENT{why describe fig 3 again} The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
 
-\textit{Hadoop results (filled circles)}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that the Bowtie provides almost  linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}.
-Since the ratio~$T_{mapping}/T_{comm}\approx \mbox{const}$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication is in the preprocessing stage.
+\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
+It is known that Bowtie provides almost linear scaling between mapping time and dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}.
+Since the ratio~$T_{mapping}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
 
-\textit{HPC results}: The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in parallel, which are simultaneously being pushed to the local scratches of the nodes, allocated by SLURM  (open circles).
-One can see two distinct linear stretches, each with a different slope.
+\textit{HPC results}: The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}  (open circles in Figure~\ref{fig:fig3}) is as follows: the data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratches of the nodes allocated by SLURM.
+One can see two distinct linear stretches.
 One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
-The former (horizontal stretch) is explained as for Hadoop~-- the more nodes being involved the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e.
-a linear dependency, which one can observe in the plot.
-The transition area between the two modes~-- saturated and unsaturated~-- has the next origin: each hard disk drive on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
+The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
+The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e., a linear dependency, which one can observe in the plot.
+The transition area between the two modes~-- saturated and unsaturated~-- has the next origin \COMMENT{huh?}: each hard disk drive on the local node can write the data at a speed of about \sout{100MB/sec\,$\approx$\,} 1Gbit/sec.
+Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
 The nodes are being allocated on the same rack, which is the default SLURM behaviour.
 
-The scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
+Scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
 Allocating the nodes on random racks allows one to engage more nodes without network saturation.
@@
 The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
-The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approach suffer from the network saturation at higher number of nodes.
+The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
 
@@
 %linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
-Linear fit  done with the least-squares method.}
+Linear fit  done with the least-squares method. \COMMENT{Figure legend too complex. Y-axis label inconsistant with previous figure.}}
 	\label{fig:fig3}
@@
 \begin{table}[htdp]
-\caption{Timings  for mapping, and the ratio~$T_{mapping}/T_{comm}$  for HPC and  Hadoop clusters for Dataset IV.
-For ``HPC random'' approach data chunks have to be copied to the local scratch disks first, and the alignments (SAM files) copied back, while Hadoop keeps all the data inside HDFS and hence do not need data staging.
-Hadoop however needs to preprocess reads before the actual alignment stage, in order  to be able to operate in MR manner, also resulting in what we term ``communication costs''.
+\caption{Timings  for mapping and the ratio~$T_{mapping}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
+For the ``HPC random'' approach, data chunks have to be copied to the local scratch disks first and the alignments (SAM files) copied back while Hadoop keeps all the data inside HDFS and hence does not need data staging.
+Hadoop however needs to preprocess reads before the actual alignment stage in order to be able to operate in MR manner resulting in what we term ``communication costs''.
 Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
@@
 
-Number nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$	&Number  nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$\\
+Number of nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$	&Number of nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$\\
 (cores)					&minutes		&							&(cores)			&minutes&\\
@@
 \end{table}%
- At the same time, the HDFS keeps data locality, aiming to reduce the amount of communications, resulting less data move and, therefore, better scalability.
-Our Hadoop-in-the-Cloud cluster has no more~($\approx 40$) free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the exposed behaviour is a generic for Hadoop with HDFS.
+ At the same time, the HDFS keeps data locality \COMMENT{'keeps data locally' or 'maintains data locality'}, aiming to reduce the amount of communications, resulting less data move and, therefore, better scalability.
+Our Hadoop-in-the-Cloud cluster has no more~($\approx 40$) \COMMENT{how does 40 approximate zero?} free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
+\COMMENT{need ref?}
 
@@
 
-A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying an experience for the end user.
-One of the alternatives for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private cloud.
+A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying the experience for the end user.
+One of the alternatives for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private clouds.
 We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}(a)).
-For the particular task in DNA-seq experiment, Cloudgene provides the intuitive interface making one easy to follow.
-The most of data managing job is done automatically, and the results can be downloaded to the client machine.
-Modular structure allows one easy to modify the source code to integrate to the existing computing center’s architecture, Figure~\ref{fig:fig4}(b).
-For example the UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will stay locally, reducing amount of unnecessary risks.
+For our particular task in DNA sequencing, Cloudgene provides the intuitive interface making one easy to follow.
+Most of the data managing work is done automatically and the results can be downloaded to the client machine.
+The modular structure allows modification of the source code to adapt to the existing computing center’s architecture, Figure~\ref{fig:fig4}(b).
+For example, UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will be held locally, reducing the amount of unnecessary risks.
 
@@
 \textbf{a)} Pipeline selection - in our case containing the Crossbow pipeline.
-\textbf{b)} The UPPMAX-adapted functionality to browsing and import data from the users home folder in the shared file system.}
+\textbf{b)} The UPPMAX-adapted functionality to browsing and import data from the user's home folder in the shared file system.}
 	\label{fig:fig4}
@@
 A key result is the data size where Hadoop is favorable to regular HPC batch systems, and we developed highly optimized pipelines for both scenarios.
-We found that dataset sizes larger than 100\,Gbases is where Hadoop excel over HPC, and also that Hadoop shows almost linear scalability (Figure~\ref{fig:fig1}).
+We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability (Figure~\ref{fig:fig1}).
 Extrapolation to infinite dataset size (Figure~\ref{fig:fig2}) reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
@@
 The calculation time reduction can be observed in the Table~\ref{table:preprocess}.
+\COMMENT{So, is Hadoop better after the modifications for infinite datasets?}
 %A state-of-art Bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
@@
 Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show how performance of the HPC approach can be improved redefining the queueing system's default behaviour.
-Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene), provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
+In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behaviour.
+\COMMENT{Should maybe mention somewhere that this sort of cheating.} Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene), provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
 
@@
 
-\subsection{Used datasets}
+\subsection{Datasets}
 
@@
 HPC:
-Multinode short read mapping was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8cores CPUs Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
+Multinode short read mapping was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
 
@@
 
-\item Our  Hadoop test platform was deployed on the private cloud at UPPMAX, using the OpenNebula~\cite{opennebula} cloud management system.
-Each node in this deployment was equipped with dual 4-core CPUs (Intel Xeon 5420; clock frequency of 2.50GHz\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
+\item Our Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud management system.
+Each node in this deployment was equipped with dual 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
 The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-One should notice that the used physical hardware provided with less RAM, as desired.
-Each VM node has 7 cores, each of each can use less than 2 GB of RAM, that is a little for Hadoop.
-We would expect much better performance would have twice as much memory, as recommended.
+Note that the physical hardware provided less RAM than desired.
+Each VM node has 7 cores, each of each can use less than 2 GB of RAM, which is little for Hadoop.
+We would expect much better performance with twice as much memory, as recommended.
+
+\COMMENT{Journal abbreviations in references are inconsistant.
+The journal name for ref 2 is wrong.
+Capitalization of refs 4, 14, 16, 18, 19, 25, 27, etc are wrong.
+All refs should be checked for accuracy and style consistancy.
+Many or all footnotes should be changed to endnotes (references).}
 

### 2014-02-20 text1.tex
Adapted for part of Wes comments.

@@
 As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in the future of biological data analysis.
-\COMMENT{Brit spelling assumed since that's most common in this document}
 }
@@
 As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 days~\cite{illumina}.
-Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} \COMMENT{[unusual to use both footnotes and endnotes]} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) \COMMENT{huh?} storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
+%Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} \COMMENT{[unusual to use both footnotes and endnotes]} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) \COMMENT{huh?} storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
 
@@
 A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-\sout{Most of the bioinformatic programs are written for the Linux operating system.} /COMMENT{awkward but maybe it's needed} Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made with distributed computing in mind.
-Many others do not even have the native ability to use multiple cores \sout{on the same computer}\footnote{Samtools acquired this feature in v.0.1.19}.
+% \sout{Most of the bioinformatic programs are written for the Linux operating system.} /COMMENT{awkward but maybe it's needed} 
+Many of the most widespread tools for this purpuse, like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools}, are not made with distributed computing in mind.
+Many others do not even have the native ability to use multiple cores.
+%\sout{on the same computer}\footnote{Samtools acquired this feature in v.0.1.19}.
+%\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications}
 
@@
 
-The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16.
-For the tools which do not support the OMP natively, e.g.
-Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
-Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating \COMMENT{huh?} to split the dataset and use multi-node parallelisation.
+The most common approach to speed up NGS tools is to parallelise within a compute node (multi-core parallelisation) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16.
+\COMMENT{confirm this!} For the tools which do not support OMP natively, e.g.
+Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel Linux utility~\cite{gnuparallel}.
+Of great importance is that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and multi-node parallelisation generally are not bound by these constraints.
 
-Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronisation (or load balancing) has to be woven into the software code \sout{by a programmer} and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
+Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronisation (or load balancing) has to be woven into the software code and there are few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
@@
 This kind of solution benefits from both MPI-like and OMP parallelisation and provides good performance, but the development requires substantial expertise in order to be efficient.
-Since the process is tightly coupled to the local computational cluster and network architectures, it might not be possible to be re-used in other settings.
+Since the process is tightly coupled to the local computational cluster and network architectures, it might also not be possible to be re-used in other settings.
 
@@
 %Introduce what we did in the manuscript.
-In this manuscript we focus on the question {\it if} and {\it when} \COMMENT{that's two questions} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
+In this manuscript we focus on the question {\it when} \COMMENT{that's two questions} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
 Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
@@
 For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short read alignment and exploiting the network fully, thereby speeding up calculations.
-The execution times for each dataset were collected and scaling relations were analysed to answer the question\COMMENT{s} in focus.
+The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.
 
-The manuscript is structured as follows: \COMMENT{[odd to do this in a normal publication, but OK]} In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
+%The manuscript is structured as follows: \COMMENT{[odd to do this in a normal publication, but OK]} In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
 Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources.
@@
 \subsection{Datasets}
-We used publicly available DNA-seq datasets~(I--III), a synthetic dataset~(IV) of {\it A.thaliana}, the well-known model plant, and dataset (V) of two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
+We used publicly available DNA-seq datasets~(I--III) as well as a synthetic dataset~(IV) of {\it A.thaliana}, the well-known model plant, and a dataset (V) of two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
 Data for datasets I-III and V were obtained using Illumina/HiSeq sequencing platforms.
@@
 \footnotesize
-\caption{Datasets used}
+\caption{Datasets used in the comparison. }
 \begin{center}
@@
 All experiments were repeated several times and the results averaged \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
-Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (e.g., GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass \COMMENT{huh?} the same stages on Hadoop and HPC:
+Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (e.g., GATK~\cite{gatk}), we decided to create the analysis pipeline as simple as possible to be able to make a better comparison with the one implemented for Hadoop.
+
+%pass \COMMENT{huh?} the same stages on Hadoop and HPC:
 
@@
 \subitem Short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: Samtool~ver.~0.1.19
+\subitem SNP calling: Samtools~ver.~0.1.19
 \item Hadoop approach: Crossbow~\cite{crossbow}
@@
 Where available these features were used to improve the analysis speed.
-The exact workflow used is available in the code repository created for this work~\cite{code_repo}.
+The exact workflow used is available in \COMMENT{github?} the code repository created for this work~\cite{code_repo}.
 \COMMENT{missing ref}
@@
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation.
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available in \COMMENT{github?} the code repository~\cite{code_repo}).
 
 \subsection{Computational resources}
-To run the HPC analysis pipeline we used a computational cluster at UPPMAX on nodes equipped with 8 dual-core CPUs.
-Data and reference genomes were stored on a parallel, shared-storage system\cite{gulo}.
+To run the HPC analysis pipeline we used a computational cluster at UPPMAX, which is heavily used for analysis of NGS-data~\cite{lampa}, on nodes equipped with 8 dual-core CPUs.
+Data and reference genomes were stored on a parallel, shared-storage system.
 The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager.
@@
 For details on computational resources, see Supplementary material.
+%\cite{gulo}
 
@@
 \subsection{Modified preprocessing stage}
-A native preprocessing stage in Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster.
-Data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}.
+A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}.
 The only way to introduce parallelisations in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
@@
 
-Being pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both storage and computers, therefore we considered that the data already been downloaded to storage Hadoop could access.
-We rewrote the preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
+%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
+
+Assuming that sequencing platforms are usually in close proximity with facilities that provide both storage and computers, we considered that the data already had been downloaded to storage that Hadoop could access.
+We rewrote the Crossbow preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
 The requirement is that the FASTQ data are BZIP2 archived and accessible by Hadoop.
-For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested \COMMENT{special word?} to the HDFS.
-BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}.
-Also, BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e., \sout{for the developer} there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
+For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS storage.
+BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner.
+Also, BZIP2 format is natively supported by Hadoop, i.e., there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
 
-The Hadoop streaming library offers a possibility to write Mapper and Reducer for Hadoop jobs in any programming language.
+%\footnote{other options like splittable LZO are also possible}
+
+The Hadoop streaming library offers a possibility to write Mapper and Reducer \COMMENT{Introduce in introduction} for Hadoop jobs in any programming language.
 Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive).\COMMENT{ref?}
@@
 
-In order to compare the native preprocessor and the proposed one, the data was put on the HDFS beforehand, to eliminate possible network delays, e.g., while downloading data from some Amazon cloud.
+In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays, e.g., while downloading data from Amazon.
 Table~\ref{table:preprocess} illustrates the benefits of our approach.
-\begin{table}[htdp]
+To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
+The script (see Supplementary information) ran on a single HPC node, utilizes multiple cores, with the limiting factor being the hard disk IO performance.%, which effectively leaves with 5-7 cores of 16.
+\COMMENT{awkward}
+
+\begin{table}[tdp]
 \small
-\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on \sout{$p=56$} 56 cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node \sout{, $p=16$} with 16 cores.
-\COMMENT{A bit odd to use the vertical lines before/after table, but OK.}}
+\caption{Timings (in minutes) with Crossbow's native read preprocessor and with derived approach for  datasets II, III, IV  on a 56 cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node with 16 cores.}
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|}
+\begin{tabular}{l|c|c|c}
 Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
@@
 \end{table}%
-To complete the comparison the same preprocessing was performed with the BASH scripting against the data located on the HPC storage.
-The script\footnote{\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}} ran on a single HPC node, utilizes multiple cores, with the limiting factor being the HDD IO performance, which effectively leaves with 5-7 cores of 16.
-\COMMENT{awkward}
 
@@
 To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
-\sout{As one can see} the aligning process with the Bowtie scales fairly well, but the SNP calling part is a bottleneck, scaling worse than the Bowtie, and consuming a progressively larger portion of the total calculation time.
+The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome\footnote{\url{http://www.biostars.org/p/48781}}.
 %\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
@@
 One of the attractive sides of using Hadoop~MR is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie, etc.).
-Figure~\ref{fig:fig1} \COMMENT{this figure should be closer, table and figure layout odd in general}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ \COMMENT{is this 'p=' notation necessary} cores Hadoop cluster.
-Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
-\COMMENT{so what}
-To stress the linear nature of scaling, both sets of points were fit \sout{to linear polynomial} using the least squares method.
+Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
+To stress the linear nature of scaling, both sets of points were fit using the least squares method.
 
-\begin{table}[htb]
+\begin{table}[tb]
 \small
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|ccc|}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c|}{speed-ups} \\
+\begin{tabular}{c|c|c|c|ccc}
+$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c}{speed-ups} \\
 \hline
-	& mapping 	&	SNP calling\tablefootnote{We used special tricks to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781}.}	&	total  &total & map& SNP call\\
+	& mapping 	&	SNP calling	&	total  &total & map& SNP call\\
 \hline
@@
 
-\begin{table}[htdp]
+\begin{table}[tdp]
 \small
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\begin{tabular}{l|c|c|c|c|c|c|c}
 
@@
 
-\begin{figure}
+\begin{figure}[tb]
 	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time depending on the size of selected datasets for~$p=56$ \COMMENT{p= returns} cores Hadoop cluster on the private Cloud.
-\sout{The least squares fit reveals linear scaling.} Datasets are marked with roman numerals.
-\COMMENT{Rather large graph for such a simple result.} }
+	\caption{Calculation time for selected datasets (labeled with roman numerals) for the 56 cores Hadoop cluster shows a linear scaling with dataset size. }
 	\label{fig:fig1}
@@
 
-\subsection{Comparing Hadoop and HPC runtime efficiency for different dataset sizes}
-\COMMENT{delete 'for different dataset sizes'} Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
+\subsection{Comparing Hadoop and HPC runtime efficiency}
+Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
 One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
@@
 where~$T_{p}$ is the calculation time on~$p$ cores. \COMMENT{This trivial/obvious formula doesn't deserve such special treatment.}
-Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ \sout{, keeping in mind that the closer the ratio is to the unity, the closer Hadoop's efficiency is to that of the HPC approach}.
+
+Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
+Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
+As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
+The parabolic extrapolation for the ratio is $1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
@@
 \end{figure}
-
-The curve  in~Figure~\ref{fig:fig2}, based on Table~\ref{table:4}\sout{, is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=16$), and} displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
-Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
-As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for the ratio is $1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
@@
 We would expect much better performance with twice as much memory, as recommended.
+\end{enumerate}
+
+\subsection{Bash script for the preprocessing stage}
+
+\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
 
@@
 
-\end{enumerate}
-
 \bibliography{text1}

### 2014-02-23 text1.tex
More work on wes comments.

@@
 To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
-The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome\footnote{\url{http://www.biostars.org/p/48781}}.
+The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome.
+%\footnote{\url{http://www.biostars.org/p/48781}
 %\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
@@
 Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
-To stress the linear nature of scaling, both sets of points were fit using the least squares method.
+%To stress the linear nature of scaling, both sets of points were fit using the least squares method.
 
@@
 
-\begin{figure}[tb]
+\begin{figure}[bt]
 	\input{./figures/fig1/fig1.tex}
@@
 	\label{fig:fig1}
+\end{figure}
+
+\begin{figure}[bt]
+	\input{./figures/fig2/fig2.tex}
+	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases. Calculations were carried out for 56 cores Hadoop and 16 cores HPC cluster correspondingly.
+The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are labeled with roman numerals.}
+	\label{fig:fig2}
 \end{figure}
@@
 
-Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
+Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
 Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
@@
 
-\begin{figure}
-	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases. Calculations were carried out for~$p=56$ and $p=16$ cores for Hadoop and HPC correspondingly.
-The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are marked with roman numerals.}
-	\label{fig:fig2}
-\end{figure}
+The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
+Considering the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
+%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
+being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
+\COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
+We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
+If lacking RAM, Samtools implementation swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
+Table~\ref{table:5} shows the sorting runtime performance of Samtools for the dataset~IV run with different amount of RAM available on a single HPC node.
+The large uncertainty for small amounts of RAM we believe is due to other users' load on the cluster network.
+As a result, a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling on much more economic hardware in linear time with just 12GB RAM per virtual node.
 
-The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
-Considering a {\it linear} scaling for the Hadoop approach for those datasets, Figure~\ref{fig:fig1}, one can conclude that HPC approach scales worse than linearly\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
-being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
-\COMMENT{Why jump back and forth in the discussion of figures 1 and 2?} We suspect the main reason for such a behavior is the fact that for large datasets (starting from dataset~III in our case) the SNP calling routine with the Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
-If lacking  RAM, the Samtools swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
-Table~\ref{table:5} shows the role of the amount of allocated RAM on the Samtools sort timing, as a most time expensive \COMMENT{huh?}, for the dataset~IV, starting from 8GB up to 80GB, performing on a single HPC node.
-The large uncertainty for small amounts of RAM we believe is due to other users' load on the cluster network.
-As a result, a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling on much more economic hardware in linear time with just 12GB RAM per VM node.
-
-\begin{table}[htdp]
+ \begin{table}[hbtp]
 \small
@@
 The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
-We provide a comparison of HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gb, dataset~IV).
+To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gb, dataset~IV).
 
@@
 
-There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}.
-Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}.
-\COMMENT{why footnote rather than endnote} However, e.g., the Pmap package functions poorly \sout{on UPPMAX cluster} for datasets larger than 20 Gbases \sout{, raising memory exceptions}.
-\COMMENT{Needs an intro like To circumvent this limitation} We implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, \COMMENT{missing reference, written as 'repo' but assumed to be same as 'code\_repo' above} and compared the network performance with the standard Hadoop HDFS approach.
+There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}, and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
+However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
+To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, and compared the network performance with the standard Hadoop HDFS approach.
 We separated the mapping time, $T_{mapping}$, and the communication time, $T_{comm}$, and plotted $T_{mapping}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
@@
 We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}} files back to the delivery location over the network.
+For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.
+%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
 
-Figure~\ref{fig:fig3} shows the~$T_{mapping}/T_{comm}$ ratio as a function of the reciprocal number of nodes~$1/N$ for Hadoop and HPC approaches.
-\COMMENT{why describe fig 3 again} The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
+%The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
 
@@
 
-\textit{HPC results}: The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}  (open circles in Figure~\ref{fig:fig3}) is as follows: the data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratches of the nodes allocated by SLURM.
+%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
+\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM.
 One can see two distinct linear stretches.
@@
 The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e., a linear dependency, which one can observe in the plot.
-The transition area between the two modes~-- saturated and unsaturated~-- has the next origin \COMMENT{huh?}: each hard disk drive on the local node can write the data at a speed of about \sout{100MB/sec\,$\approx$\,} 1Gbit/sec.
+The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
+%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
+and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
+The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
 Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
@@
 For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}  for the data.
+$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}.
 The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
- \multicolumn{3}{|c|}{Hadoop} & \multicolumn{3}{c|}{ HPC random} \\
+\begin{tabular}{c|c|c|c|c|c}
+ \multicolumn{3}{c|}{Hadoop} & \multicolumn{3}{c}{ HPC random} \\
  \hline
@@
 \end{table}%
- At the same time, the HDFS keeps data locality \COMMENT{'keeps data locally' or 'maintains data locality'}, aiming to reduce the amount of communications, resulting less data move and, therefore, better scalability.
-Our Hadoop-in-the-Cloud cluster has no more~($\approx 40$) \COMMENT{how does 40 approximate zero?} free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
-\COMMENT{need ref?}
+HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
+Our Hadoopcluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
+\COMMENT{need ref for this!}
 

### 2014-02-24 paper-plos.tex
Set paper in PLoS Comp Biol style.

--- 2014-02-23 text1.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-\geometry{a4paper}                   		% ... or a4paper or a5paper or ... 
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or epsБ•√ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
+% Template for PLoS
+% Version 1.0 January 2009
+%
+% To compile to pdf, run:
+% latex plos.template
+% bibtex plos.template
+% latex plos.template
+% latex plos.template
+% dvipdf plos.template
+
+\documentclass[10pt]{article}
+
+% amsmath package, useful for mathematical formulas
+\usepackage{amsmath}
+% amssymb package, useful for mathematical symbols
+\usepackage{amssymb}
+
+% graphicx package, useful for including eps and pdf graphics
+% include graphics with the command \includegraphics
+\usepackage{graphicx}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+\usepackage{color}
+
+% Use doublespacing - comment out for single spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\topmargin 0.0cm
+\oddsidemargin 0.5cm
+\evensidemargin 0.5cm
+\textwidth 16cm 
+\textheight 21cm
+
+% Bold the 'Figure #' in the caption and separate it with a period
+% Captions will be left justified
+\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
+
+% Use the PLoS provided bibtex style
+\bibliographystyle{plos2009}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+\pagestyle{myheadings}
+%% ** EDIT HERE **
+\newcommand{\COMMENT}[1]{{\color{red} #1 }}
+\graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
 \usepackage[normalem]{ulem}
@@
 \usepackage{amsmath}
-\usepackage{multirow}% http://ctan.org/pkg/multirow
-\usepackage{tablefootnote}
-\usepackage{import}
+%\usepackage{multirow}% http://ctan.org/pkg/multirow
+%\usepackage{tablefootnote}
+%\usepackage{import}
 \usepackage{subcaption}
 \usepackage{caption}
-\usepackage[format=hang,font=small,labelfont=bf]{caption}
+%\usepackage[format=hang,font=small,labelfont=bf]{caption}
 
-\newcommand{\COMMENT}[1]{{\color{red} #1 }}
+%% ** EDIT HERE **
+%% PLEASE INCLUDE ALL MACROS BELOW
 
-\title{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)}
-\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth \COMMENT{[specify affiliation]}} 
-%\date{}							% Activate to display a given date or no date
-
-\graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
+%% END MACROS SECTION
 
 \begin{document}
-\maketitle
 
-\abstract{New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field.
+% Title must be 150 characters or less
+\begin{flushleft}
+{\Large
+\textbf{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data}
+}
+% Insert Author names, affiliations and corresponding author email.
+\\
+Alexey Siretskiy$^{1 \ast}$, 
+Luca Pireddu$^{2}$, 
+Ola Spjuth$^{3,4}$
+\\
+\bf{1} Department of Information Technology, Uppsala University, Uppsala, Sweden
+\\
+\bf{2} CRS4, Polaris, Pula, Italy
+\\
+\bf{3} Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden
+\\
+\bf{4} Science for Life Laboratory, Uppsala University, Uppsala, Sweden
+\\
+$\ast$ E-mail: alexey.siretskiy@it.uu.se
+\end{flushleft}
+
+% Please keep the abstract between 250 and 300 words
+\section*{Abstract}
+New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field.
 With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
@@
 As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in the future of biological data analysis.
-}
 
-\section{Introduction}
-\label{sectionI}
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLoS ONE authors please skip this step. 
+% Author Summary not valid for PLoS ONE submissions.   
+\section*{Author Summary}
+New experimental technologies generate vast amounts of data and require supercomputers to store and analyse.
+As datasets increase in volume there is a need for new computational tools and technologies that can cope with the scale of data, and in this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
+While there is an overhead of using Hadoop, we show that when sequencing data sets reach sizes of 100 gigabases then Hadoop outperforms traditional analysis.
+We also show that Hadoop scales linearly while traditional systems become network-saturated at larger data sizes.
+As sequencing data are bound to increase in the future, Hadoop is a very interesting framework to base future analysis on.
+
+%%%%%%%%%%%%
+% Introduction
+%%%%%%%%%%%
+
+\section*{Introduction}
 %The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}.
-Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 days~\cite{illumina}.
-%Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} \COMMENT{[unusual to use both footnotes and endnotes]} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) \COMMENT{huh?} storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.
+Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
+Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline that requires a high-performance computing (HPC) infrastructure to analyse.
+As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately \COMMENT{XXXX} gigabases (Gb) corresponding to 3\,terabytes (TB) of raw data in 10 days~\cite{illumina}.
 
@@
 A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-% \sout{Most of the bioinformatic programs are written for the Linux operating system.} /COMMENT{awkward but maybe it's needed} 
-Many of the most widespread tools for this purpuse, like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools}, are not made with distributed computing in mind.
+Many of the most widespread tools for this purpuse, like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools}, are not made with distributed computing in mind.
 Many others do not even have the native ability to use multiple cores.
-%\sout{on the same computer}\footnote{Samtools acquired this feature in v.0.1.19}.
-%\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications}
 
 %Current practices: How to parallelize
-
 The most common approach to speed up NGS tools is to parallelise within a compute node (multi-core parallelisation) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16.
@@
 %Introduce what we did in the manuscript.
-In this manuscript we focus on the question {\it when} \COMMENT{that's two questions} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis.
+In this manuscript we focus on the question {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. 
 Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
-We use five datasets with short reads of different sizes, align them against the appropriate reference genome, and call the variances.
+We also study execution times for datasets of different sizes and analyze Hadoop and HPC approaches from a scaling perspective.
+
+%%%%%%%%%%%%
+% Results
+%%%%%%%%%%%%
+% Results and Discussion can be combined.
+\section*{Results}
+
+We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as HPC method) (Methods).
+Our experiments then consisted of running the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage. 
+All experiments were repeated several times and the results averaged \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
 For the existing Hadoop software we propose modifications, in order to benefit fully from massively parallel nature of MR computations.
 For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short read alignment and exploiting the network fully, thereby speeding up calculations.
-The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.
 
-%The manuscript is structured as follows: \COMMENT{[odd to do this in a normal publication, but OK]} In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used.
-Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources.
-The results are discussed in Section~IV, and conclusions in Section~V.
-Supplementary materials are provided in the corresponding section.
+\subsection*{Modified preprocessing stage}
+A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet~\cite{Langmead:2009kx}.
+The only way to introduce parallelisations in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
+This, however, is lacking the massively parallel way of treating the data.
 
-\section{Methods}
-\label{sectionII}
+%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
 
-\subsection{Datasets}
+Assuming that sequencing platforms are usually in close proximity with facilities that provide both storage and computers, we considered that the data already had been downloaded to storage that Hadoop could access.
+We rewrote the Crossbow preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
+The requirement is that the FASTQ data are BZIP2 archived and accessible by Hadoop.
+For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS storage.
+BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner.
+Also, BZIP2 format is natively supported by Hadoop, i.e., there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
+
+%\footnote{other options like splittable LZO are also possible}
+
+The Hadoop streaming library offers a possibility to write code for Hadoop jobs in any programming language.
+Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive~\cite{ncbi-sra}).
+The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
+To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.
+
+In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays, e.g., while downloading data from Amazon.
+Table~\ref{table:preprocess} illustrates the benefits of our approach.
+To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
+The script (see Supplementary information) ran on a single HPC node, utilizes multiple cores, with the limiting factor being the hard disk IO performance.%, which effectively leaves with 5-7 cores of 16.
+\COMMENT{awkward}
+
+\subsection*{Accuracy of pipelines}
+Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp, correspondingly) we should not expect them to deliver perfectly matching SNP lists, but still we expect them to capture and correctly identify the mutations.
+We tested the correctness just for the smallest dataset (I).
+The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.
+
+\subsection*{Scalability of HPC and Hadoop approaches}
+To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
+The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome.
+%\footnote{\url{http://www.biostars.org/p/48781}
+%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
+During its sorting phase the lack of RAM results in spilling to the HDD.
+For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM.
+For our tasks we used the default Samtool's settings.}.
+For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for the HPC and Hadoop approaches were collected (Table~\ref{table:4}).
+
+One of the attractive sides of using Hadoop~MR is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{Langmead:2009kx,Pireddu:2011vn}, regardless the scaling nature of the underlying program (Samtools, Bowtie, etc.).
+Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
+%To stress the linear nature of scaling, both sets of points were fit using the least squares method.
+
+\subsection*{Comparing Hadoop and HPC runtime efficiency}
+Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
+One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
+In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the following function:
+$$F=T_{p}\times p,$$ 
+where~$T_{p}$ is the calculation time on~$p$ cores. \COMMENT{This trivial/obvious formula doesn't deserve such special treatment.}
+
+Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
+Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
+As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
+The parabolic extrapolation for the ratio is $1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gb (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+
+The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
+Considering the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
+%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
+being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
+\COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
+We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
+If lacking RAM, Samtools implementation swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
+Table~\ref{table:5} shows the sorting runtime performance of Samtools for the dataset~IV run with different amount of RAM available on a single HPC node.
+The large uncertainty for small amounts of RAM we believe is due to other users' load on the cluster network.
+As a result, a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling on much more economic hardware in linear time with just 12GB RAM per virtual node.
+
+ \begin{table}[hbtp]
+\small
+
+\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58 Gb BAM file).
+The measurements are done on a single HPC node with $p=16$ cores.
+Averaging is done over 5 independent simulations.
+The amount of RAM is given for all 16 cores.
+}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+RAM, GB		&	8			&		16			&			32		& 		64				\\
+timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
+\end{tabular}
+\end{center}
+\label{table:5}
+\normalsize
+\end{table}%
+
+ \subsection*{Comparing the network communication efficiency for the Hadoop and HPC approaches }
+The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
+The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
+To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gb, dataset~IV).
+
+Due to the trivially parallelisability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
+Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
+
+There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}, and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
+However, e.g., the Pmap package functions poorly for datasets larger than 20 Gb.
+To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, and compared the network performance with the standard Hadoop HDFS approach.
+We separated the mapping time, $T_{mapping}$, and the communication time, $T_{comm}$, and plotted $T_{mapping}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
+This measure is applicable to both HPC and Hadoop, however, $T_{comm}$ has a different explanation.
+For Hadoop, the short reads in FASTQ format have to be preprocessed (involving node communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data ingestion into the HDFS.
+We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
+For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.
+%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
+
+%The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
+
+\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
+It is known that Bowtie provides almost linear scaling between mapping time and dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and Figure~\ref{fig:fig1}.
+Since the ratio~$T_{mapping}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
+
+%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
+\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM.
+One can see two distinct linear stretches.
+One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
+The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
+The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
+%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
+and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
+The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
+Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
+The nodes are being allocated on the same rack, which is the default SLURM behaviour.
+
+Scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
+Allocating the nodes on random racks allows one to engage more nodes without network saturation.
+For our cluster we could go up to 30-35 nodes with perfect linear scaling.
+For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
+$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}.
+The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
+The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
+HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
+Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
+\COMMENT{need ref for this!}
+
+\subsection*{Usability aspects}
+\label{subsectionIV_2}
+
+A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying the experience for the end user.
+An alternative for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private clouds.
+We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).
+For our particular task in DNA sequencing, Cloudgene makes it easy to select data and parameters for its execution.
+Most of the data managing work is done automatically and the results can be downloaded to the client machine.
+The modular structure allows modification of the source code to adapt to the existing computing center architecture.
+For example, UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data is kept privately.
+
+%%%%%%%%%%%%
+% Discussion
+%%%%%%%%%%%%
+\section*{Discussion}
+
+In this report we have described two approaches for high-performance analysis of DNA sequencing data; one based on regular HPC using a batch system and one based on Hadoop.
+We show that Hadoop installed on a private cloud is an appealing solution both on terms of runtime performance and also in terms of usability.
+A key result is the data size where Hadoop is favorable to regular HPC batch systems, and in order to compare them we developed highly optimized pipelines for both scenarios.
+We found that dataset sizes larger than 100\,Gb is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability for this problem.
+Extrapolation to infinite dataset size reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
+
+We modified the preprocessing stage of the Crossbow software to enable it to make use of all nodes in the Hadoop system, end used a highly optimized Bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes on an HPC system.
+Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
+In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behaviour.
+This is however not of practical use in an HPC center since the network bandwidth is shared among multiple users.
+Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene) provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
+
+%%%%%%%%%%%
+% Methods
+%%%%%%%%%%%
+% You may title this section "Methods" or "Models". 
+% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
+% authors may use "Analysis" 
+\section*{Methods}
+
+\subsection*{Datasets}
 We used publicly available DNA-seq datasets~(I--III) as well as a synthetic dataset~(IV) of {\it A.thaliana}, the well-known model plant, and a dataset (V) of two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
@@
 
-\begin{table}[htdp]
-\small
-\footnotesize
-\caption{Datasets used in the comparison. }
-\begin{center}
-\begin{tabular}{|l|l|l|}
-dataset &	organism &	size in Gbases\\
-\hline
- I		&	{\it A.thaliana}	&	1.4	\\
- II	&	{\it A.thaliana}	&	7.0\\
-  III	&	{\it A.thaliana}	&	30.0	\\
- IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
- V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
-
-\end{tabular}
-\end{center}
-\label{table:datasets}
-\normalsize
-\end{table}%
-
-\subsection{Analysis pipelines}
-We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as HPC).
-Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage.
-All experiments were repeated several times and the results averaged \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
+\subsection*{Analysis pipelines}
+We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data; one for Hadoop and one for HPC.
 Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (e.g., GATK~\cite{gatk}), we decided to create the analysis pipeline as simple as possible to be able to make a better comparison with the one implemented for Hadoop.
-
-%pass \COMMENT{huh?} the same stages on Hadoop and HPC:
 
@@
 \subitem SNP calling: Samtools~ver.~0.1.19
-\item Hadoop approach: Crossbow~\cite{crossbow}
+\item Hadoop approach: Crossbow
 \subitem Short read alignment: Bowtie~ver.~0.12.8
@@
 
-\subsection{Computational resources}
+\subsection*{Computational resources}
 To run the HPC analysis pipeline we used a computational cluster at UPPMAX, which is heavily used for analysis of NGS-data~\cite{lampa}, on nodes equipped with 8 dual-core CPUs.
@@
 
-\section{Results}
-\label{sectionIII}
+%%%%%%%%%%%
+% Acknowledgments
+%%%%%%%%%%%
+% Do NOT remove this, even if you are not including acknowledgments
+\section*{Acknowledgments}
+The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
+This work was supported by the Swedish strategic research programme eSSENCE.
+We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
 
-\subsection{Modified preprocessing stage}
-A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}.
-The only way to introduce parallelisations in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
-This, however, is lacking the massively parallel way of treating the data.
+%%%%%%%%%%%
+% Bibliography
+%%%%%%%%%%%
+%\section*{References}
+% The bibtex filename
+\bibliography{paper-plos}
 
-%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
+%%%%%%%%%%%
+% Supplementary material
+%%%%%%%%%%%
+\section*{Supplementary material}
 
-Assuming that sequencing platforms are usually in close proximity with facilities that provide both storage and computers, we considered that the data already had been downloaded to storage that Hadoop could access.
-We rewrote the Crossbow preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
-The requirement is that the FASTQ data are BZIP2 archived and accessible by Hadoop.
-For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS storage.
-BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner.
-Also, BZIP2 format is natively supported by Hadoop, i.e., there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
+\subsection*{Datasets}
 
-%\footnote{other options like splittable LZO are also possible}
+The datasets used in the paper are publicly available at:
 
-The Hadoop streaming library offers a possibility to write Mapper and Reducer \COMMENT{Introduce in introduction} for Hadoop jobs in any programming language.
-Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive).\COMMENT{ref?}
-The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
-To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.
+I: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0\\/data/reads/Schneeberger.2009/Schneeberger.2009.single\_end.gz}
 
-In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays, e.g., while downloading data from Amazon.
-Table~\ref{table:preprocess} illustrates the benefits of our approach.
-To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
-The script (see Supplementary information) ran on a single HPC node, utilizes multiple cores, with the limiting factor being the hard disk IO performance.%, which effectively leaves with 5-7 cores of 16.
-\COMMENT{awkward}
+II: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads1.fq.gz, http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads2.fq.gz}
 
-\begin{table}[tdp]
+III: \url{ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611084//SRR611084.sra, ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611085//SRR611085.sra}
+
+IV: artificial pair-ended dataset for {\it A.thaliana} created with the {\tt wgsim} program from the Samtools package.
+
+V: \url{http://www.ncbi.nlm.nih.gov/sra/SRX148888}
+
+\subsection*{Reference genomes}
+\begin{itemize}
+\item TAIR10 for datasets II-IV \url{ftp://ftp.arabidopsis.org/home/tair/Sequences/whole\_chromosomes/*.fas}
+\item TAIR8 for dataset I \url{ftp://ftp.arabidopsis.org/home/tair/Genes/TAIR8\_genome\_release/}
+\item H.sapiens, NCBI v37 \url{ftp://ftp.ccb.jhu.edu/pub/data/bowtie\_indexes/h\_sapiens\_37\_asm.ebwt.zip}
+\end{itemize}
+
+\subsection*{Description of computational facilities}
+
+\begin{enumerate}
+%\item  HPC1:
+%The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
+The data and reference genomes were read and written to a parallel shared storage system.
+
+\item 
+HPC:
+Multinode short read mapping was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
+
+\item Storage: 
+Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600 storage boxes and an additional node for metadata handling.
+In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
+It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
+
+\item Our Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud management system.
+Each node in this deployment was equipped with dual 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
+The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
+Note that the physical hardware provided less RAM than desired.
+Each VM node has 7 cores where each can use less than 2 GB of RAM, which is little for Hadoop.
+We expect much better performance with twice as much memory.
+\end{enumerate}
+
+\subsection*{Bash script for the preprocessing stage}
+
+\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
+
+\COMMENT{Journal abbreviations in references are inconsistant.
+The journal name for ref 2 is wrong.
+Capitalization of refs 4, 14, 16, 18, 19, 25, 27, etc are wrong.
+All refs should be checked for accuracy and style consistancy.
+Many or all footnotes should be changed to endnotes (references).}
+
+%%%%%%%%%%%
+% Figure Legends
+%%%%%%%%%%%
+\section*{Figure Legends}
+
+\begin{figure}[!ht]
+	\input{./figures/fig1/fig1.tex}
+	\caption{Calculation time for selected datasets (labeled with roman numerals) for the 56 cores Hadoop cluster shows a linear scaling with dataset size. }
+	\label{fig:fig1}
+\end{figure}
+
+\begin{figure}[!ht]
+	\input{./figures/fig2/fig2.tex}
+	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases. Calculations were carried out for 56 cores Hadoop and 16 cores HPC cluster correspondingly.
+The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are labeled with roman numerals.}
+	\label{fig:fig2}
+\end{figure}
+
+\begin{figure}[!ht]
+ 	\begin{subfigure}[b]{0.4\textwidth}
+        	\includegraphics[width=\textwidth]{a.png}
+		\subcaption{}
+	\end{subfigure}
+	\hspace{0.5cm}
+	\begin{subfigure}[b]{0.6\textwidth}
+	        	\includegraphics[width=\textwidth]{c.png}
+		\subcaption{}
+	\end{subfigure}
+%	\begin{subfigure}[b]{0.6\textwidth}
+%		\includegraphics[width=\textwidth]{b.png}
+%		\subcaption{Cloudgene: specifying job parameters}
+%	\end{subfigure}
+	\caption{An example of a job setup with the graphical Hadoop front-end Cloudgene, providing a smooth user experience even for novice users.
+\textbf{a)} Pipeline selection - in our case containing the Crossbow pipeline.
+\textbf{b)} The UPPMAX-adapted functionality to browsing and import data from the user's home folder in the shared file system.}
+	\label{fig:fig4}
+\end{figure}
+
+\begin{figure}[!ht]
+	\small
+	\input{./figures/fig9/fig3.tex}
+	\normalsize
+	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
+Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
+%One  can see the differences in the Hadoop and HPC communication patterns: for Hadoop the ratio is almost constant, while for HPC cluster there are two defined regions.
+%The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
+%The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
+%linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
+Linear fit  done with the least-squares method. \COMMENT{Figure legend too complex. Y-axis label inconsistant with previous figure.}}
+	\label{fig:fig3}
+\end{figure}
+
+%%%%%%%%%%%
+% Tables
+%%%%%%%%%%%
+\section*{Tables}
+
+\begin{table}[!ht]
 \small
@@
 \begin{tabular}{l|c|c|c}
-Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
+Dataset			&		II (7 Gb )	& III (30 Gb)	& IV (100 Gb)\\
 \hline
@@
 
-\subsection{Accuracy of pipelines}
-Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp, correspondingly) we should not expect them to deliver perfectly matching SNP lists, but still we expect them to capture and correctly identify the mutations.
-We tested the correctness just for the smallest dataset (I).
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.
-
-\subsection{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
-The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome.
-%\footnote{\url{http://www.biostars.org/p/48781}
-%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
-During its sorting phase the lack of RAM results in spilling to the HDD.
-For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM.
-For our tasks we used the default Samtool's settings.}.
-For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for the HPC and Hadoop approaches were collected (Table~\ref{table:4}).
-
-One of the attractive sides of using Hadoop~MR is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie, etc.).
-Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
-%To stress the linear nature of scaling, both sets of points were fit using the least squares method.
-
-\begin{table}[tb]
+\begin{table}[!ht]
 \small
@@
 
-\begin{table}[tdp]
+\begin{table}[!ht]
 \small
@@
 
-data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
+data, Gb		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
 				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
@@
 
-\begin{figure}[bt]
-	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time for selected datasets (labeled with roman numerals) for the 56 cores Hadoop cluster shows a linear scaling with dataset size. }
-	\label{fig:fig1}
-\end{figure}
+\begin{table}[!ht]
+\small
+\footnotesize
+\caption{Datasets used in the comparison. }
+\begin{center}
+\begin{tabular}{|l|l|l|}
+dataset &	organism &	size in Gb\\
+\hline
+ I		&	{\it A.thaliana}	&	1.4	\\
+ II	&	{\it A.thaliana}	&	7.0\\
+  III	&	{\it A.thaliana}	&	30.0	\\
+ IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
+ V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
 
-\begin{figure}[bt]
-	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases. Calculations were carried out for 56 cores Hadoop and 16 cores HPC cluster correspondingly.
-The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are labeled with roman numerals.}
-	\label{fig:fig2}
-\end{figure}
-
-\subsection{Comparing Hadoop and HPC runtime efficiency}
-Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the following function:
-$$F=T_{p}\times p,$$ 
-where~$T_{p}$ is the calculation time on~$p$ cores. \COMMENT{This trivial/obvious formula doesn't deserve such special treatment.}
-
-Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
-Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
-As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for the ratio is $1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
-
-The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
-Considering the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
-%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
-being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
-\COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
-We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
-If lacking RAM, Samtools implementation swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
-Table~\ref{table:5} shows the sorting runtime performance of Samtools for the dataset~IV run with different amount of RAM available on a single HPC node.
-The large uncertainty for small amounts of RAM we believe is due to other users' load on the cluster network.
-As a result, a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling on much more economic hardware in linear time with just 12GB RAM per virtual node.
-
- \begin{table}[hbtp]
-\small
-
-\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58Gb BAM file).
-The measurements are done on a single HPC node with $p=16$ cores.
-Averaging is done over 5 independent simulations.
-The amount of RAM is given for all 16 cores.
-}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-RAM, GB		&	8			&		16			&			32		& 		64				\\
-timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
 \end{tabular}
 \end{center}
-\label{table:5}
+\label{table:datasets}
 \normalsize
@@
 
- \subsection{Comparing the network communication efficiency for the Hadoop and HPC approaches }
-The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
-The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
-To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gb, dataset~IV).
-
-Due to the trivially parallelisability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
-
-There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}, and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
-However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
-To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, and compared the network performance with the standard Hadoop HDFS approach.
-We separated the mapping time, $T_{mapping}$, and the communication time, $T_{comm}$, and plotted $T_{mapping}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
-This measure is applicable to both HPC and Hadoop, however, $T_{comm}$ has a different explanation.
-For Hadoop, the short reads in FASTQ format have to be preprocessed (involving node communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data ingestion into the HDFS.
-We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.
-%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
-
-%The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
-
-\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that Bowtie provides almost linear scaling between mapping time and dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}.
-Since the ratio~$T_{mapping}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
-
-%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
-\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM.
-One can see two distinct linear stretches.
-One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
-The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
-%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
-and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
-The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
-Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated on the same rack, which is the default SLURM behaviour.
-
-Scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
-Allocating the nodes on random racks allows one to engage more nodes without network saturation.
-For our cluster we could go up to 30-35 nodes with perfect linear scaling.
-For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}.
-The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
-The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
-
-\begin{figure}
-	\small
-	\input{./figures/fig9/fig3.tex}
-	\normalsize
-	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
-Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
-%One  can see the differences in the Hadoop and HPC communication patterns: for Hadoop the ratio is almost constant, while for HPC cluster there are two defined regions.
-%The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
-%The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
-%linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
-Linear fit  done with the least-squares method. \COMMENT{Figure legend too complex. Y-axis label inconsistant with previous figure.}}
-	\label{fig:fig3}
-\end{figure}
-
-\begin{table}[htdp]
+\begin{table}[!ht]
 \caption{Timings  for mapping and the ratio~$T_{mapping}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
@@
 \end{table}%
-HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
-Our Hadoopcluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
-\COMMENT{need ref for this!}
-
-\subsection{Usability aspects}
-\label{subsectionIV_2}
-
-A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying the experience for the end user.
-One of the alternatives for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private clouds.
-We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}(a)).
-For our particular task in DNA sequencing, Cloudgene provides the intuitive interface making one easy to follow.
-Most of the data managing work is done automatically and the results can be downloaded to the client machine.
-The modular structure allows modification of the source code to adapt to the existing computing center’s architecture, Figure~\ref{fig:fig4}(b).
-For example, UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will be held locally, reducing the amount of unnecessary risks.
-
-\begin{figure}
- 	\begin{subfigure}[b]{0.4\textwidth}
-        	\includegraphics[width=\textwidth]{a.png}
-		\subcaption{}
-	\end{subfigure}
-	\hspace{0.5cm}
-	\begin{subfigure}[b]{0.6\textwidth}
-	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{}
-	\end{subfigure}
-%	\begin{subfigure}[b]{0.6\textwidth}
-%		\includegraphics[width=\textwidth]{b.png}
-%		\subcaption{Cloudgene: specifying job parameters}
-%	\end{subfigure}
-	\caption{An example of a job setup with the graphical Hadoop front-end Cloudgene, providing a smooth user experience even for novice users.
-\textbf{a)} Pipeline selection - in our case containing the Crossbow pipeline.
-\textbf{b)} The UPPMAX-adapted functionality to browsing and import data from the user's home folder in the shared file system.}
-	\label{fig:fig4}
-\end{figure}
-
-\section{Conclusions}
-\label{sectionV}
-
-In this report we have described two approaches for high-performance analysis of DNA sequencing data; one based on regular HPC using a batch system and one based on Hadoop.
-We show that Hadoop installed on a private cloud is an appealing solution both on terms of runtime performance and also in terms of usability.
-A key result is the data size where Hadoop is favorable to regular HPC batch systems, and we developed highly optimized pipelines for both scenarios.
-We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability (Figure~\ref{fig:fig1}).
-Extrapolation to infinite dataset size (Figure~\ref{fig:fig2}) reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
-
-To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow/Myrna  were suggested.
-The calculation time reduction can be observed in the Table~\ref{table:preprocess}.
-\COMMENT{So, is Hadoop better after the modifications for infinite datasets?}
-%A state-of-art Bash script was created to engage multiple nodes (up to 928 cores) in the HPC approach to align the Illumina-produced short reads with almost linear speed-ups.
-
-%We concentrated ourselves on the Hadoop  in the private cloud installation, Hadoop-in-the-Cloud, orchestrated by the OpenNebula. 
-%Picking the appropriate program settings we found out that DNA-seq dataset size larger than 100\,Gbases is suitable for Hadoop, with competitive  execution time compare to  that of HPC.
-%The scaling graphs (Figure~\ref{fig:fig1}) confirms the known fact for almost linear scalability of Hadoop.
-%The extrapolation to the infinite dataset size (Figure~\ref{fig:fig2}) revealed, that HPC, however, provides the results faster, given the same amount of resources, than the Hadoop-in-the-Cloud.
-
-Exploiting the embarrassingly parallel nature of the short reads mapping we used a highly optimized Bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes (Figure~\ref{fig:fig3}).
-Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behaviour.
-\COMMENT{Should maybe mention somewhere that this sort of cheating.} Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene), provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
-
-\section{Acknowledgements}
-The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE~\url{http://essenceofescience.se/home/}.
-We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage.
-We also thank Jonas Hagberg (BILS, Stockholm, Sweden), for implementing the Cloudgene extensions to import data from UPPMAX filesystem.
-
-\section{Supplementary material}
-
-\subsection{Datasets}
-
-The datasets used in the paper are publicly available at:
-
-I: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0\\/data/reads/Schneeberger.2009/Schneeberger.2009.single\_end.gz}
-
-II: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads1.fq.gz, http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads2.fq.gz}
-
-III: \url{ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611084//SRR611084.sra, ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611085//SRR611085.sra}
-
-IV: artificial pair-ended dataset for {\it A.thaliana} created with the {\tt wgsim} program from the Samtools package.
-
-V: \url{http://www.ncbi.nlm.nih.gov/sra/SRX148888}
-
-\subsection{Reference genomes}
-\begin{itemize}
-\item TAIR10 for datasets II-IV \url{ftp://ftp.arabidopsis.org/home/tair/Sequences/whole\_chromosomes/*.fas}
-\item TAIR8 for dataset I \url{ftp://ftp.arabidopsis.org/home/tair/Genes/TAIR8\_genome\_release/}
-\item H.sapiens, NCBI v37 \url{ftp://ftp.ccb.jhu.edu/pub/data/bowtie\_indexes/h\_sapiens\_37\_asm.ebwt.zip}
-\end{itemize}
-
-\subsection{Description of computational facilities}
-
-\begin{enumerate}
-%\item  HPC1:
-%The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
-The data and reference genomes were read and written to a parallel shared storage system.
-
-\item 
-HPC:
-Multinode short read mapping was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
-
-\item Storage: 
-Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600 storage boxes and an additional node for metadata handling.
-In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
-It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
-
-\item Our Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud management system.
-Each node in this deployment was equipped with dual 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
-The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
-Note that the physical hardware provided less RAM than desired.
-Each VM node has 7 cores, each of each can use less than 2 GB of RAM, which is little for Hadoop.
-We would expect much better performance with twice as much memory, as recommended.
-\end{enumerate}
-
-\subsection{Bash script for the preprocessing stage}
-
-\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
-
-\COMMENT{Journal abbreviations in references are inconsistant.
-The journal name for ref 2 is wrong.
-Capitalization of refs 4, 14, 16, 18, 19, 25, 27, etc are wrong.
-All refs should be checked for accuracy and style consistancy.
-Many or all footnotes should be changed to endnotes (references).}
-
-\bibliography{text1}
-\bibliographystyle{unsrt}
 

### 2014-02-24 text1.tex
Set paper in PLoS Comp Biol style.

@@
 %The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}.
+Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
 Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.
@@
 HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
-Our Hadoopcluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
+Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
 \COMMENT{need ref for this!}
@@
 A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying the experience for the end user.
-One of the alternatives for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private clouds.
-We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}(a)).
-For our particular task in DNA sequencing, Cloudgene provides the intuitive interface making one easy to follow.
+An alternative for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private clouds.
+We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).
+For our particular task in DNA sequencing, Cloudgene makes it easy to select data and parameters for its execution.
 Most of the data managing work is done automatically and the results can be downloaded to the client machine.
-The modular structure allows modification of the source code to adapt to the existing computing center’s architecture, Figure~\ref{fig:fig4}(b).
-For example, UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data will be held locally, reducing the amount of unnecessary risks.
+The modular structure allows modification of the source code to adapt to the existing computing center architecture.
+For example, UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data is kept privately.
 
@@
 
-\section{Conclusions}
+\section{Discussions/Conclusions}
 \label{sectionV}
@@
 We show that Hadoop installed on a private cloud is an appealing solution both on terms of runtime performance and also in terms of usability.
-A key result is the data size where Hadoop is favorable to regular HPC batch systems, and we developed highly optimized pipelines for both scenarios.
-We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability (Figure~\ref{fig:fig1}).
-Extrapolation to infinite dataset size (Figure~\ref{fig:fig2}) reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
+A key result is the data size where Hadoop is favorable to regular HPC batch systems, and in order to compare them we developed highly optimized pipelines for both scenarios.
+We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability for this problem.
+Extrapolation to infinite dataset size reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
 
-To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow/Myrna  were suggested.
+%To increase the performance of the existing Hadoop software, the modification to the preprocessing stage of the Crossbow/Myrna  were suggested.
 The calculation time reduction can be observed in the Table~\ref{table:preprocess}.
@@
 
-Exploiting the embarrassingly parallel nature of the short reads mapping we used a highly optimized Bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes (Figure~\ref{fig:fig3}).
+We modified the preprocessing stage of the Crossbow software to enable it to make use of all nodes in the Hadoop system, end used a highly optimized Bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes on an HPC system.
 Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
 In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behaviour.
-\COMMENT{Should maybe mention somewhere that this sort of cheating.} Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene), provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
+This is however not of practical use in an HPC center since the network bandwidth is shared among multiple users.
+Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene) provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
 
@@
 The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE~\url{http://essenceofescience.se/home/}.
-We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage.
-We also thank Jonas Hagberg (BILS, Stockholm, Sweden), for implementing the Cloudgene extensions to import data from UPPMAX filesystem.
+This work was supported by the Swedish strategic research programme eSSENCE.
+We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
 
@@
 Note that the physical hardware provided less RAM than desired.
-Each VM node has 7 cores, each of each can use less than 2 GB of RAM, which is little for Hadoop.
-We would expect much better performance with twice as much memory, as recommended.
+Each VM node has 7 cores where each can use less than 2 GB of RAM, which is little for Hadoop.
+We expect much better performance with twice as much memory.
 \end{enumerate}

### 2014-02-25 paper-plos.tex
modification after meeting on 25 of February

@@
 \newcommand{\COMMENT}[1]{{\color{red} #1 }}
+\newcommand{\COM}[1]{{\color{blue} #1 }}
+
 \graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
@@
 {\Large
-\textbf{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data}
+\textbf{A quantitative assessment of the Hadoop \textcolor{blue}{and HPC} framework for analysing massively parallel DNA sequencing data}
 }
@@
 Hadoop, emerging in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-Herein we report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
+Herein we \textcolor{blue}{introduce the metrics and} report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
 In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts.
@@
 As datasets increase in volume there is a need for new computational tools and technologies that can cope with the scale of data, and in this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
-While there is an overhead of using Hadoop, we show that when sequencing data sets reach sizes of 100 gigabases then Hadoop outperforms traditional analysis.
+While there is an overhead of using Hadoop, we show that when sequencing data sets reach sizes of 100 gigabases then Hadoop \textcolor{blue}{(worth to use perhaps.
+Outperforms is too strong)} outperforms traditional analysis.
 We also show that Hadoop scales linearly while traditional systems become network-saturated at larger data sizes.
@@
 Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline that requires a high-performance computing (HPC) infrastructure to analyse.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately \COMMENT{XXXX} gigabases (Gb) corresponding to 3\,terabytes (TB) of raw data in 10 days~\cite{illumina}.
+As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gb) corresponding to 2\,terabytes (TB) of raw data in 3 days~\cite{illumina}.
 
@@
 %Current practices: How to parallelize
-The most common approach to speed up NGS tools is to parallelise within a compute node (multi-core parallelisation) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16.
-\COMMENT{confirm this!} For the tools which do not support OMP natively, e.g.
-Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel Linux utility~\cite{gnuparallel}.
+The most common approach to speed up NGS tools is to parallelise within a compute node (multi-core parallelisation) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16, see~\cite{top500}.
+For the tools which do not support OMP natively, e.g.
+Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel Linux utility~\cite{Tange2011a}.
 Of great importance is that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and multi-node parallelisation generally are not bound by these constraints.
@@
 This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
-The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. \COMMENT{Fix bib error (can't have both editor and author for this ref type)} 
+The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}.
+\textcolor{blue}{Please clarify, I checked the bib file and...
+found nothing...}\COMMENT{Fix bib error (can't have both editor and author for this ref type)}
 The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
@@
 Our experiments then consisted of running the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage. 
-All experiments were repeated several times and the results averaged \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
+All experiments were repeated several times and the results averaged \textcolor{blue}{to use medians I had to run much more experiments...
+each takes 10 hours ....
+not possible.
+} \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
 For the existing Hadoop software we propose modifications, in order to benefit fully from massively parallel nature of MR computations.
@@
 A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet~\cite{Langmead:2009kx}.
-The only way to introduce parallelisations in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
+The only way to introduce parallelisation in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
 This, however, is lacking the massively parallel way of treating the data.
@@
 We rewrote the Crossbow preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
-The requirement is that the FASTQ data are BZIP2 archived and accessible by Hadoop.
-For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS storage.
+The requirement is that the FASTQ data are i.e.
+BZIP2 archived\COMMENT {there was a footnote about LZO.
+Why was is bad?}, (however, any splittable archive format is suitable) and accessible by Hadoop.
+For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS.
 BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner.
@@
 To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
-The script (see Supplementary information) ran on a single HPC node, utilizes multiple cores, with the limiting factor being the hard disk IO performance.%, which effectively leaves with 5-7 cores of 16.
-\COMMENT{awkward}
+The script (see Supplementary information) ran on a single HPC node, utilizes multiple cores, with the limiting factor being \sout{the hard disk IO performance.}\COM{More correctly to say, that the pipe is not filled completely.
+i.e.
+the degree of achieved parallelization does not utilize all 16 cores.
+Perhaps not the best bash commands etc.}
+%, which effectively leaves with 5-7 cores of 16. \COMMENT{awkward}
 
@@
 Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp, correspondingly) we should not expect them to deliver perfectly matching SNP lists, but still we expect them to capture and correctly identify the mutations.
-We tested the correctness just for the smallest dataset (I).
+We tested the correctness just for the smallest dataset~(I).
 The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.
@@
 \subsection*{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
-The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome.
+To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}).
+\COMMENT{Consider replacing with a graph.} \COM{no, lets leave it as a table...
+with the better caption.}
+The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome  as suggested at~\cite{biostars_samtools}.
 %\footnote{\url{http://www.biostars.org/p/48781}
+\COM{here was a footnote referring to biostars site. Ok, if the footnotes are not welcome, lets then have it as a reference. That code is not trivial. Maybe we can leave the ref?}
 %\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
@@
 
-One of the attractive sides of using Hadoop~MR is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{Langmead:2009kx,Pireddu:2011vn}, regardless the scaling nature of the underlying program (Samtools, Bowtie, etc.).
+One of the attractive sides to run i.e.
+Samtools, Bowtie, etc.
+in the  Hadoop~MR framework is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{Langmead:2009kx,Pireddu:2011vn}, regardless the scaling nature of the underlying program.
 Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
@@
 One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the following function:
-$$F=T_{p}\times p,$$ 
-where~$T_{p}$ is the calculation time on~$p$ cores. \COMMENT{This trivial/obvious formula doesn't deserve such special treatment.}
+In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the function showing the amount of used core-hours:
+$F=T_{p}\times p,$ 
+where~$T_{p}$ is the calculation time on~$p$ cores.
 
@@
 As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for the ratio is $1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gb (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gb (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
-The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one.
-Considering the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
+The simplest curve to fit the points nicely is a parabolic function, importantly not a constant. 
+Since the Hadoop approach for the datasets reveals the linear scaling (Figure~\ref{fig:fig1}), a deviation of the ratio from the constant shows the 
+lack of scalability of the HPC approach.
+%Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
 %\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
@@
 We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
-If lacking RAM, Samtools implementation swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
+If lacking RAM, Samtools \sout{implementation} swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
 Table~\ref{table:5} shows the sorting runtime performance of Samtools for the dataset~IV run with different amount of RAM available on a single HPC node.
@@
 HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
-Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS.
-\COMMENT{need ref for this!}
+Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS\cite{lin2010, Hadoop:Guide}.
 
@@
 	\normalsize
-	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset IV as a function of reciprocal cluster size $1/N$.
+	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset~IV as a function of reciprocal number of nodes~$1/N$.
 Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
-%One  can see the differences in the Hadoop and HPC communication patterns: for Hadoop the ratio is almost constant, while for HPC cluster there are two defined regions.
+Linear fit  to $ax+b$ was done with the least-squares method.
+The estimated values for Hadoop approach are $(a,b)\approx(0.95,2.07)$.
+One  can see two defined linear regions for the HPC approach with very different tangents for linear and non-linear scalings correspondingly.
+For  the ``HPC SLURM'' $(a,b)\approx (0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random'' $(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.
+ The constant term for HPC goes below unity, i.e. for the large number of nodes the communication will take more time than actual mapping. 
 %The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
@@
 %linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
-Linear fit  done with the least-squares method. \COMMENT{Figure legend too complex. Y-axis label inconsistant with previous figure.}}
+}
 	\label{fig:fig3}
@@
 \small
-\caption{Calculation time for Dataset I executed on different number of cores on a node of HPC cluster. \COMMENT{awkward title, consider replacing with graph (and new title)}}
+\caption{Timings for Dataset I processed on different number of cores of a HPC node. \COMMENT{awkward title, consider replacing with graph (and new title)} \COM{better?}}
 \begin{center}

### 2014-02-25 paper-plos.tex
Updates to remove comments and changed to Gbases and alignment

@@
 {\Large
-\textbf{A quantitative assessment of the Hadoop \textcolor{blue}{and HPC} framework for analysing massively parallel DNA sequencing data}
+\textbf{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
 }
@@
 New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field.
-With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
+With increasing data volumes comes the necessity to analyzee data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
 Hadoop, emerging in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-Herein we \textcolor{blue}{introduce the metrics and} report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
+Herein we introduce metrics and report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
 In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts.
-From the observered scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100\,gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution.
+From the observed scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100\,gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution.
 As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in the future of biological data analysis.
@@
 \section*{Author Summary}
-New experimental technologies generate vast amounts of data and require supercomputers to store and analyse.
+New experimental technologies generate vast amounts of data and require supercomputers to store and analyze.
 As datasets increase in volume there is a need for new computational tools and technologies that can cope with the scale of data, and in this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
-While there is an overhead of using Hadoop, we show that when sequencing data sets reach sizes of 100 gigabases then Hadoop \textcolor{blue}{(worth to use perhaps.
-Outperforms is too strong)} outperforms traditional analysis.
+While there is an overhead of using Hadoop, we show that when sequencing data sets reach sizes of 100 gigabases then Hadoop becomes attractive compared with traditional methods.
 We also show that Hadoop scales linearly while traditional systems become network-saturated at larger data sizes.
@@
 Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
-Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline that requires a high-performance computing (HPC) infrastructure to analyse.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gb) corresponding to 2\,terabytes (TB) of raw data in 3 days~\cite{illumina}.
+Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline that requires a high-performance computing (HPC) infrastructure to analyze.
+As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gbases) corresponding to 2\,terabytes (TB) of raw data in 3 days~\cite{illumina}.
 
 %Current practices: HPC systems
-A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Many of the most widespread tools for this purpuse, like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools}, are not made with distributed computing in mind.
+A common step of NGS data analysis consists of alignment short reads to a reference sequence and then finding the genetic variations specific to the sample.
+Many of the most widespread tools for this purpose, like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools}, are not made with distributed computing in mind.
 Many others do not even have the native ability to use multiple cores.
@@
 The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}.
-\textcolor{blue}{Please clarify, I checked the bib file and...
-found nothing...}\COMMENT{Fix bib error (can't have both editor and author for this ref type)}
 The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
@@
 Our experiments then consisted of running the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage. 
-All experiments were repeated several times and the results averaged \textcolor{blue}{to use medians I had to run much more experiments...
-each takes 10 hours ....
-not possible.
-} \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform.
+All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
 For the existing Hadoop software we propose modifications, in order to benefit fully from massively parallel nature of MR computations.
@@
 The requirement is that the FASTQ data are i.e.
-BZIP2 archived\COMMENT {there was a footnote about LZO.
-Why was is bad?}, (however, any splittable archive format is suitable) and accessible by Hadoop.
+BZIP2 archived (however, any splittable archive format is suitable) and accessible by Hadoop.
 For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS.
@@
 To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
-The script (see Supplementary information) ran on a single HPC node, utilizes multiple cores, with the limiting factor being \sout{the hard disk IO performance.}\COM{More correctly to say, that the pipe is not filled completely.
+The script utilizes multiple cores and was executed on a single HPC node (see Supplementary information).
+%ran on a single HPC node, utilizes multiple cores, with the limiting factor being \sout{the hard disk IO performance.}\COM{More correctly to say, that the pipe is not filled completely.
 i.e.
@@
 To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). 
-\COMMENT{Consider replacing with a graph.} \COM{no, lets leave it as a table...
-with the better caption.}
-The aligning \COMMENT{mapping written in table, be consistent} process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelising analysis by chromosome  as suggested at~\cite{biostars_samtools}.
+The alignment process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelizing analysis by chromosome as previously suggested~\cite{biostars_samtools}.
 %\footnote{\url{http://www.biostars.org/p/48781}
-\COM{here was a footnote referring to biostars site. Ok, if the footnotes are not welcome, lets then have it as a reference. That code is not trivial. Maybe we can leave the ref?}
+%\COM{here was a footnote referring to biostars site. Ok, if the footnotes are not welcome, lets then have it as a reference. That code is not trivial. Maybe we can leave the ref?}
 %\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
@@
 As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gb (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
@@
 \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
-We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
+We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual alignment with Bowtie.
 If lacking RAM, Samtools \sout{implementation} swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
@@
 
- \begin{table}[hbtp]
-\small
-
-\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58 Gb BAM file).
-The measurements are done on a single HPC node with $p=16$ cores.
-Averaging is done over 5 independent simulations.
-The amount of RAM is given for all 16 cores.
-}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-RAM, GB		&	8			&		16			&			32		& 		64				\\
-timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
-\end{tabular}
-\end{center}
-\label{table:5}
-\normalsize
-\end{table}%
-
  \subsection*{Comparing the network communication efficiency for the Hadoop and HPC approaches }
@@
 The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
-To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gb, dataset~IV).
+To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gbases, dataset~IV).
 
 Due to the trivially parallelisability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
+Reducing the size of each data chunk reduces the aligner  job,~$T_{alignment}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
 
 There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}, and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
-However, e.g., the Pmap package functions poorly for datasets larger than 20 Gb.
+However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
 To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, and compared the network performance with the standard Hadoop HDFS approach.
-We separated the mapping time, $T_{mapping}$, and the communication time, $T_{comm}$, and plotted $T_{mapping}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
+We separated the alignment time, $T_{alignment}$, and the communication time, $T_{comm}$, and plotted $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
 This measure is applicable to both HPC and Hadoop, however, $T_{comm}$ has a different explanation.
@@
 We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
-For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.
+For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual alignment happens, and the sending of the aligned SAM files back to the delivery location over the network.
 %\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
@@
 
-\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that Bowtie provides almost linear scaling between mapping time and dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and Figure~\ref{fig:fig1}.
-Since the ratio~$T_{mapping}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
+\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: One can see that the ratio~$T_{alignment}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
+It is known that Bowtie provides almost linear scaling between alignment time and dataset chunk size~$D$:~$T_{alignment}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and Figure~\ref{fig:fig1}.
+Since the ratio~$T_{alignment}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
 
@@
 The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
-%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
-and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
-The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
-Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
+%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
+and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
+The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbasesit/sec.
+Ten nodes will consume the data with the rate of~10Gbasesit/sec, which is the limiting speed for the standard 10Gbasesit Ethernet cable connecting the cluster's rack with the switch.
 The nodes are being allocated on the same rack, which is the default SLURM behaviour.
@@
 $5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}.
-The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
+The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbasesit/sec being saturated.
 The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
@@
 A key result is the data size where Hadoop is favorable to regular HPC batch systems, and in order to compare them we developed highly optimized pipelines for both scenarios.
-We found that dataset sizes larger than 100\,Gb is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability for this problem.
+We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability for this problem.
 Extrapolation to infinite dataset size reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
 
-We modified the preprocessing stage of the Crossbow software to enable it to make use of all nodes in the Hadoop system, end used a highly optimized Bash script to compare the scaling relations between the ratio of the mapping time to the communication time as a function of the reciprocal number of nodes on an HPC system.
+We modified the preprocessing stage of the Crossbow software to enable it to make use of all nodes in the Hadoop system, end used a highly optimized Bash script to compare the scaling relations between the ratio of the alignment time to the communication time as a function of the reciprocal number of nodes on an HPC system.
 Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
@@
 Where available these features were used to improve the analysis speed.
-The exact workflow used is available in \COMMENT{github?} the code repository created for this work~\cite{code_repo}.
-\COMMENT{missing ref}
+The exact workflow used is available from github~\cite{code_repo}.
 
@@
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available in \COMMENT{github?} the code repository~\cite{code_repo}).
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from github~\cite{code_repo}).
 
@@
 %\item  HPC1:
-%The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbit/s uplink.
+%The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbasesit/s uplink.
 The data and reference genomes were read and written to a parallel shared storage system.
@@
 HPC:
-Multinode short read mapping was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
+Multinode short read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbasesit/s uplink.
 
@@
 In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
-It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
+It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbasesit per second.
 
@@
 	\normalsize
-	\caption{Ratios of the mapping time~$T_{mapping}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset~IV as a function of reciprocal number of nodes~$1/N$.
+	\caption{Ratios of the alignment time~$T_{alignment}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset~IV as a function of reciprocal number of nodes~$1/N$.
 Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
@@
 For  the ``HPC SLURM'' $(a,b)\approx (0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random'' $(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.
- The constant term for HPC goes below unity, i.e. for the large number of nodes the communication will take more time than actual mapping. 
+ The constant term for HPC goes below unity, i.e. for the large number of nodes the communication will take more time than actual alignment. 
 %The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
 %The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
-%linearly to the reciprocal cluster size, and becomes more of less a constant, while the mapping time continues to satisfy~$T_{mapping}\propto 1/N$.
+%linearly to the reciprocal cluster size, and becomes more of less a constant, while the alignment time continues to satisfy~$T_{alignment}\propto 1/N$.
 }
@@
 
+ \begin{table}[hbtp]
+\small
+
+\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58 Gbases BAM file).
+The measurements are done on a single HPC node with $p=16$ cores.
+Averaging is done over 5 independent simulations.
+The amount of RAM is given for all 16 cores.
+}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+RAM, GB		&	8			&		16			&			32		& 		64				\\
+timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
+\end{tabular}
+\end{center}
+\label{table:5}
+\normalsize
+\end{table}%
+
 \begin{table}[!ht]
@@
 \begin{tabular}{l|c|c|c}
-Dataset			&		II (7 Gb )	& III (30 Gb)	& IV (100 Gb)\\
+Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
 \hline
@@
 \hline
-	& mapping 	&	SNP calling	&	total  &total & map& SNP call\\
+	& alignment 	&	SNP calling	&	total  &total & map& SNP call\\
 \hline
@@
 
-data, Gb		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
+data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
 				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
@@
 \begin{tabular}{|l|l|l|}
-dataset &	organism &	size in Gb\\
+dataset &	organism &	size in Gbases\\
 \hline
@@
 \begin{table}[!ht]
-\caption{Timings  for mapping and the ratio~$T_{mapping}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
+\caption{Timings  for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
 For the ``HPC random'' approach, data chunks have to be copied to the local scratch disks first and the alignments (SAM files) copied back while Hadoop keeps all the data inside HDFS and hence does not need data staging.
@@
 
-Number of nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$	&Number of nodes	&Mapping time,	&$\frac{T_{mapping}}{T_{comm}}$\\
+Number of nodes	&Mapping time,	&$\frac{T_{alignment}}{T_{comm}}$	&Number of nodes	&Mapping time,	&$\frac{T_{alignment}}{T_{comm}}$\\
 (cores)					&minutes		&							&(cores)			&minutes&\\

### 2014-02-25 paper-plos.tex
More updates to comments and changes to american english.

@@
 %Current practices: How to parallelize
-The most common approach to speed up NGS tools is to parallelise within a compute node (multi-core parallelisation) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16, see~\cite{top500}.
+The most common approach to speed up NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16, see~\cite{top500}.
 For the tools which do not support OMP natively, e.g.
-Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel Linux utility~\cite{Tange2011a}.
-Of great importance is that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and multi-node parallelisation generally are not bound by these constraints.
+Samtools, variant calling can be parallelized by creating a separate process  for  each chromosome, or using GNU Parallel Linux utility~\cite{Tange2011a}.
+Of great importance is that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and multi-node parallelization generally are not bound by these constraints.
 
-Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronisation (or load balancing) has to be woven into the software code and there are few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
+Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelization, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code and there are few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
-Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting.
+Another common way to introduce parallelization to NGS analysis pipelines in Linux systems is to use Bash scripting.
 This involves using existing utilities and cluster tools to split the data into chunks, process them on the separate nodes, and merge the results afterwards.
-This kind of solution benefits from both MPI-like and OMP parallelisation and provides good performance, but the development requires substantial expertise in order to be efficient.
+This kind of solution benefits from both MPI-like and OMP parallelization and provides good performance, but the development requires substantial expertise in order to be efficient.
 Since the process is tightly coupled to the local computational cluster and network architectures, it might also not be possible to be re-used in other settings.
@@
 A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet~\cite{Langmead:2009kx}.
-The only way to introduce parallelisation in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
+The only way to introduce parallelization in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
 This, however, is lacking the massively parallel way of treating the data.
@@
 
-Due to the trivially parallelisability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
+Due to the trivially parallelizability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
 Reducing the size of each data chunk reduces the aligner  job,~$T_{alignment}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
@@
 For Hadoop, the short reads in FASTQ format have to be preprocessed (involving node communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data ingestion into the HDFS.
-We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
+We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelization.
 For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual alignment happens, and the sending of the aligned SAM files back to the delivery location over the network.
@@
 and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
-The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbasesit/sec.
-Ten nodes will consume the data with the rate of~10Gbasesit/sec, which is the limiting speed for the standard 10Gbasesit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated on the same rack, which is the default SLURM behaviour.
+The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
+Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
+The nodes are being allocated on the same rack, which is the default SLURM behavior.
 
-Scalability can be improved by overriding the default behaviour of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
+Scalability can be improved by overriding the default behavior of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
 Allocating the nodes on random racks allows one to engage more nodes without network saturation.
@@
 $5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}.
-The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbasesit/sec being saturated.
+The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
 The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
 HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
-Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behaviour is a generic for Hadoop with HDFS\cite{lin2010, Hadoop:Guide}.
+Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behavior is a generic for Hadoop with HDFS\cite{lin2010, Hadoop:Guide}.
 
@@
 Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behaviour.
+In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behavior.
 This is however not of practical use in an HPC center since the network bandwidth is shared among multiple users.
-Finally, we demonstrate that an existing and extensible publically available web-based GUI (Cloudgene) provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
+Finally, we demonstrate that an existing and extensible publicly available web-based GUI (Cloudgene) provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
 
@@
 The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE.
+This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 “Next Generation Sequencing Data Analysis Network”, SeqAhead.
 We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
@@
 HPC:
-Multinode short read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbasesit/s uplink.
+Multinode short read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
 
@@
 In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
-It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbasesit per second.
+It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
 
@@
 \texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
-
-\COMMENT{Journal abbreviations in references are inconsistant.
-The journal name for ref 2 is wrong.
-Capitalization of refs 4, 14, 16, 18, 19, 25, 27, etc are wrong.
-All refs should be checked for accuracy and style consistancy.
-Many or all footnotes should be changed to endnotes (references).}
 
@@
 	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases. Calculations were carried out for 56 cores Hadoop and 16 cores HPC cluster correspondingly.
+	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gbases. Calculations were carried out for 56 cores Hadoop and 16 cores HPC cluster correspondingly.
 The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are labeled with roman numerals.}
@@
 	\caption{Ratios of the alignment time~$T_{alignment}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset~IV as a function of reciprocal number of nodes~$1/N$.
-Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behaviour and a modified one where nodes are being allocated from random racks.
+Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behavior and a modified one where nodes are being allocated from random racks.
 Linear fit  to $ax+b$ was done with the least-squares method.
@@
 \small
-\caption{Timings for Dataset I processed on different number of cores of a HPC node. \COMMENT{awkward title, consider replacing with graph (and new title)} \COM{better?}}
+\caption{Timings for Dataset I processed on different number of cores of a HPC node.}
 \begin{center}
@@
 \caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.
-Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.
-Big deviation \COMMENT{what aspect of deviation, not fixable?} for the HPC is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.
-\COMMENT{No p= used here.} }
+Dataset are labelled by roman numerals, ``f.r.'' stands for ``forward reads''.
+The large variance for the HPC deployment is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.}
 \begin{center}

### 2014-02-27 paper-plos.tex
minor text clarifications and ref. addition

@@
 We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual alignment with Bowtie.
-If lacking RAM, Samtools \sout{implementation} swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
+If lacking RAM, Samtools  swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
 Table~\ref{table:5} shows the sorting runtime performance of Samtools for the dataset~IV run with different amount of RAM available on a single HPC node.
@@
 However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
-To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo}, and compared the network performance with the standard Hadoop HDFS approach.
+To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo_bash}, and compared the network performance with the standard Hadoop HDFS approach.
 We separated the alignment time, $T_{alignment}$, and the communication time, $T_{comm}$, and plotted $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
@@
 
-In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools.
+In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the aligned reads and~SNP calling with Samtools.
 The Bowtie aligner natively implements~OMP, meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core.
@@
 Where available these features were used to improve the analysis speed.
-The exact workflow used is available from github~\cite{code_repo}.
+The exact workflow used is available from github~\cite{code_repo_bash}.
 
@@
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from github~\cite{code_repo}).
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from github~\cite{code_repo_mr}).
 
@@
 The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 “Next Generation Sequencing Data Analysis Network”, SeqAhead.
+This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 ÒNext Generation Sequencing Data Analysis NetworkÓ, SeqAhead.
 We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
@@
 Each node in this deployment was equipped with dual 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
-The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.4.0~\cite{cloudera}.
+The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.5.0~\cite{cloudera}.
 Note that the physical hardware provided less RAM than desired.
@@
 
-\subsection*{Bash script for the preprocessing stage}
+\subsection*{Crossbow preprocessing stage as a Bash script}
+Current Bash script mimics functionality of the Crossbow's preprocessing stage, and utilizes the multicore parallelism.
+The script reformats the BZIP2 archived short reads files in FASTQ format to a single text file, each line containing following tab-separated fields:
+\begin{itemize}
+\item read header
+\item forward read
+\item forward read qualities
+\item reverse read
+\item reverse read qualities
+\end{itemize}
 
@@
 \hline
-	& alignment 	&	SNP calling	&	total  &total & map& SNP call\\
+	& alignment 	&	SNP calling	&	total  &total & alignment& SNP call\\
 \hline
@@
 
-Number of nodes	&Mapping time,	&$\frac{T_{alignment}}{T_{comm}}$	&Number of nodes	&Mapping time,	&$\frac{T_{alignment}}{T_{comm}}$\\
+Number of nodes	&Alignment time,	&$\frac{T_{alignment}}{T_{comm}}$	&Number of nodes	&Alignment time,	&$\frac{T_{alignment}}{T_{comm}}$\\
 (cores)					&minutes		&							&(cores)			&minutes&\\

### 2014-02-27 paper-plos.tex
Edits up to the end of the subsection "Modified preprocessing stage"

@@
 \\
-\bf{1} Department of Information Technology, Uppsala University, Uppsala, Sweden
+{\bf 1} Department of Information Technology, Uppsala University, Uppsala, Sweden
  \\
-\bf{2} CRS4, Polaris, Pula, Italy
+{\bf 2} CRS4, Polaris, Pula, Italy
  \\
-\bf{3} Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden
+{\bf 3} Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden
  \\
-\bf{4} Science for Life Laboratory, Uppsala University, Uppsala, Sweden
+{\bf 4} Science for Life Laboratory, Uppsala University, Uppsala, Sweden
 \\
@@
 \section*{Abstract}
-New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field.
-With increasing data volumes comes the necessity to analyzee data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
-Hadoop, emerging in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
+New high-throughput technologies such as massively parallel sequencing have transformed the life sciences into a data-intensive field.
+With increasing data volumes comes the necessity to analyze data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
+Hadoop, which has emerged over the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
 Herein we introduce metrics and report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
@@
 \section*{Author Summary}
-New experimental technologies generate vast amounts of data and require supercomputers to store and analyze.
-As datasets increase in volume there is a need for new computational tools and technologies that can cope with the scale of data, and in this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
-While there is an overhead of using Hadoop, we show that when sequencing data sets reach sizes of 100 gigabases then Hadoop becomes attractive compared with traditional methods.
+New experimental technologies in biology generate vast amounts of data that require
+high-performance computing
+to analyze and store them.
+Indeed, with the increase in dataset volumes there is a need for new computational tools and technologies that can cope with the scale of data.
+In this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
+While there is an overhead in using Hadoop, we show that as sequencing data sets reach sizes of 100 gigabases Hadoop becomes an attractive option compared with traditional methods.
 We also show that Hadoop scales linearly while traditional systems become network-saturated at larger data sizes.
-As sequencing data are bound to increase in the future, Hadoop is a very interesting framework to base future analysis on.
+As sequencing data are bound to increase in the future, Hadoop is a very interesting framework on which to base future analysis.
 
@@
 %The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
-Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline that requires a high-performance computing (HPC) infrastructure to analyze.
+Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data providing insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
+Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes.
+Extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline requiring a high-performance computing (HPC) infrastructure to produce results.
 As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gbases) corresponding to 2\,terabytes (TB) of raw data in 3 days~\cite{illumina}.
@@
 %Current practices: HPC systems
-A common step of NGS data analysis consists of alignment short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Many of the most widespread tools for this purpose, like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools}, are not made with distributed computing in mind.
+A common step of NGS data analysis consists of aligning short reads to a reference sequence and then finding the genetic variations specific to the sample.
+Many of the most widespread tools for this purpose -- like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools} -- are not made with distributed computing in mind.
 Many others do not even have the native ability to use multiple cores.
@@
 %Current practices: How to parallelize
-The most common approach to speed up NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16, see~\cite{top500}.
+The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16 (see~\cite{top500}).
 For the tools which do not support OMP natively, e.g.
-Samtools, variant calling can be parallelized by creating a separate process  for  each chromosome, or using GNU Parallel Linux utility~\cite{Tange2011a}.
-Of great importance is that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and multi-node parallelization generally are not bound by these constraints.
+Samtools, variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
+It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.
 
-Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelization, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code and there are few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
+The Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelization.
+However, writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code and there are few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
-Another common way to introduce parallelization to NGS analysis pipelines in Linux systems is to use Bash scripting.
+Another common way to introduce parallelization into NGS analysis pipelines in Linux systems is to use Bash scripting.
 This involves using existing utilities and cluster tools to split the data into chunks, process them on the separate nodes, and merge the results afterwards.
 This kind of solution benefits from both MPI-like and OMP parallelization and provides good performance, but the development requires substantial expertise in order to be efficient.
-Since the process is tightly coupled to the local computational cluster and network architectures, it might also not be possible to be re-used in other settings.
+Since the process is tightly coupled to the local computational cluster and network architectures, it might also nearly impossible to re-use such developments in other settings.
 
@@
 %Properties of Hadoop and HDFS- data localization and programming model etc
-The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running tasks in a {\it massively} parallel way.
+The Map-Reduce~(MR) programming paradigm~\cite{dean.2004.mapreduce} offers a compelling alternative for running tasks in a {\it massively} parallel way.
 This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
 The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}.
-The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
-Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively moving the computation to the data and reducing network traffic.
+The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies of data on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
+Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively enabling moving the computation to the data and thus reducing network traffic.
 HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
@@
 %Introduce what we did in the manuscript.
-In this manuscript we focus on the question {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. 
-Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective.
-We also study execution times for datasets of different sizes and analyze Hadoop and HPC approaches from a scaling perspective.
+In this manuscript we focus on the question of {\it when} is Hadoop an appealing alternative to the programs for DNA-seq analysis generally found in~HPC centers.
+Since the Hadoop framework poses a significant performance overhead, we seek to estimate the average data size at which point, from a performance perspective, using it starts to pay off.
+We also study the analysis execution times for datasets of different sizes and evaluate Hadoop and the HPC approaches from a scaling perspective.
 
@@
 
-We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as HPC method) (Methods).
-Our experiments then consisted of running the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage. 
+We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as the HPC method) (Methods).
+Our experiments consisted of executing the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage.
 All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
-For the existing Hadoop software we propose modifications, in order to benefit fully from massively parallel nature of MR computations.
-For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short read alignment and exploiting the network fully, thereby speeding up calculations.
+Finally, we propose modifications for existing Hadoop software to better benefit fully from massively parallel nature of MR computations; we also
+developed a set of Bash scripts for the classical HPC DNA-seq analysis programs for short read alignment to use multiple nodes and fully exploit the network, thereby speeding up calculations.
 
@@
 A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet~\cite{Langmead:2009kx}.
-The only way to introduce parallelization in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks.
+However, the only way to introduce parallelization in the native preprocessing stage is to split the read files into smaller chunks and to explicitly configure Crossbow's preprocessor to download each chunk (this is done by listing them in a special \textit{manifest file}.
+listing all these chunks.
 This, however, is lacking the massively parallel way of treating the data.
@@
 
-Assuming that sequencing platforms are usually in close proximity with facilities that provide both storage and computers, we considered that the data already had been downloaded to storage that Hadoop could access.
-We rewrote the Crossbow preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
-The requirement is that the FASTQ data are i.e.
-BZIP2 archived (however, any splittable archive format is suitable) and accessible by Hadoop.
-For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were transferred to the HDFS.
-BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner.
-Also, BZIP2 format is natively supported by Hadoop, i.e., there is no difference in dealing with the FASTQ data or with its BZIP2 archive.
+Assuming that sequencing platforms are in close proximity of storage and
+computing facilities, we considered that the data was already stored where
+Hadoop could access it.
+We rewrote the Crossbow preprocessing stage scripts as MR programs, thus
+benefiting from its massively parallel nature.  For our tests, we required that
+the input FASTQ data be compressed with \texttt{bzip2}.
+\texttt{bzip2} provides very good compression and is splittable,
+While this is not an
+intrinsic constraint in our set up, we reasoned that it is typical to store
+sequencing data in a compressed file format; \texttt{bzip2} provides very good
+compression and its file format is splittable,
+meaning that the archive can be expanded in a parallel manner allowing Hadoop to process multiple segments simultaneously.
+In addition, the \texttt{bzip2} format is natively supported by Hadoop -- i.e. FASTQ data can be read whether it is uncompressed or \texttt{bzip2}-compressed.
+
+In our tests, the storage volume containing the input data was mounted with SSHFS to one of the Hadoop nodes, from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
 
@@
 
-The Hadoop streaming library offers a possibility to write code for Hadoop jobs in any programming language.
+The Hadoop streaming library offers the flexibility to write Hadoop programs in any programming language.
 Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive~\cite{ncbi-sra}).
-The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
+% LP: I think this sentence could be removed since to appreciate what is being
+% described one would have to understand quite well how that program works.
+The various formats that our script reads lack a standard FASTQ record id, which needs to be interpreted properly at least to extract the correct read number.
+To circumvent this problem we rewrote the id based on a SHA-1 hash function and labelling reads mates ``.1" and ``.2" suffixes.
 To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.
 
-In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays, e.g., while downloading data from Amazon.
-Table~\ref{table:preprocess} illustrates the benefits of our approach.
+In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays -- e.g., while downloading data from Amazon.
+Table~\ref{table:preprocess} illustrates the benefits of our MR approach.
 To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
@@
 The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 ÒNext Generation Sequencing Data Analysis NetworkÓ, SeqAhead.
+This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 Next Generation Sequencing Data Analysis Network, SeqAhead.
 We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
+Part of L.P.'s activity was performed within the context of the Ph.D.\ program in
+Biomedical Engineering at the University of Cagliari, Italy.
 

### 2014-02-28 paper-plos.tex
Fixme and Change macros
Small progress

@@
 %\doublespacing
+
+\definecolor{ChangeColor}{rgb}{0,1,0}
+\newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}
+
+\definecolor{FixmeColor}{rgb}{0,1,0}
+\newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}
 
@@
 \subsection*{Accuracy of pipelines}
-Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp, correspondingly) we should not expect them to deliver perfectly matching SNP lists, but still we expect them to capture and correctly identify the mutations.
-We tested the correctness just for the smallest dataset~(I).
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.
+Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and SOAPsnp~\cite{soapsnp}, respectively) we do not expect them to deliver perfectly matching SNP lists, but we do expect them to capture and correctly identify the
+% which mutation?
+\Fixme{mutations}.
+We only tested for correctness with the smallest dataset~(I).
+The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$~\cite{schneeberger} was successfully localized by both variant callers.
 
 \subsection*{Scalability of HPC and Hadoop approaches}
-To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). 
-The alignment process in Bowtie scales fairly well, but the SNP calling part in Samtools is a bottleneck, scaling worse than Bowtie, and consuming a progressively larger portion of the total calculation time even after parallelizing analysis by chromosome as previously suggested~\cite{biostars_samtools}.
+\Change{To demonstrate the scalability of the HPC approach we ran the HPC pipeline using dataset~I as input.
+We collected the running times} as a function of the number of cores used (Table~\ref{table:3}).
+We found that the alignment process with Bowtie scales fairly well, while SNP calling with Samtools is the bottleneck of the workflow.
+Indeed, this latter part scales worse than Bowtie and consumes a progressively larger portion of the total calculation time as the input grows, even after parallelizing analysis by chromosome as previously suggested~\cite{biostars_samtools}.
 %\footnote{\url{http://www.biostars.org/p/48781}
@@
 For our tasks we used the default Samtool's settings.}.
-For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for the HPC and Hadoop approaches were collected (Table~\ref{table:4}).
+\Change{For both the HPC and Hadoop approaches, and using all the test datasets (I-V), we collected the time required for the combined task of mapping the reads to the corresponding reference genome and SNP calling.
+The results are shown in Table~\ref{table:4}.}
 
-One of the attractive sides to run i.e.
-Samtools, Bowtie, etc.
-in the  Hadoop~MR framework is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{Langmead:2009kx,Pireddu:2011vn}, regardless the scaling nature of the underlying program.
+\Change{The results confirm that one the advantages to using the Hadoop~MR} framework is its almost linear scalability -- i.e.\ the calculation time linearly depends on the size of the dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
+\Fixme{Deleting this since one could dream/find pathological cases that don't work: regardless the scaling nature of the underlying program.}
 Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
@@
 \subitem Short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: SOAPsnp~1.02~\cite{soapsnp}
+\subitem SNP calling: SOAPsnp~1.02
 \end{enumerate}

### 2014-03-03 paper-plos.tex
Slightly more progress

@@
 
-\definecolor{ChangeColor}{rgb}{0,1,0}
+\definecolor{ChangeColor}{rgb}{0.24,0.56,0.18}
 \newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}
 
-\definecolor{FixmeColor}{rgb}{0,1,0}
+\definecolor{FixmeColor}{rgb}{1,0,0}
 \newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}
@@
 \Change{The results confirm that one the advantages to using the Hadoop~MR} framework is its almost linear scalability -- i.e.\ the calculation time linearly depends on the size of the dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
-\Fixme{Deleting this since one could dream/find pathological cases that don't work: regardless the scaling nature of the underlying program.}
-Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56 cores Hadoop cluster and datasets I-IV.
+\Fixme{Deleting this since one could dream/find pathological cases that don't work: ``regardless the scaling nature of the underlying program''.}
+Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56-core Hadoop cluster and datasets I-IV.
 %To stress the linear nature of scaling, both sets of points were fit using the least squares method.
@@
 \subsection*{Comparing Hadoop and HPC runtime efficiency}
-Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the function showing the amount of used core-hours:
+\Change{%
+Hadoop was designed to process huge datasets\cite{hadoop,lin2010}.
+Therefore, one could image then that the larger the dataset the more advantageous Hadoop should become compared to the HPC approach.
+In order to compare the efficiency of the two processing platforms (Hadoop and HPC) as the size of the dataset varied, we looked at the number of core-hours used for same analysis:
 $F=T_{p}\times p,$ 
 where~$T_{p}$ is the calculation time on~$p$ cores. 
+}
 
-Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
+\Change{%
+Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
 Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
-As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+It is evident from the data that the Hadoop pipeline becomes increasingly more efficient than the HPC-based pipeline as the dataset size increases.
+The parabolic extrapolation of our data yields the value~$1.70\pm0.01$ for the ratio, meaning that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+}
 
-The simplest curve to fit the points nicely is a parabolic function, importantly not a constant. 
-Since the Hadoop approach for the datasets reveals the linear scaling (Figure~\ref{fig:fig1}), a deviation of the ratio from the constant shows the 
+\Change{%
+The simplest curve that fits the data points nicely is a parabolic function which, importantly, is not a constant. 
+Since our measurements using the Hadoop-based pipeline, summmarized in Figure~\ref{fig:fig1}, showed linear scaling properties of Hadoop in this setting, a deviation of the ratio $F_{Hadoop}/F_{HPC}$ from the constant is evidence of the
 lack of scalability of the HPC approach.
+
 %Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
@@
 \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
-We suspect the main reason for such a behavior is the fact that for large datasets (dataset~III-V in our case) the SNP calling routine with Samtools becomes more memory and time demanding than the actual alignment with Bowtie.
-If lacking RAM, Samtools  swaps to disk, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and time-expensive IO on a high rate.
-Table~\ref{table:5} shows the sorting runtime performance of Samtools for the dataset~IV run with different amount of RAM available on a single HPC node.
-The large uncertainty for small amounts of RAM we believe is due to other users' load on the cluster network.
-As a result, a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling on much more economic hardware in linear time with just 12GB RAM per virtual node.
+We suspect the main reason for such a behavior is that for large datasets (dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory- and time-demanding than the alignment with Bowtie.
+Since its memory requirements scale with input size, with sufficient input data Samtools swaps data to disk, creating multiple (up to hundreds) temporary files while it sorts the BAM file -- incidentally, these files can end up on shared network-accessed file systems and thus generate large amounts of network traffic.
+Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with different amount of RAM available on a single HPC node.
+Though there is a large deviation when using small amounts of RAM, we believe due to other users' load on the cluster network, the trend quite clearly indicates that a relatively large amount of RAM per node is needed in order to use Samtools effectively.
+In contrast, in our tests Hadoop performed the SNP calling on much more economic hardware in linear time with just 12GB of RAM per \textit{virtual} node.
+}
 

### 2014-03-12 paper-plos.tex
Extensive changes all over

@@
 
+% required by the coloured \Fixme and \Change macros
 \usepackage{color}
@@
 \definecolor{ChangeColor}{rgb}{0.24,0.56,0.18}
+%\definecolor{ChangeColor}{rgb}{0,0,0}
 \newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}
@@
 \definecolor{FixmeColor}{rgb}{1,0,0}
+%\definecolor{FixmeColor}{rgb}{0,0,0}
 \newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}
@@
 The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16 (see~\cite{top500}).
-For the tools which do not support OMP natively, e.g.
-Samtools, variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
+For the tools which do not support OMP natively -- e.g.
+Samtools -- variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
 It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.
@@
 
+%%%%%%%%%%%
+% Methods
+%%%%%%%%%%%
+% You may title this section "Methods" or "Models". 
+% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
+% authors may use "Analysis" 
+\section*{Methods}
+
+\Fixme{move this section back to the its original position if you don't like it
+here (it was just before Acknowledgements)}
+
+\subsection*{Analysis pipelines}
+
+For our work to determine when Hadoop is an appealing option for genomic
+sequence analysis, we chose as a representative 
+use case for our experiments the task identifying
+single-nucleotide polymorphisms (SNPs) from short-read data.  We constructed
+two pipelines that perform the same task: one that runs in an HPC environment
+and the other on the Hadoop platform.  Although there are many different
+approaches and software tools to perform this task in an HPC environment (e.g.,
+GATK~\cite{gatk}, \Fixme{Other example? Otherwise none}), there is only a limited
+number of options for Hadoop.  Given this limitation, we decided to keep the
+analysis pipeline as simple as possible to ensure a fair comparison between the
+two platforms.
+
+\begin{enumerate}
+\item HPC approach
+\subitem Short-read alignment: Bowtie~ver.~0.12.8
+\subitem SNP calling: Samtools~ver.~0.1.19
+\item Hadoop approach: Crossbow
+\subitem Short-read alignment: Bowtie~ver.~0.12.8
+\subitem SNP calling: SOAPsnp~1.02
+\end{enumerate}
+
+In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the aligned reads and~SNP calling with Samtools.
+The Bowtie aligner natively implements~OMP -- therefore, using 8 cores on the same computer will theoretically produce a result 8 times faster than using a single core.
+Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
+Where available these features were used to improve the analysis speed.
+The exact workflow used is available from GitHub~\cite{code_repo_bash}.
+
+\Change{%
+The equivalent Hadoop-based pipeline was implemented with Crossbow. The input
+data and the indexed genome reference were copied to Hadoop's storage
+system~(HDFS) before starting the experiments.  Specifically, the storage volume
+containing the input data was mounted with SSHFS to one of the Hadoop nodes,
+from where the \texttt{bzip2}-compressed data were transferred to the HDFS\@.
+This preparation was done to eliminate unpredictable effects on the experiments
+from possible network delays.
+}
+Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
+Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program).
+Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from GitHub~\cite{code_repo_mr}).
+Our implementation is described in Section~\ref{sec:preprocessor}.
+
+\subsection*{Datasets}
+For our experiments we used three publicly available DNA-seq datasets~(I--III),
+a synthetic dataset~(IV) of {\it A.thaliana} -- the well-known model plant --
+and a dataset (V) of two {\it H.sapiens} individuals
+(Table~\ref{table:datasets}). Datasets I-III and V are composed of Illumina
+HiSeq sequencing data. Additional information about the source of the datasets is
+provided in the Supplementary Material section.
+
+\Change{%
+All data was standardized to FASTQ format and compressed with \texttt{bzip2},
+since it is typical to store sequencing data in a compressed file format. The
+\texttt{bzip2} compression codec was selected it provides very good compression
+and is splittable by Hadoop, meaning that the archive can be expanded in a
+parallel manner allowing Hadoop to process multiple segments simultaneously.
+Also, since some of the datasets lack a standard FASTQ record id, which is
+required by our test pipelines to extract the read number, we also standardised
+the read ids in the datasets by generating new ids based on the SHA-1 hash
+function and labelled read mates with ``.1" and ``.2" suffixes.
+}
+
+\subsection*{MapReduce-based Data Preprocessor}
+\label{sec:preprocessor}
+
+\Change{%
+The Crossbow package includes a native preprocessing stage that provides great
+flexibility in delivering the data to the Hadoop cluster; it can autonomously
+download data from Amazon S3, FTP and HTTP servers over the
+Internet~\cite{Langmead:2009kx}. However, this preprocessor does not parallelize
+its operations. In theory one can manually introduce some degree of
+parallelization by manually splitting the original input read data into smaller
+files and to explicitly configuring Crossbow's preprocessor to download each one
+(this is done by listing all of them in a special \textit{manifest file}).
+Nevertheless, in practice this approach is rather error-prone and laborious.
+
+%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
+We reimplemented the main functionality in the Crossbow preprocessor as an MR
+program to benefit from its massively parallel nature. The new MR preprocessor
+was implemented in Python using the Hadoop streaming library.
+\Fixme{Consider deleting the following implementation detail.}
+To route both reads of the same pair to the same Reducer, a secondary
+key-sorting mechanism was employed.
+The resulting program efficiently processes short reads produced by different
+versions of the Illumina sequencing platform as well as FASTQ files generated
+from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
+Hadoop-based approach imposes the requirement that the input data be stored
+where the Hadoop nodes can access it, which is in our opinion a reasonable since
+sequencing platforms generally have direct access to extensive storage and
+computing facilities.
+}
+
+\subsection*{Computational resources}
+To run the HPC analysis pipeline we used a computational cluster at UPPMAX,
+which is heavily used for analysis of NGS-data~\cite{lampa}.
+Data and reference genomes were stored on a parallel, shared storage system.
+On the other hand, the Hadoop test platform was deployed on a private cloud at
+UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was
+configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}.
+For the fully detailed hardware configuration of the two computational resources
+please see the Supplementary Material section.
+
 %%%%%%%%%%%%
@@
 
-We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as the HPC method) (Methods).
-Our experiments consisted of executing the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage.
-All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
-Finally, we propose modifications for existing Hadoop software to better benefit fully from massively parallel nature of MR computations; we also
-developed a set of Bash scripts for the classical HPC DNA-seq analysis programs for short read alignment to use multiple nodes and fully exploit the network, thereby speeding up calculations.
+\Change{%
+As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
+(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
+with a batch processing system (hereafter referred to as the HPC method)
+(Methods).  Our experiments consisted of executing the pipelines for each of the
+five input 
+datasets (see Table~\ref{table:datasets}) and
+measuring the wall-clock run time for each pipeline stage.  All experiments were
+repeated
+\Fixme{several times (how many?)}
+and the results averaged to obtain a data point for the
+particular combination of data size and computational platform.
+}
 
-\subsection*{Modified preprocessing stage}
-A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet~\cite{Langmead:2009kx}.
-However, the only way to introduce parallelization in the native preprocessing stage is to split the read files into smaller chunks and to explicitly configure Crossbow's preprocessor to download each chunk (this is done by listing them in a special \textit{manifest file}.
-listing all these chunks.
-This, however, is lacking the massively parallel way of treating the data.
+\Fixme{Belongs with Discussion:
+In light of our observations, we
+propose modifications for existing Hadoop software to better exploit the
+massively parallel nature of MR computations. Finally, we developed a set of Bash
+scripts to parallelize the operations of classical DNA-seq analysis programs for
+short-read alignment over multiple nodes and fully exploit the network bandwidth, thereby
+accelerating calculations.
+}
 
-%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
+\subsection*{Accuracy of Pipelines}
+\Change{%
+Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
+SOAPsnp~\cite{soapsnp}, respectively) we could not expect them to deliver perfectly
+matching SNP lists. Instead, we verified the accuracy of the test pipelines by
+verifying their ability to correctly identify the mutation $C\rightarrow T$ on
+chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.
+}
 
-Assuming that sequencing platforms are in close proximity of storage and
-computing facilities, we considered that the data was already stored where
-Hadoop could access it.
-We rewrote the Crossbow preprocessing stage scripts as MR programs, thus
-benefiting from its massively parallel nature.  For our tests, we required that
-the input FASTQ data be compressed with \texttt{bzip2}.
-\texttt{bzip2} provides very good compression and is splittable,
-While this is not an
-intrinsic constraint in our set up, we reasoned that it is typical to store
-sequencing data in a compressed file format; \texttt{bzip2} provides very good
-compression and its file format is splittable,
-meaning that the archive can be expanded in a parallel manner allowing Hadoop to process multiple segments simultaneously.
-In addition, the \texttt{bzip2} format is natively supported by Hadoop -- i.e. FASTQ data can be read whether it is uncompressed or \texttt{bzip2}-compressed.
+\subsection*{Performance evaluation of MR-based preprocessing stage}
 
-In our tests, the storage volume containing the input data was mounted with SSHFS to one of the Hadoop nodes, from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
+\Change{%
+To quantify the processing time saved by adopting the MR approach for
+preprocessing the read data, we ran an
+experiment comparing the new preprocessor described in
+Section~\ref{sec:preprocessor}  to the original preprocessor furnished with
+Crossbow.  To further validate our strategy, we also compared the performance of
+our MapReduce approach to a different parallelization strategy based on a custom Bash script
+that uses of all processing cores available on a single computing node to accelerate the operation (see
+Supplementary Material for details of the implementation). This program read its
+input directly from conventional HPC storage, rather than HDFS\@.  The experiment
+was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
+with 56 cores (7 slave nodes) while the Bash approach ran on a single node
+equipped with 16 CPU cores.
+}
 
@@
 
-The Hadoop streaming library offers the flexibility to write Hadoop programs in any programming language.
-Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive~\cite{ncbi-sra}).
-% LP: I think this sentence could be removed since to appreciate what is being
-% described one would have to understand quite well how that program works.
-The various formats that our script reads lack a standard FASTQ record id, which needs to be interpreted properly at least to extract the correct read number.
-To circumvent this problem we rewrote the id based on a SHA-1 hash function and labelling reads mates ``.1" and ``.2" suffixes.
-To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.
+\Fixme{Alexey, please verify whether my interpretation seems reasonable. LP}
 
-In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays -- e.g., while downloading data from Amazon.
-Table~\ref{table:preprocess} illustrates the benefits of our MR approach.
-To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage.
-The script utilizes multiple cores and was executed on a single HPC node (see Supplementary information).
-%ran on a single HPC node, utilizes multiple cores, with the limiting factor being \sout{the hard disk IO performance.}\COM{More correctly to say, that the pipe is not filled completely.
-i.e.
-the degree of achieved parallelization does not utilize all 16 cores.
-Perhaps not the best bash commands etc.}
-%, which effectively leaves with 5-7 cores of 16. \COMMENT{awkward}
+\Change{%
+Table~\ref{table:preprocess} shows the results of the experiment. Clearly,
+either of the parallel approaches are much faster than the original tool, with
+all dataset sizes.  Also, we noticed that the difference between the Bash and Hadoop
+approaches is minimal when running on Dataset II\@.  Indeed, with the relatively
+small dataset size the overhead incurred by the Hadoop framework likely
+overwhelms the useful compute time used for each available CPU core. However,
+with increasing input size the superior scalability of MR becomes apparent, with
+the largest dataset taking requiring less than half the time of the Bash
+approach to be processed; at the same time, it is important to also notice that
+the Hadoop approach employed approx.\ 1.6 times more CPU time compared to the
+Bash approach.
+}
 
-\subsection*{Accuracy of pipelines}
-Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and SOAPsnp~\cite{soapsnp}, respectively) we do not expect them to deliver perfectly matching SNP lists, but we do expect them to capture and correctly identify the
-% which mutation?
-\Fixme{mutations}.
-We only tested for correctness with the smallest dataset~(I).
-The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$~\cite{schneeberger} was successfully localized by both variant callers.
+%ran on a single HPC node, utilizes multiple cores, with the limiting factor
+%being \sout{the hard disk IO performance.}\COM{More correctly to say, that the
+%pipe is not filled completely. i.e. the degree of achieved parallelization does
+%not utilize all 16 cores. Perhaps not the best bash commands etc.} , which
+%effectively leaves with 5-7 cores of 16. \COMMENT{awkward}
 
 \subsection*{Scalability of HPC and Hadoop approaches}
-\Change{To demonstrate the scalability of the HPC approach we ran the HPC pipeline using dataset~I as input.
-We collected the running times} as a function of the number of cores used (Table~\ref{table:3}).
-We found that the alignment process with Bowtie scales fairly well, while SNP calling with Samtools is the bottleneck of the workflow.
-Indeed, this latter part scales worse than Bowtie and consumes a progressively larger portion of the total calculation time as the input grows, even after parallelizing analysis by chromosome as previously suggested~\cite{biostars_samtools}.
-%\footnote{\url{http://www.biostars.org/p/48781}
-%\COM{here was a footnote referring to biostars site. Ok, if the footnotes are not welcome, lets then have it as a reference. That code is not trivial. Maybe we can leave the ref?}
-%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance.
-During its sorting phase the lack of RAM results in spilling to the HDD.
-For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM.
-For our tasks we used the default Samtool's settings.}.
-\Change{For both the HPC and Hadoop approaches, and using all the test datasets (I-V), we collected the time required for the combined task of mapping the reads to the corresponding reference genome and SNP calling.
-The results are shown in Table~\ref{table:4}.}
+\Change{%
+Given the wide range in genome dataset sizes processed in practice -- e.g.,
+exomes,targeted sequencing, whole-genome sequencing at various depths -- it is
+important to understand the scaling characteristics of each processing
+platforms (Hadoop and HPC) with respect to input size. For each platform, we
+measured the time required for to execute their respective variant calling
+pipeline (see Methods) for each of the test datasets (I-V). The results are
+shown in Table~\ref{table:4}.
 
-\Change{The results confirm that one the advantages to using the Hadoop~MR} framework is its almost linear scalability -- i.e.\ the calculation time linearly depends on the size of the dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
-\Fixme{Deleting this since one could dream/find pathological cases that don't work: ``regardless the scaling nature of the underlying program''.}
-Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56-core Hadoop cluster and datasets I-IV.
-%To stress the linear nature of scaling, both sets of points were fit using the least squares method.
+The measurements highlight that one of the advantages of using the Hadoop~MR} framework is
+its almost linear scalability with respect to input size -- i.e.\ the
+calculation time linearly depends on the size of the
+dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
+%Deleting this since one could dream/find pathological cases that don't work: ``
+\Fixme{\sout{regardless the scaling nature of the underlying program.}}
+Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the wall clock
+running time as a function of the dataset size for the 56-core Hadoop cluster and datasets
+I-IV.
+%To stress the linear nature of scaling, both sets of points were fit using the
+%least squares method.
 
-\subsection*{Comparing Hadoop and HPC runtime efficiency}
 \Change{%
-Hadoop was designed to process huge datasets\cite{hadoop,lin2010}.
-Therefore, one could image then that the larger the dataset the more advantageous Hadoop should become compared to the HPC approach.
-In order to compare the efficiency of the two processing platforms (Hadoop and HPC) as the size of the dataset varied, we looked at the number of core-hours used for same analysis:
-$F=T_{p}\times p,$ 
-where~$T_{p}$ is the calculation time on~$p$ cores. 
+Analogously, we investigated the scalability characteristics of the HPC method
+with respect to input size.
+\Fixme{The following works better the aid of a plot.}
+Analyzing the observations in Table~\ref{table:4}, we can see with the HPC
+method run time increases superlinearly as a function of dataset size.  As
+expected, the result is that the HPC method is not able to complete the analysis
+with our larger tests datasets.  On the other hand, this approach is not much
+slower than our Hadoop cluster considering the much smaller quantity of computing
+resources that is uses -- an issue that is analyzed more closely in the section
+on the efficiency of the two systems.
+}
+
+\Fixme{The following paragraph comments on how the HPC approaches' runtime
+varies with the number of cores available, rather than input size. If we're
+going to leave it in, I think we should make an analogous discussion regarding
+how Hadoop's runtime varies with increasing resources. The data in Table 6 might
+be suitable for this.}
+
+To investigate the scalability characteristics of the HPC approach we ran the
+HPC pipeline using Dataset~I as input. We collected the running times as a
+function of the number of cores used (Table~\ref{table:3}). The results show that the
+alignment process with Bowtie scales fairly well, but overall the pipeline
+suffers because of the SNP calling step with Samtools. Indeed, this operation
+is the bottleneck of the workflow as it scales
+worse than Bowtie and consumes a progressively larger portion of the total
+calculation time as the input grows -- even after parallelizing analysis by
+chromosome as previously suggested~\cite{biostars_samtools}.
+
+%\footnote{\url{http://www.biostars.org/p/48781} \COM{here was a footnote
+%referring to biostars site. Ok, if the footnotes are not welcome, lets then
+%have it as a reference. That code is not trivial. Maybe we can leave the ref?}
+%\footnote{The amount of available memory, specified with the \texttt{-m}
+%option, is of big importance for the Samtools performance. During its sorting
+%phase the lack of RAM results in spilling to the HDD. For example for an half
+%of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and
+%15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our
+%tasks we used the default Samtool's settings.}.
+
+\subsection*{Comparing the efficiency of Hadoop and HPC}
+\Change{%
+Hadoop was designed to process huge datasets by distributing computations to a
+large number of computers\cite{hadoop,lin2010}.  Therefore, while allowing us to
+process very large datasets, the mechanisms required for distributing the work
+add a non-trivial overhead to the computation, which becomes more and more
+significant as the amount of work to be done is reduced -- i.e.\ the dataset
+size decreases.
+
+In order to measure the overhead incurred by the Hadoop platform, we compared
+the efficiency of the two processing methods (Hadoop and HPC) as the size of the
+dataset varies. We measure the efficiency indirectly as the number of core-hours
+($F$) used for same analysis: $F=T_{p}\times p,$ where~$T_{p}$ is the running time on~$p$ cores.
 }
@@
 \Change{%
-Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
-Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
-It is evident from the data that the Hadoop pipeline becomes increasingly more efficient than the HPC-based pipeline as the dataset size increases.
-The parabolic extrapolation of our data yields the value~$1.70\pm0.01$ for the ratio, meaning that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
-}
+\Fixme{The way of reorganized this text, at the end of the subsection it kind of
+excessively teeter-totters too much between HPC and Hadoop.}
 
-\Change{%
-The simplest curve that fits the data points nicely is a parabolic function which, importantly, is not a constant. 
-Since our measurements using the Hadoop-based pipeline, summmarized in Figure~\ref{fig:fig1}, showed linear scaling properties of Hadoop in this setting, a deviation of the ratio $F_{Hadoop}/F_{HPC}$ from the constant is evidence of the
-lack of scalability of the HPC approach.
-
+Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the
+ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
+%
+The simplest curve that fits the data points nicely is a parabolic function
+which, importantly, is not a constant.  Since our measurements using the
+Hadoop-based pipeline, summarized in Figure~\ref{fig:fig1}, showed linear
+scaling properties of Hadoop in this setting, a deviation of the ratio
+$F_{Hadoop}/F_{HPC}$ from the constant is evidence of the lack of scalability of
+the HPC approach.
+}%
 %Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
@@
 \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
-We suspect the main reason for such a behavior is that for large datasets (dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory- and time-demanding than the alignment with Bowtie.
-Since its memory requirements scale with input size, with sufficient input data Samtools swaps data to disk, creating multiple (up to hundreds) temporary files while it sorts the BAM file -- incidentally, these files can end up on shared network-accessed file systems and thus generate large amounts of network traffic.
-Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with different amount of RAM available on a single HPC node.
-Though there is a large deviation when using small amounts of RAM, we believe due to other users' load on the cluster network, the trend quite clearly indicates that a relatively large amount of RAM per node is needed in order to use Samtools effectively.
-In contrast, in our tests Hadoop performed the SNP calling on much more economic hardware in linear time with just 12GB of RAM per \textit{virtual} node.
+We suspect that the main reason for such a behavior is that for large datasets
+(Dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory-
+and time-demanding than the alignment with Bowtie.  Since its memory
+requirements scale with input size, with sufficient input data Samtools swaps
+data to disk, creating multiple (up to hundreds) temporary files while it sorts
+the BAM file -- incidentally, these files can end up on shared network-accessed
+file systems and thus generate large amounts of network traffic.
+Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with
+different amount of RAM available on a single HPC node. Though there is a large
+deviation when using small amounts of RAM -- we believe due to other users' load
+on the cluster network -- the trend quite clearly indicates that a relatively
+large amount of RAM per node is needed in order to use Samtools effectively. In
+contrast, in our tests Hadoop performed the SNP calling on much more economic
+hardware, in linear time, with just 12GB of RAM per \textit{virtual} node.
+
+\Change{%
+Nevertheless, it is evident from the plot that the HPC method is always more
+efficient in its use of CPU time than the Hadoop-based approach, and thus takes fewer computing
+resources to perform the same amount of work.  However, Hadoop becomes
+increasingly more efficient with respect to the HPC method as the dataset size
+increases.
+}
+However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
+infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+\Change{%
+In addition it is quite important to note that the price paid for
+the efficiency in HPC's method is in overall runtime and in its inability to
+finish the analysis on Dataset~V in a reasonable time (over 48 hours to Hadoop's 19.4
+hours; see Table~\ref{table:4}).
 }
 
-\subsection*{Comparing the network communication efficiency for the Hadoop and HPC approaches }
+\subsection*{Comparing the efficiency of network communication with the Hadoop and HPC approaches}
+
 The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
 The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
-To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gbases, dataset~IV).
+To study this phenomenon, we compared HPC and Hadoop network communication costs
+\Fixme{(maybe ``network load''?)}
+as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).
 
-Due to the trivially parallelizability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
-Reducing the size of each data chunk reduces the aligner  job,~$T_{alignment}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.
+Since the alignment process is embarrassingly parallel~-- the read-pairs are
+independent of each other and thus can be aligned independently~-- it is
+possible to distribute the work over more than one computational resource -- e.g., by
+splitting the initial data into chunks to processing them independently. In this
+approach, reducing the size of each data chunk reduces the size of the aligner
+job,~$T_{alignment}$ but, at the same time, the more chunks almost
+simultaneously have to be sent over the network, potentially causing traffic
+jams, and therefore increasing the communication costs,~$T_{comm}$.
 
-There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}, and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
+There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
+%
 However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
-To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo_bash}, and compared the network performance with the standard Hadoop HDFS approach.
-We separated the alignment time, $T_{alignment}$, and the communication time, $T_{comm}$, and plotted $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}).
-This measure is applicable to both HPC and Hadoop, however, $T_{comm}$ has a different explanation.
-For Hadoop, the short reads in FASTQ format have to be preprocessed (involving node communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data ingestion into the HDFS.
-We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelization.
-For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual alignment happens, and the sending of the aligned SAM files back to the delivery location over the network.
+\Fixme{(so, after reading the previous sentence, the logical question is ``why
+not use gnumap?'')}
+%
+To circumvent this limitation we implemented a highly optimized Bash script
+based on standard Unix utilities that makes very efficient use of the HPC
+cluster network~\cite{code_repo_bash}; we then compared the network
+performance with the standard Hadoop HDFS approach. We separated alignment
+time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
+the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
+(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
+however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in
+FASTQ format have to be preprocessed (involving node communication~$T_{comm}$)
+to be able to run in MR-fashion, while the data locality will be automatically
+achieved during the data ingestion into the HDFS\@.  We rewrote the code for the
+preprocessing stage for Crossbow to make it suitable for MR-style
+parallelization.  For the HPC approach, $T_{comm}$ involves the chunks that are
+sent from the sequence delivery location to the local node scratch disks where
+the actual alignment happens, and the sending of the aligned SAM files back to
+the delivery location over the network.
 %\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
@@
 %The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
-\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM.
+\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@.
 One can see two distinct linear stretches.
@@
 
-A popular way to construct and execute bioinformatic pipelines on HPC resources is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical user interface~(GUI) to bioinformatic programs, simplifying the experience for the end user.
-An alternative for Hadoop is Cloudgene\cite{cloudgene}, which is a light-weight and flexible Web-based solution for both public and private clouds.
-We implemented our Hadoop-pipeline in Cloudgene and extended the platform with functions to import data from the central file system at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).
-For our particular task in DNA sequencing, Cloudgene makes it easy to select data and parameters for its execution.
-Most of the data managing work is done automatically and the results can be downloaded to the client machine.
-The modular structure allows modification of the source code to adapt to the existing computing center architecture.
-For example, UPPMAX users can import their data from the sequencing platform directly to the Hadoop cluster by pressing a button and entering the credentials, being at the same time sure that their sensitive data is kept privately.
+\Fixme{Galaxy seems a bit out of place to me.
+\sout{A popular way to define and run bioinformatics pipelines on HPC resources
+is via the} Galaxy\cite{galaxy} \sout{platform, which provides a Web-based graphical
+user interface~(GUI) to bioinformatic programs, simplifying the experience for
+the end user.}
+}
+
+Hadoop presents a computing paradigm that is different from what most users are
+accustomed to using.  Therefore, adopting it requires an investment in learning
+how it works, how to use it, and how to understand what the programs running on
+the platform are doing.
+The Cloudgene\cite{cloudgene} GUI is available to help lower the learning curve
+for Hadoop adopters and is specifically targeted at users in the life
+sciences. It consists of a light-weight and flexible Web-based solution for both
+public and private clouds. We implemented our Hadoop-pipeline in Cloudgene and
+extended the platform with functions to import data from the central file system
+at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).  For our
+particular task related to DNA sequencing, Cloudgene makes it easy to select
+data and parameters for its execution. Most of the data managing work is done
+automatically and the results can be downloaded to the client machine. The
+modular structure allows modification of the source code to adapt to the
+existing computing center architecture. For example, UPPMAX users can import
+their data from the sequencing platform directly to the Hadoop cluster by
+pressing a button and entering the credentials; at the same time, the can be
+sure that their sensitive data is kept private.
 
@@
 
-In this report we have described two approaches for high-performance analysis of DNA sequencing data; one based on regular HPC using a batch system and one based on Hadoop.
-We show that Hadoop installed on a private cloud is an appealing solution both on terms of runtime performance and also in terms of usability.
-A key result is the data size where Hadoop is favorable to regular HPC batch systems, and in order to compare them we developed highly optimized pipelines for both scenarios.
-We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability for this problem.
-Extrapolation to infinite dataset size reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.
+In this report we have described two approaches for high-performance analysis of
+DNA sequencing data; one based on regular HPC using a batch system and one based
+on Hadoop. We show that Hadoop installed on a private cloud is an appealing
+solution both on terms of runtime performance and also in terms of usability. A
+key result is the data size where Hadoop becomes an advantageous option as compared
+to regular HPC batch systems.
+In order to compare the two platforms we developed highly optimized pipelines
+for both scenarios. We found that Hadoop excels over HPC when processing
+datasets larger than 100\,Gbases. Also, Hadoop shows almost linear scalability for
+this problem. Extrapolation to infinite dataset size reveals however that HPC
+provides the results faster, given the same amount of resources as Hadoop on a
+private cloud. \Fixme{But the HPC method tested ran on a single node so it
+didn't incur distribution costs!}
 
-We modified the preprocessing stage of the Crossbow software to enable it to make use of all nodes in the Hadoop system, end used a highly optimized Bash script to compare the scaling relations between the ratio of the alignment time to the communication time as a function of the reciprocal number of nodes on an HPC system.
-Our results show that the calculations on Hadoop with HDFS scales better than the network attached parallel storage commonly used in the HPC centers.
-In addition we show how performance of the HPC approach can be improved by redefining the queueing system's default behavior.
-This is however not of practical use in an HPC center since the network bandwidth is shared among multiple users.
-Finally, we demonstrate that an existing and extensible publicly available web-based GUI (Cloudgene) provides an easy way to execute bioinformatics analysis on Hadoop for those who are less experienced with Linux scripting.
-
-%%%%%%%%%%%
-% Methods
-%%%%%%%%%%%
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Methods}
-
-\subsection*{Datasets}
-We used publicly available DNA-seq datasets~(I--III) as well as a synthetic dataset~(IV) of {\it A.thaliana}, the well-known model plant, and a dataset (V) of two {\it H.sapiens} individuals (Table~\ref{table:datasets}).
-Data for datasets I-III and V were obtained using Illumina/HiSeq sequencing platforms.
-Further information about the datasets is provided in the Supplementary material section.
-
-\subsection*{Analysis pipelines}
-We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data; one for Hadoop and one for HPC.
-Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (e.g., GATK~\cite{gatk}), we decided to create the analysis pipeline as simple as possible to be able to make a better comparison with the one implemented for Hadoop.
-
-\begin{enumerate}
-\item HPC approach
-\subitem Short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: Samtools~ver.~0.1.19
-\item Hadoop approach: Crossbow
-\subitem Short read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: SOAPsnp~1.02
-\end{enumerate}
-
-In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the aligned reads and~SNP calling with Samtools.
-The Bowtie aligner natively implements~OMP, meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core.
-Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
-Where available these features were used to improve the analysis speed.
-The exact workflow used is available from github~\cite{code_repo_bash}.
-
-The equivalent Hadoop-based pipeline was implemented with Crossbow.
-The input data and the indexed genome reference were copied to Hadoop's storage system~(HDFS) before starting the experiments.
-Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
-Unfortunately, Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way.
-Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from github~\cite{code_repo_mr}).
-
-\subsection*{Computational resources}
-To run the HPC analysis pipeline we used a computational cluster at UPPMAX, which is heavily used for analysis of NGS-data~\cite{lampa}, on nodes equipped with 8 dual-core CPUs.
-Data and reference genomes were stored on a parallel, shared-storage system.
-The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager.
-The cluster was set up with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}.
-For details on computational resources, see Supplementary material.
-%\cite{gulo}
+We implemented a new MapReduce program to perform the preprocessing stage
+required by Crossbow in order to enable to Hadoop-based pipeline to effectively
+use the available computing resources. We also wrote a highly optimized Bash
+script to compare the scaling relations between the ratio of the alignment time
+to the communication time as a function of the reciprocal number of nodes on an
+HPC system.  Our results show that the calculations on Hadoop with HDFS scales
+better than the network attached parallel storage commonly used in the HPC
+centers.  In addition, we showed how performance of the HPC approach can be
+improved by redefining the queueing system's default behavior. This is however
+not of practical use in an HPC center since the network bandwidth is shared
+among multiple users. Finally, we demonstrated that an existing and extensible
+publicly available web-based GUI (Cloudgene) provides an easy way to execute
+bioinformatics analysis on Hadoop for those who are less experienced with Linux
+scripting.
 
@@
 %%%%%%%%%%%
-% Supplementary material
+% Supplementary Material
 %%%%%%%%%%%
-\section*{Supplementary material}
+\section*{Supplementary Material}
 
@@
 \begin{enumerate}
-%\item  HPC1:
+%\item  HPC:
 %The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbasesit/s uplink.
@@
 HPC:
-Multinode short read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8--core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
+Multinode short-read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8-core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
 
 \item Storage: 
-Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600 storage boxes and an additional node for metadata handling.
-In total, it provides roughly 1 PB of storage and is accessed with Lustre's own protocol.
-It supports data striping over multiple nodes and disk targets and can give a theoretical single file read performance of up to 80 Gbit per second.
+Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600
+storage boxes and an additional node for metadata handling. In total, it
+provides roughly 1 PB of storage and is accessed with Lustre's own protocol. It
+supports data striping over multiple nodes and disk targets and can give a
+theoretical single file read performance of up to 80 Gbits per second.
 
-\item Our Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud management system.
-Each node in this deployment was equipped with dual 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache), 16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet.
-The cluster was set up with Cloudera Hadoop Distribution version~2.0.0-cdh4.5.0~\cite{cloudera}.
-Note that the physical hardware provided less RAM than desired.
-Each VM node has 7 cores where each can use less than 2 GB of RAM, which is little for Hadoop.
-We expect much better performance with twice as much memory.
+\item Our Hadoop test platform was deployed on a private cloud at UPPMAX using
+	the OpenNebula~\cite{opennebula} cloud management system. Each physical node
+	was equipped with two 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache),
+	16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet. The cluster was set up
+	with Cloudera Hadoop Distribution version~2.0.0-cdh4.5.0~\cite{cloudera}.
+	Note that the physical hardware provided less RAM than desired. Each VM node
+	has 7 cores and 14~GB of RAM (2~GB/core), which is less than recommended for
+	Hadoop. We would expect to see much better performance with twice as much memory.
 \end{enumerate}
@@
 \subsection*{Crossbow preprocessing stage as a Bash script}
-Current Bash script mimics functionality of the Crossbow's preprocessing stage, and utilizes the multicore parallelism.
-The script reformats the BZIP2 archived short reads files in FASTQ format to a single text file, each line containing following tab-separated fields:
+
+The following Bash script mimics functionality of the Crossbow's preprocessing stage,
+and uses multi-core parallelism. The script reformats the
+\texttt{bzip2}-compressed short-reads in FASTQ format into a single text file
+where each line contains following tab-separated fields:
 \begin{itemize}
@@
 
-\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
+\begin{verbatim}
+pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}
+\end{verbatim}
 
@@
 	\input{./figures/fig1/fig1.tex}
-	\caption{Calculation time for selected datasets (labeled with roman numerals) for the 56 cores Hadoop cluster shows a linear scaling with dataset size. }
+	\caption{%
+	\Change{%
+		Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
+		increasing input dataset size. Each test dataset (labelled with Roman
+		numerals) was used. The image clearly highlights the linear scaling
+		characteristics of the Hadoop-based approach.
+		}
+		\Fixme{\\ What would you think of inserting line plotting the 16-core HPC
+		tests from Table 4}
+		}
 	\label{fig:fig1}
@@
 	\input{./figures/fig2/fig2.tex}
-	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gbases. Calculations were carried out for 56 cores Hadoop and 16 cores HPC cluster correspondingly.
-The points are fit to a quadratic least-squares curve, which makes it possible to predict {\it infinite} dataset size. Datasets are labeled with roman numerals.}
+	\caption{%
+	\Change{%
+	The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
+	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
+	a 16-core HPC node, respectively.
+	The points are fit to a quadratic least-squares curve, which makes it possible
+	to predict {\it infinite} dataset size. Datasets are labeled with Roman
+	numerals.
+	}
+	}
 	\label{fig:fig2}
@@
 %	\end{subfigure}
-	\caption{An example of a job setup with the graphical Hadoop front-end Cloudgene, providing a smooth user experience even for novice users.
-\textbf{a)} Pipeline selection - in our case containing the Crossbow pipeline.
-\textbf{b)} The UPPMAX-adapted functionality to browsing and import data from the user's home folder in the shared file system.}
+	\caption{%
+	An example of a job setup view with the graphical Hadoop front-end Cloudgene,
+	providing a smooth user experience even for novice users. \textbf{a)} Pipeline
+	selection -- in our case containing the Crossbow pipeline. \textbf{b)} The
+	UPPMAX-adapted functionality to browsing and import data from the user's home
+	folder in the shared file system.
+	}
 	\label{fig:fig4}
@@
 	\normalsize
-	\caption{Ratios of the alignment time~$T_{alignment}$ to the communication costs~$T_{comm}$ for HPC and Hadoop clusters for Dataset~IV as a function of reciprocal number of nodes~$1/N$.
-Two HPC scenarios are shown as ``HPC SLURM'' and ``HPC random'', which correspond to standard SLURM behavior and a modified one where nodes are being allocated from random racks.
-Linear fit  to $ax+b$ was done with the least-squares method.
-The estimated values for Hadoop approach are $(a,b)\approx(0.95,2.07)$.
-One  can see two defined linear regions for the HPC approach with very different tangents for linear and non-linear scalings correspondingly.
-For  the ``HPC SLURM'' $(a,b)\approx (0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random'' $(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.
- The constant term for HPC goes below unity, i.e. for the large number of nodes the communication will take more time than actual alignment. 
+\caption{%
+	\Change{%
+	Ratios of alignment time~$T_{alignment}$ to communication costs~$T_{comm}$ for
+	the HPC and Hadoop clusters as a function of reciprocal number of nodes~$1/N$.
+	All runs used Dataset~IV as input. Two HPC scenarios are shown as ``HPC
+	SLURM'' and ``HPC random'', which correspond to standard SLURM behavior and a
+	modified one that allocates nodes from random racks.  Linear fit to $ax+b$ was
+	done with the least-squares method.  The estimated coefficients for the
+	Hadoop-based approach are $(a,b)\approx(0.95,2.07)$.  One can see two defined
+	linear regions for the HPC approach with very different tangents for linear
+	and non-linear scalings correspondingly.  For  the ``HPC SLURM'' $(a,b)\approx
+	(0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random''
+	$(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.  The constant
+	term for HPC goes below unity -- i.e.\ for a large number of nodes more time
+	will be spent on communication than the actual alignment. 
+	}
+}
+
 %The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
@@
 %linearly to the reciprocal cluster size, and becomes more of less a constant, while the alignment time continues to satisfy~$T_{alignment}\propto 1/N$.
-}
 	\label{fig:fig3}
@@
 
-\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58 Gbases BAM file).
-The measurements are done on a single HPC node with $p=16$ cores.
-Averaging is done over 5 independent simulations.
-The amount of RAM is given for all 16 cores.
+\caption{%
+	\Change{%
+	Time (in minutes) required by Samtools to sort
+	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
+	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
+	amount of available RAM reported is for all 16 cores.
+	}
 }
@@
 \small
-\caption{Timings (in minutes) with Crossbow's native read preprocessor and with derived approach for  datasets II, III, IV  on a 56 cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node with 16 cores.}
+\caption{%
+	\Change{%
+	Time (in minutes) employed to preprocess Datasets II, III, and IV by
+	Crossbow's native read preprocessor and our MR preprocessor on a 56-core Hadoop
+	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
+	}
+}
+
 \begin{center}
@@
 Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
-BASH, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
+Bash, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
 \end{tabular}
@@
 
-\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.
-Dataset are labelled by roman numerals, ``f.r.'' stands for ``forward reads''.
-The large variance for the HPC deployment is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.}
+\caption{%
+	\Change{%
+	Timings (in minutes) for processing different dataset sizes on the HPC and
+	Hadoop deployments with their respective pipelines. Datasets are labelled by
+	Roman numerals, (``f.r.'' stands for ``forward reads''). The large variance
+	for the HPC deployment is due to Samtools BAM sorting, which is I/O and memory
+	intensive; the data splillage to the file system makes it very susceptible to
+	the load of the entire shared HPC cluster.
+	}
+}
+
 \begin{center}
@@
 
-data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
+Data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
 				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
@@
 Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
-HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
+HPC, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
 HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\
@@
 \begin{tabular}{|l|l|l|}
-dataset &	organism &	size in Gbases\\
+Dataset &	Organism &	Size in Gbases\\
 \hline
@@
 \begin{table}[!ht]
-\caption{Timings  for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
-For the ``HPC random'' approach, data chunks have to be copied to the local scratch disks first and the alignments (SAM files) copied back while Hadoop keeps all the data inside HDFS and hence does not need data staging.
-Hadoop however needs to preprocess reads before the actual alignment stage in order to be able to operate in MR manner resulting in what we term ``communication costs''.
-Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
+\caption{%
+	\Change{%
+	Timings for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and
+	Hadoop clusters for Dataset IV\@.  For the ``HPC random'' approach, data
+	chunks have to be copied to the local scratch disks first and the output alignments
+	(SAM files) copied back. On the contrary, Hadoop keeps all the data inside HDFS and hence
+	does not need data staging. However, Hadoop needs to preprocess reads before
+	the actual alignment stage to group read mates into the same record,
+	resulting in what we term ``communication costs''. Note that each HPC node has
+	16 cores, while each Hadoop node has 7 (one core is dedicated to run the
+	virtual machine).
+	}
+}
+
 \begin{center}
 \begin{tabular}{c|c|c|c|c|c}
- \multicolumn{3}{c|}{Hadoop} & \multicolumn{3}{c}{ HPC random} \\
+ \multicolumn{3}{c|}{Hadoop} & \multicolumn{3}{c}{HPC random} \\
  \hline

### 2014-03-14 paper-plos.tex
final

@@
 
-\Fixme{move this section back to the its original position if you don't like it
-here (it was just before Acknowledgements)}
-
 \subsection*{Analysis pipelines}
@@
 
-\Change{%
 The equivalent Hadoop-based pipeline was implemented with Crossbow. The input
@@
 containing the input data was mounted with SSHFS to one of the Hadoop nodes,
-from where the \texttt{bzip2}-compressed data were transferred to the HDFS\@.
-This preparation was done to eliminate unpredictable effects on the experiments
-from possible network delays.
-}
+from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
 Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
@@
 To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from GitHub~\cite{code_repo_mr}).
-Our implementation is described in Section~\ref{sec:preprocessor}.
+Our implementation is described in Section~{\it MapReduce-based Data Preprocessor}.
 
@@
 
-\Change{%
-All data was standardized to FASTQ format and compressed with \texttt{bzip2},
-since it is typical to store sequencing data in a compressed file format. The
-\texttt{bzip2} compression codec was selected it provides very good compression
+All data was standardized to FASTQ format and compressed. The
+\texttt{bzip2} compression codec was selected as it provides very good compression
 and is splittable by Hadoop, meaning that the archive can be expanded in a
 parallel manner allowing Hadoop to process multiple segments simultaneously.
-Also, since some of the datasets lack a standard FASTQ record id, which is
+Also, due to  lack a standard FASTQ record id, which is
 required by our test pipelines to extract the read number, we also standardised
@@
 function and labelled read mates with ``.1" and ``.2" suffixes.
-}
 
@@
 
-\Change{%
 The Crossbow package includes a native preprocessing stage that provides great
@@
 (this is done by listing all of them in a special \textit{manifest file}).
-Nevertheless, in practice this approach is rather error-prone and laborious.
 
@@
 was implemented in Python using the Hadoop streaming library.
-\Fixme{Consider deleting the following implementation detail.}
-To route both reads of the same pair to the same Reducer, a secondary
-key-sorting mechanism was employed.
+%\Fixme{Consider deleting the following implementation detail.}
+%To route both reads of the same pair to the same Reducer, a secondary key-sorting mechanism was employed.
 The resulting program efficiently processes short reads produced by different
@@
 computing facilities.
-}
 
@@
 
-\Change{%
 As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
@@
 measuring the wall-clock run time for each pipeline stage.  All experiments were
-repeated
-\Fixme{several times (how many?)}
+repeated at least three times, 
 and the results averaged to obtain a data point for the
 particular combination of data size and computational platform.
-}
-
-\Fixme{Belongs with Discussion:
-In light of our observations, we
-propose modifications for existing Hadoop software to better exploit the
-massively parallel nature of MR computations. Finally, we developed a set of Bash
-scripts to parallelize the operations of classical DNA-seq analysis programs for
-short-read alignment over multiple nodes and fully exploit the network bandwidth, thereby
-accelerating calculations.
-}
 
 \subsection*{Accuracy of Pipelines}
-\Change{%
 Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
@@
 chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.
-}
 
@@
 
-\Change{%
 To quantify the processing time saved by adopting the MR approach for
@@
 experiment comparing the new preprocessor described in
-Section~\ref{sec:preprocessor}  to the original preprocessor furnished with
+Section~{\it MapReduce-based Data Preprocessor} to the original preprocessor furnished with
 Crossbow.  To further validate our strategy, we also compared the performance of
@@
 equipped with 16 CPU cores.
-}
 
@@
 
-\Fixme{Alexey, please verify whether my interpretation seems reasonable. LP}
-
-\Change{%
 Table~\ref{table:preprocess} shows the results of the experiment. Clearly,
@@
 Bash approach.
-}
 
@@
 \subsection*{Scalability of HPC and Hadoop approaches}
-\Change{%
 Given the wide range in genome dataset sizes processed in practice -- e.g.,
-exomes,targeted sequencing, whole-genome sequencing at various depths -- it is
+exomes, targeted sequencing, whole-genome sequencing at various depths -- it is
 important to understand the scaling characteristics of each processing
@@
 
-The measurements highlight that one of the advantages of using the Hadoop~MR} framework is
+The measurements highlight that one of the advantages of using the Hadoop~MR framework is
 its almost linear scalability with respect to input size -- i.e.\ the
 calculation time linearly depends on the size of the
-dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
-%Deleting this since one could dream/find pathological cases that don't work: ``
-\Fixme{\sout{regardless the scaling nature of the underlying program.}}
+dataset, regardless the scaling nature of the underlying program~\cite{Langmead:2009kx,Pireddu:2011vn}.
 Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the wall clock
@@
 %least squares method.
-
-\Change{%
-Analogously, we investigated the scalability characteristics of the HPC method
-with respect to input size.
-\Fixme{The following works better the aid of a plot.}
-Analyzing the observations in Table~\ref{table:4}, we can see with the HPC
-method run time increases superlinearly as a function of dataset size.  As
-expected, the result is that the HPC method is not able to complete the analysis
-with our larger tests datasets.  On the other hand, this approach is not much
-slower than our Hadoop cluster considering the much smaller quantity of computing
-resources that is uses -- an issue that is analyzed more closely in the section
-on the efficiency of the two systems.
-}
-
-\Fixme{The following paragraph comments on how the HPC approaches' runtime
-varies with the number of cores available, rather than input size. If we're
-going to leave it in, I think we should make an analogous discussion regarding
-how Hadoop's runtime varies with increasing resources. The data in Table 6 might
-be suitable for this.}
 
@@
 \subsection*{Comparing the efficiency of Hadoop and HPC}
-\Change{%
-Hadoop was designed to process huge datasets by distributing computations to a
-large number of computers\cite{hadoop,lin2010}.  Therefore, while allowing us to
-process very large datasets, the mechanisms required for distributing the work
-add a non-trivial overhead to the computation, which becomes more and more
-significant as the amount of work to be done is reduced -- i.e.\ the dataset
-size decreases.
+Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
+One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
+In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the function showing the amount of used core-hours:
+$F=T_{p}\times p,$ 
+where~$T_{p}$ is the calculation time on~$p$ cores.
 
-In order to measure the overhead incurred by the Hadoop platform, we compared
-the efficiency of the two processing methods (Hadoop and HPC) as the size of the
-dataset varies. We measure the efficiency indirectly as the number of core-hours
-($F$) used for same analysis: $F=T_{p}\times p,$ where~$T_{p}$ is the running time on~$p$ cores.
-}
+Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
+Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
+As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
+The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
-\Change{%
-\Fixme{The way of reorganized this text, at the end of the subsection it kind of
-excessively teeter-totters too much between HPC and Hadoop.}
-
-Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the
-ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
-%
-The simplest curve that fits the data points nicely is a parabolic function
-which, importantly, is not a constant.  Since our measurements using the
-Hadoop-based pipeline, summarized in Figure~\ref{fig:fig1}, showed linear
-scaling properties of Hadoop in this setting, a deviation of the ratio
-$F_{Hadoop}/F_{HPC}$ from the constant is evidence of the lack of scalability of
-the HPC approach.
-}%
+The simplest curve to fit the points nicely is a parabolic function, importantly not a constant. 
+Since the Hadoop approach for the datasets reveals the linear scaling (Figure~\ref{fig:fig1}), a deviation of the ratio from the constant shows the 
+lack of scalability of the HPC approach.
 %Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
@@
 
-\Change{%
 Nevertheless, it is evident from the plot that the HPC method is always more
@@
 increases.
-}
+
 However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
 infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
-\Change{%
 In addition it is quite important to note that the price paid for
@@
 hours; see Table~\ref{table:4}).
-}
 
@@
 To study this phenomenon, we compared HPC and Hadoop network communication costs
-\Fixme{(maybe ``network load''?)}
 as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).
@@
 %
-However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
-\Fixme{(so, after reading the previous sentence, the logical question is ``why
-not use gnumap?'')}
-%
+However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.%
 To circumvent this limitation we implemented a highly optimized Bash script
@@
 however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in
-FASTQ format have to be preprocessed (involving node communication~$T_{comm}$)
+FASTQ format have to be preprocessed (referred to as~$T_{comm}$)
 to be able to run in MR-fashion, while the data locality will be automatically
@@
 preprocessing stage for Crossbow to make it suitable for MR-style
-parallelization.  For the HPC approach, $T_{comm}$ involves the chunks that are
+parallelization.  For the HPC approach,~$T_{comm}$ involves the chunks that are
 sent from the sequence delivery location to the local node scratch disks where
@@
 %\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
-and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot.
+and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency in the selected coordinates, which can be observed in the plot.
 The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
@@
 
-\Fixme{Galaxy seems a bit out of place to me.
-\sout{A popular way to define and run bioinformatics pipelines on HPC resources
-is via the} Galaxy\cite{galaxy} \sout{platform, which provides a Web-based graphical
+A popular way to define and run bioinformatics pipelines on HPC resources
+is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical
 user interface~(GUI) to bioinformatic programs, simplifying the experience for
-the end user.}
-}
+the end user.
 
@@
 In order to compare the two platforms we developed highly optimized pipelines
-for both scenarios. We found that Hadoop excels over HPC when processing
+for both scenarios. We found that Hadoop becomes comparable with HPC when processing
 datasets larger than 100\,Gbases. Also, Hadoop shows almost linear scalability for
@@
 provides the results faster, given the same amount of resources as Hadoop on a
-private cloud. \Fixme{But the HPC method tested ran on a single node so it
-didn't incur distribution costs!}
+private cloud.
 
@@
 centers.  In addition, we showed how performance of the HPC approach can be
-improved by redefining the queueing system's default behavior. This is however
-not of practical use in an HPC center since the network bandwidth is shared
-among multiple users. Finally, we demonstrated that an existing and extensible
+improved by redefining the queueing system's default behavior. Finally, we demonstrated that an existing and extensible
 publicly available web-based GUI (Cloudgene) provides an easy way to execute
@@
 
-\begin{verbatim}
-pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}
-\end{verbatim}
+\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc \${file2}) 
+| paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
 
@@
 	\caption{%
-	\Change{%
 		Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
@@
 		characteristics of the Hadoop-based approach.
-		}
-		\Fixme{\\ What would you think of inserting line plotting the 16-core HPC
-		tests from Table 4}
 		}
@@
 	\caption{%
-	\Change{%
 	The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
@@
 	numerals.
-	}
 	}
@@
 \caption{%
-	\Change{%
 	Ratios of alignment time~$T_{alignment}$ to communication costs~$T_{comm}$ for
@@
 	will be spent on communication than the actual alignment. 
-	}
 	}
@@
 \caption{%
-	\Change{%
 	Time (in minutes) required by Samtools to sort
@@
 	amount of available RAM reported is for all 16 cores.
-	}
 }
@@
 \caption{%
-	\Change{%
 	Time (in minutes) employed to preprocess Datasets II, III, and IV by
@@
 	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
-	}
 	}
@@
 \caption{%
-	\Change{%
-	Timings (in minutes) for processing different dataset sizes on the HPC and
+	Time (in minutes) for processing different dataset sizes on the HPC and
 	Hadoop deployments with their respective pipelines. Datasets are labelled by
@@
 	for the HPC deployment is due to Samtools BAM sorting, which is I/O and memory
-	intensive; the data splillage to the file system makes it very susceptible to
+	intensive; the data spillage to the file system makes it very susceptible to
 	the load of the entire shared HPC cluster.
-	}
 	}
@@
 \hline
-Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
 Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
-HPC, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
 HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\
@@
 \begin{table}[!ht]
-\caption{%
-	\Change{%
-	Timings for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and
-	Hadoop clusters for Dataset IV\@.  For the ``HPC random'' approach, data
-	chunks have to be copied to the local scratch disks first and the output alignments
-	(SAM files) copied back. On the contrary, Hadoop keeps all the data inside HDFS and hence
-	does not need data staging. However, Hadoop needs to preprocess reads before
-	the actual alignment stage to group read mates into the same record,
-	resulting in what we term ``communication costs''. Note that each HPC node has
-	16 cores, while each Hadoop node has 7 (one core is dedicated to run the
-	virtual machine).
-	}
-}
-
+\caption{Timings  for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
+For the ``HPC random'' approach, data chunks have to be copied to the local scratch disks first and the alignments (SAM files) copied back while Hadoop keeps all the data inside HDFS and hence does not need data staging.
+Hadoop however needs to preprocess reads before the actual alignment stage in order to be able to operate in MR manner resulting in what we term ``communication costs''.
+Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
 \begin{center}

### 2014-03-18 paper-plos.tex
Text editions + rewrote discussion  * Updated ordering of tables/figures

@@
 New experimental technologies in biology generate vast amounts of data that require
-high-performance computing
-to analyze and store them.
-Indeed, with the increase in dataset volumes there is a need for new computational tools and technologies that can cope with the scale of data.
+high-performance computing to analyze and store.
+With the increase in dataset volumes there is a need for new computational tools and technologies that can cope with the scale of data.
 In this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
@@
 A common step of NGS data analysis consists of aligning short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Many of the most widespread tools for this purpose -- like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools} -- are not made with distributed computing in mind.
+Many of the most widespread tools for this purpose -- like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools} -- are not designed with distributed computing in mind.
 Many others do not even have the native ability to use multiple cores.
@@
 %Current practices: How to parallelize
-The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16 (see~\cite{top500}).
+The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16~\cite{top500}.
 For the tools which do not support OMP natively -- e.g.
@@
 It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.
+A common way to implement multi-node parallelization is by using Message Passing Interface~(MPI)~\cite{mpi1}.
+However, writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code, and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
 
-The Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelization.
-However, writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code and there are few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
-
-Another common way to introduce parallelization into NGS analysis pipelines in Linux systems is to use Bash scripting.
+Another common way to introduce parallelization into NGS analysis pipelines in Linux systems is to use custom scripting in a language such as Bash, Perl, or Python.
 This involves using existing utilities and cluster tools to split the data into chunks, process them on the separate nodes, and merge the results afterwards.
 This kind of solution benefits from both MPI-like and OMP parallelization and provides good performance, but the development requires substantial expertise in order to be efficient.
-Since the process is tightly coupled to the local computational cluster and network architectures, it might also nearly impossible to re-use such developments in other settings.
+Since the process is tightly coupled to the local computational cluster and network architectures, it might also be nearly impossible to re-use such developments in other settings.
 
@@
 Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively enabling moving the computation to the data and thus reducing network traffic.
-HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
+HDFS provides a storage system where the bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
 
@@
 In this manuscript we focus on the question of {\it when} is Hadoop an appealing alternative to the programs for DNA-seq analysis generally found in~HPC centers.
-Since the Hadoop framework poses a significant performance overhead, we seek to estimate the average data size at which point, from a performance perspective, using it starts to pay off.
+Since the Hadoop framework poses a significant performance overhead, we seek to estimate the average data size at which point it starts to pay off from a performance perspective.
 We also study the analysis execution times for datasets of different sizes and evaluate Hadoop and the HPC approaches from a scaling perspective.
@@
 sequence analysis, we chose as a representative 
-use case for our experiments the task identifying
+use case for our experiments the task of identifying
 single-nucleotide polymorphisms (SNPs) from short-read data.  We constructed
@@
 and the other on the Hadoop platform.  Although there are many different
-approaches and software tools to perform this task in an HPC environment (e.g.,
-GATK~\cite{gatk}, \Fixme{Other example? Otherwise none}), there is only a limited
+approaches and software tools to perform this task in an HPC environment~\cite{Li:2013fk} there is only a limited
 number of options for Hadoop. Given this limitation, we decided to keep the
@@
 two platforms.
+%(e.g., GATK~\cite{gatk}, \Fixme{Other example? Otherwise none}),
 
@@
 
-In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the aligned reads and~SNP calling with Samtools.
+In the HPC pipeline, reads were aligned with Bowtie~\cite{Langmead:2009uq}, followed by sorting the aligned reads and~SNP calling with Samtools~\cite{samtools}.
 The Bowtie aligner natively implements~OMP -- therefore, using 8 cores on the same computer will theoretically produce a result 8 times faster than using a single core.
@@
 
-The equivalent Hadoop-based pipeline was implemented with Crossbow. The input
+The equivalent Hadoop-based pipeline was implemented with Crossbow~\cite{Langmead:2009kx}. The input
 data and the indexed genome reference were copied to Hadoop's storage
@@
 from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
-Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.
+Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP~\cite{soapsnp} to call SNPs.
 Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program).
@@
 parallel manner allowing Hadoop to process multiple segments simultaneously.
-Also, due to  lack a standard FASTQ record id, which is
-required by our test pipelines to extract the read number, we also standardised
+Also, due to the lack of a standard FASTQ record id, which is
+required by our test pipelines to extract the read number, we also standardized
 the read ids in the datasets by generating new ids based on the SHA-1 hash
@@
 from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
-Hadoop-based approach imposes the requirement that the input data be stored
-where the Hadoop nodes can access it, which is in our opinion a reasonable since
-sequencing platforms generally have direct access to extensive storage and
+Hadoop-based approach imposes the requirement that the input data is stored
+where the Hadoop nodes can access it, which is in our opinion does not bias the comparison since
+sequencing platforms in the general case have direct access to storage and
 computing facilities.
@@
 
-As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
-(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
-with a batch processing system (hereafter referred to as the HPC method)
-(Methods).  Our experiments consisted of executing the pipelines for each of the
-five input 
-datasets (see Table~\ref{table:datasets}) and
-measuring the wall-clock run time for each pipeline stage.  All experiments were
-repeated at least three times, 
+%As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
+%(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
+%with a batch processing system (hereafter referred to as the HPC method)
+%(Methods).
+
+We executed the two pipelines for each of the five 
+datasets on the HPC and Hadoop platforms, and measured the wall-clock run time for each pipeline stage.  
+All experiments were repeated at least three times, 
 and the results averaged to obtain a data point for the
@@
 Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
-SOAPsnp~\cite{soapsnp}, respectively) we could not expect them to deliver perfectly
+SOAPsnp) we could not expect them to deliver perfectly
 matching SNP lists. Instead, we verified the accuracy of the test pipelines by
@@
 our MapReduce approach to a different parallelization strategy based on a custom Bash script
-that uses of all processing cores available on a single computing node to accelerate the operation (see
+that uses all processing cores available on a single computing node to accelerate the operation (see
 Supplementary Material for details of the implementation). This program read its
@@
 
-Table~\ref{table:preprocess} shows the results of the experiment. Clearly,
-either of the parallel approaches are much faster than the original tool, with
+Table~\ref{table:preprocess} shows the results of the experiment, and we see that all 
+of of the parallel approaches are much faster than the original tool, for
 all dataset sizes.  Also, we noticed that the difference between the Bash and Hadoop
@@
 with increasing input size the superior scalability of MR becomes apparent, with
-the largest dataset taking requiring less than half the time of the Bash
-approach to be processed; at the same time, it is important to also notice that
+the largest dataset requiring less than half the time of the Bash
+approach to be processed. At the same time, it is important to also notice that
 the Hadoop approach employed approx.\ 1.6 times more CPU time compared to the
@@
 exomes, targeted sequencing, whole-genome sequencing at various depths -- it is
-important to understand the scaling characteristics of each processing
-platforms (Hadoop and HPC) with respect to input size. For each platform, we
-measured the time required for to execute their respective variant calling
-pipeline (see Methods) for each of the test datasets (I-V). The results are
-shown in Table~\ref{table:4}.
+important to understand the scaling characteristics of the Hadoop and HPC processing
+platforms with respect to input size. For each platform, we
+measured the time required to execute their respective variant calling
+pipeline for each of the test datasets (I-V). The results are
+shown in Table~\ref{table:pipleline-timings}.
 
-The measurements highlight that one of the advantages of using the Hadoop~MR framework is
+The measurements highlight that one of the advantages of using the Hadoop framework is
 its almost linear scalability with respect to input size -- i.e.\ the
@@
 dataset, regardless the scaling nature of the underlying program~\cite{Langmead:2009kx,Pireddu:2011vn}.
-Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the wall clock
+Figure~\ref{fig:fig1} shows the wall clock
 running time as a function of the dataset size for the 56-core Hadoop cluster and datasets
@@
 To investigate the scalability characteristics of the HPC approach we ran the
-HPC pipeline using Dataset~I as input. We collected the running times as a
-function of the number of cores used (Table~\ref{table:3}). The results show that the
+HPC pipeline using Dataset~I as input and collected the running times as a
+function of the number of cores used (Table~\ref{table:timings-hpc}). The results show that the
 alignment process with Bowtie scales fairly well, but overall the pipeline
@@
 \subsection*{Comparing the efficiency of Hadoop and HPC}
-Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}.
-One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach.
-In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the function showing the amount of used core-hours:
-$F=T_{p}\times p,$ 
-where~$T_{p}$ is the calculation time on~$p$ cores.
+\Fixme{New heading propsal: Efficiency of calculations}
 
-Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
+Since Hadoop was designed to digest huge datasets\cite{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach.
+In order to compare the 'suitability' for the Hadoop and HPC platforms run on different number of cores, we construct the function showing the amount of used core-hours:
+$F=T_{p}\times p,$ where~$T_{p}$ is the calculation time on~$p$ cores.
+
+Using the data from Table~\ref{table:pipleline-timings}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
 Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
-As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+We note that the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
+The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
@@
 (Dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory-
-and time-demanding than the alignment with Bowtie.  Since its memory
-requirements scale with input size, with sufficient input data Samtools swaps
-data to disk, creating multiple (up to hundreds) temporary files while it sorts
-the BAM file -- incidentally, these files can end up on shared network-accessed
+and time-demanding than the alignment with Bowtie.  Since Samtools memory
+requirements scale with input size, the application swaps
+data to disk for large input data creating multiple (up to hundreds) temporary files while it sorts
+the BAM file. These files can end up on shared network-accessed
 file systems and thus generate large amounts of network traffic.
-Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with
+Table~\ref{table:samtools-sort} shows the time required by Samtools to sort dataset~IV with
 different amount of RAM available on a single HPC node. Though there is a large
 deviation when using small amounts of RAM -- we believe due to other users' load
-on the cluster network -- the trend quite clearly indicates that a relatively
+on the shared cluster network -- the trend clearly indicates that a relatively
 large amount of RAM per node is needed in order to use Samtools effectively. In
-contrast, in our tests Hadoop performed the SNP calling on much more economic
-hardware, in linear time, with just 12GB of RAM per \textit{virtual} node.
+contrast, our tests Hadoop performed the SNP calling on much more economic
+hardware in linear time with just 12GB of RAM per virtual node.
 
-Nevertheless, it is evident from the plot that the HPC method is always more
-efficient in its use of CPU time than the Hadoop-based approach, and thus takes fewer computing
-resources to perform the same amount of work.  However, Hadoop becomes
-increasingly more efficient with respect to the HPC method as the dataset size
-increases.
-
-However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
-infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
-In addition it is quite important to note that the price paid for
-the efficiency in HPC's method is in overall runtime and in its inability to
-finish the analysis on Dataset~V in a reasonable time (over 48 hours to Hadoop's 19.4
-hours; see Table~\ref{table:4}).
+%However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
+%infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
 \subsection*{Comparing the efficiency of network communication with the Hadoop and HPC approaches}
+\Fixme{New heading propsal: Efficiency of network communication}
 
 The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
-The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
+The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance degradation.
 To study this phenomenon, we compared HPC and Hadoop network communication costs
 as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).
-
  Since the alignment process is embarrassingly parallel~-- the read-pairs are
 independent of each other and thus can be aligned independently~-- it is
-possible to distribute the work over more than one computational resource -- e.g., by
-splitting the initial data into chunks to processing them independently. In this
+possible to distribute the work over more than one computational resource, e.g. by
+splitting the initial data into chunks to processing them in parallel. In this
 approach, reducing the size of each data chunk reduces the size of the aligner
-job,~$T_{alignment}$ but, at the same time, the more chunks almost
-simultaneously have to be sent over the network, potentially causing traffic
-jams, and therefore increasing the communication costs,~$T_{comm}$.
+job,~$T_{alignment}$, but at the same time more chunks have to be sent over the network increasing the communication time,~$T_{comm}$.
 
 There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
-%
-However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.%
-To circumvent this limitation we implemented a highly optimized Bash script
-based on standard Unix utilities that makes very efficient use of the HPC
-cluster network~\cite{code_repo_bash}; we then compared the network
+We noted however that the Pmap package functions poorly for datasets larger than 20 Gbases, and instead implemented a highly optimized Bash script for alignment based on standard Unix utilities that makes very efficient use of the HPC
+cluster network (source code available on Github~\cite{code_repo_bash}). We then compared the network
 performance with the standard Hadoop HDFS approach. We separated alignment
@@
 (Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
-however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in
-FASTQ format have to be preprocessed (referred to as~$T_{comm}$)
+however $T_{comm}$ has a different explanation: For Hadoop, the short reads in
+FASTQ format have to be preprocessed
 to be able to run in MR-fashion, while the data locality will be automatically
-achieved during the data ingestion into the HDFS\@.  We rewrote the code for the
-preprocessing stage for Crossbow to make it suitable for MR-style
-parallelization.  For the HPC approach,~$T_{comm}$ involves the chunks that are
+achieved during the data ingestion into HDFS. For the HPC approach,~$T_{comm}$ involves the chunks that are
 sent from the sequence delivery location to the local node scratch disks where
-the actual alignment happens, and the sending of the aligned SAM files back to
+the actual alignment happens, and the transfer of the aligned SAM files back to
 the delivery location over the network.
@@
 
-\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: One can see that the ratio~$T_{alignment}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that Bowtie provides almost linear scaling between alignment time and dataset chunk size~$D$:~$T_{alignment}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and Figure~\ref{fig:fig1}.
+\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: The ratio~$T_{alignment}/T_{comm}$ indicates weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
+It is known that Bowtie provides almost linear scaling between alignment time and dataset chunk size~$D$:~$T_{alignment}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and our experiments summarized in Figure~\ref{fig:fig1}.
 Since the ratio~$T_{alignment}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
@@
 Scalability can be improved by overriding the default behavior of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
-Allocating the nodes on random racks allows one to engage more nodes without network saturation.
+Allocating the nodes on random racks allows for engaging more nodes without network saturation.
 For our cluster we could go up to 30-35 nodes with perfect linear scaling.
 For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:4}.
+$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:pipleline-timings}.
 The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
@@
 
-A popular way to define and run bioinformatics pipelines on HPC resources
-is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical
-user interface~(GUI) to bioinformatic programs, simplifying the experience for
-the end user.
-
 Hadoop presents a computing paradigm that is different from what most users are
@@
 the platform are doing.
-The Cloudgene\cite{cloudgene} GUI is available to help lower the learning curve
+A popular way to define and run bioinformatics pipelines on HPC and cloud resources
+is via the Galaxy\cite{galaxy,Afgan:2010uq} platform, which provides a Web-based graphical
+user interface~(GUI) to bioinformatic programs, simplifying the experience for
+the end user.
+A similar project is Cloudgene\cite{cloudgene}, which provides a GUI to help lower the learning curve
 for Hadoop adopters and is specifically targeted at users in the life
@@
 particular task related to DNA sequencing, Cloudgene makes it easy to select
-data and parameters for its execution. Most of the data managing work is done
+data and parameters for its execution. Most of the data management work is done
 automatically and the results can be downloaded to the client machine. The
@@
 their data from the sequencing platform directly to the Hadoop cluster by
-pressing a button and entering the credentials; at the same time, the can be
+pressing a button and entering the credentials; at the same time they can be
 sure that their sensitive data is kept private.
@@
 DNA sequencing data; one based on regular HPC using a batch system and one based
-on Hadoop. We show that Hadoop installed on a private cloud is an appealing
-solution both on terms of runtime performance and also in terms of usability. A
-key result is the data size where Hadoop becomes an advantageous option as compared
-to regular HPC batch systems.
-In order to compare the two platforms we developed highly optimized pipelines
-for both scenarios. We found that Hadoop becomes comparable with HPC when processing
-datasets larger than 100\,Gbases. Also, Hadoop shows almost linear scalability for
-this problem. Extrapolation to infinite dataset size reveals however that HPC
-provides the results faster, given the same amount of resources as Hadoop on a
-private cloud.
-
-We implemented a new MapReduce program to perform the preprocessing stage
-required by Crossbow in order to enable to Hadoop-based pipeline to effectively
-use the available computing resources. We also wrote a highly optimized Bash
-script to compare the scaling relations between the ratio of the alignment time
-to the communication time as a function of the reciprocal number of nodes on an
-HPC system.  Our results show that the calculations on Hadoop with HDFS scales
-better than the network attached parallel storage commonly used in the HPC
-centers.  In addition, we showed how performance of the HPC approach can be
-improved by redefining the queueing system's default behavior. Finally, we demonstrated that an existing and extensible
+on Hadoop. In order to compare the two platforms we developed highly optimized pipelines
+for both scenarios and found that Hadoop becomes comparable with HPC when processing
+datasets larger than 100\,Gbases.
+As this corresponds to typical human genome data with a sequencing depth of about~$30x$, we consider Hadoop not only to be a technology for the future but an attractive option for bioinformatics data analysis already today.
+While implementations on HPC are more efficient on the use of CPU time than the Hadoop-based approach, the price paid for this is increased overall runtime.
+The main benefit of Hadoop is however its almost linear scaling with the problem size, and we show that this also holds for analysis of NGS data.
+Our results further reveal that the calculations on Hadoop with the file system HDFS scales
+better than the network attached parallel storage commonly used in the HPC centers.  
+Finally, we demonstrated that an existing and extensible
 publicly available web-based GUI (Cloudgene) provides an easy way to execute
@@
 scripting.
+
+Hadoop has so far seen relatively low adoption in bioinformatics.
+This is likely due to that existing high-performance infrastructure in most cases are built as traditional HPC systems with batch processing queues.
+Also, the bioinformatics software flora is still not as big for Hadoop and most analysis tools are implemented as regular linux applications.
+This may very well change soon with the demonstrated scalability and wide uptake of Hadoop in other settings.
+The results in this paper show that Hadoop enables highly scalable massively parallel analysis of DNA sequence data, and as the size of data sets are expected to increase over time we envision that Hadoop will become increasingly popular also in bioinformatics to deliver results faster.
 
@@
 \begin{figure}[!ht]
- 	\begin{subfigure}[b]{0.4\textwidth}
-        	\includegraphics[width=\textwidth]{a.png}
-		\subcaption{}
-	\end{subfigure}
-	\hspace{0.5cm}
-	\begin{subfigure}[b]{0.6\textwidth}
-	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{}
-	\end{subfigure}
-%	\begin{subfigure}[b]{0.6\textwidth}
-%		\includegraphics[width=\textwidth]{b.png}
-%		\subcaption{Cloudgene: specifying job parameters}
-%	\end{subfigure}
-	\caption{%
-	An example of a job setup view with the graphical Hadoop front-end Cloudgene,
-	providing a smooth user experience even for novice users. \textbf{a)} Pipeline
-	selection -- in our case containing the Crossbow pipeline. \textbf{b)} The
-	UPPMAX-adapted functionality to browsing and import data from the user's home
-	folder in the shared file system.
-	}
-	\label{fig:fig4}
-\end{figure}
-
-\begin{figure}[!ht]
 	\small
@@
 
+\begin{figure}[!ht]
+ 	\begin{subfigure}[b]{0.4\textwidth}
+        	\includegraphics[width=\textwidth]{a.png}
+		\subcaption{}
+	\end{subfigure}
+	\hspace{0.5cm}
+	\begin{subfigure}[b]{0.6\textwidth}
+	        	\includegraphics[width=\textwidth]{c.png}
+		\subcaption{}
+	\end{subfigure}
+%	\begin{subfigure}[b]{0.6\textwidth}
+%		\includegraphics[width=\textwidth]{b.png}
+%		\subcaption{Cloudgene: specifying job parameters}
+%	\end{subfigure}
+	\caption{%
+	An example of a job setup view with the graphical Hadoop front-end Cloudgene,
+	providing a smooth user experience even for novice users. \textbf{a)} Pipeline
+	selection -- in our case containing the Crossbow pipeline. \textbf{b)} The
+	UPPMAX-adapted functionality to browsing and import data from the user's home
+	folder in the shared file system.
+	}
+	\label{fig:fig4}
+\end{figure}
+
 %%%%%%%%%%%
@@
 
- \begin{table}[hbtp]
+\begin{table}[!ht]
 \small
+\footnotesize
+\caption{Datasets used in the comparison. }
+\begin{center}
+\begin{tabular}{|l|l|l|}
+Dataset &	Organism &	Size in Gbases\\
+\hline
+ I		&	{\it A.thaliana}	&	1.4	\\
+ II	&	{\it A.thaliana}	&	7.0\\
+ III	&	{\it A.thaliana}	&	30.0	\\
+ IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
+ V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
 
-\caption{%
-	Time (in minutes) required by Samtools to sort
-	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
-	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
-	amount of available RAM reported is for all 16 cores.
-}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-RAM, GB		&	8			&		16			&			32		& 		64				\\
-timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
 \end{tabular}
 \end{center}
-\label{table:5}
+\label{table:datasets}
 \normalsize
@@
 \small
-\caption{Timings for Dataset I processed on different number of cores of a HPC node.}
-\begin{center}
-\begin{tabular}{c|c|c|c|ccc}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c}{speed-ups} \\
-\hline
-	& alignment 	&	SNP calling	&	total  &total & alignment& SNP call\\
-\hline
-1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
-2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
-4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
-6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
-8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
-10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
-12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
-14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
-16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
-\end{tabular}
-\end{center}
-\label{table:3}
-\normalsize
-\end{table}
-
-\begin{table}[!ht]
-\small
 
@@
 \end{center}
-\label{table:4}
+\label{table:pipleline-timings}
 \normalsize
@@
 \small
-\footnotesize
-\caption{Datasets used in the comparison. }
+\caption{Timings for Dataset I processed on different number of cores of a HPC node.}
 \begin{center}
-\begin{tabular}{|l|l|l|}
-Dataset &	Organism &	Size in Gbases\\
+\begin{tabular}{c|c|c|c|ccc}
+$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c}{speed-ups} \\
 \hline
- I		&	{\it A.thaliana}	&	1.4	\\
- II	&	{\it A.thaliana}	&	7.0\\
- III	&	{\it A.thaliana}	&	30.0	\\
- IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
- V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
-
+	& alignment 	&	SNP calling	&	total  &total & alignment& SNP call\\
+\hline
+1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
+2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
+4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
+6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
+8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
+10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
+12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
+14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
+16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
 \end{tabular}
 \end{center}
-\label{table:datasets}
+\label{table:timings-hpc}
+\normalsize
+\end{table}
+
+ \begin{table}[hbtp]
+\small
+
+\caption{%
+	Time (in minutes) required by Samtools to sort
+	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
+	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
+	amount of available RAM reported is for all 16 cores.
+}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+RAM, GB		&	8			&		16			&			32		& 		64				\\
+timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
+\end{tabular}
+\end{center}
+\label{table:samtools-sort}
 \normalsize
@@
 Hadoop however needs to preprocess reads before the actual alignment stage in order to be able to operate in MR manner resulting in what we term ``communication costs''.
-Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
+Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).\Fixme{This table is not referenced in text.
+Since we have a figure, do we really need the table?
+What extra does it add?
+}}
 \begin{center}
@@
 \end{center}
-\label{table:6}
+\label{table:}
 \end{table}%

### 2014-04-04 cover-plos.tex
Added cover letter

\documentclass[11pt]{article}

\usepackage[parfill]{parskip}

%\usepackage[small,compact]{titlesec}

\usepackage[swedish]{babel} 
\usepackage[T1]{fontenc} 
\usepackage[applemac]{inputenc} 
\usepackage{soul}

%Margins
\usepackage{anysize}	% \marginsize{left}{right}{top}{bottom}
\marginsize{3.2cm}{3.2cm}{2.1cm}{2.5cm}

%Remove page numbering
\pagenumbering{gobble}

\begin{document}

\title{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data}
\author{Cover Letter}
\date{}
\maketitle

Dear editors,\\

We hereby submit our manuscript entitled \textit{'A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data'}. Hadoop is becoming increasingly popular for parallelising data analysis in a variety of disciplines. While some challenges and results on Hadoop for analysing sequencing data have been reported~\cite{Marx:2013fk,gatk}, a more comprehensible assessment of this new framework is not available. In our manuscript we present the first quantitative comparison between Hadoop and traditional methods with high-performance computing resources for DNA sequence analysis in the form of short read alignment followed by variant calling. In order to make a fair comparison, we needed to define metrics and implement comparable protocols and pipelines, but also investigate user interfaces required to deliver these technologies as services for the computational biology community.

The authors have a long experience in providing high-performance e-infrastructure for biological sequence analysis, some of which was recently reported~\cite{lampa,Pireddu:2011vn}. Founded on the challenges we see in our current e-infrastructures we were motivated to investigate the benefits and drawbacks of the Hadoop framework and, more specifically, deduce at what problem sizes it becomes advantageous with respect to the current e-infrastructure best practices. As high-performance computing nowadays is used on wide scale in the computational biology community, we think that our paper would be of wide interest to the readers of PLoS Computational Biology as it can greatly affect how future large scale studies may be analyzed.

Sincerely,

Ola Spjuth and co-authors

\scriptsize
\bibliography{paper-plos}
\bibliographystyle{ola}

\end{document}

### 2014-04-04 paper-plos.tex
Updated for alexeys comments in external pdf

@@
 \subsection*{Scalability of HPC and Hadoop approaches}
-Given the wide range in genome dataset sizes processed in practice -- e.g.,
-exomes, targeted sequencing, whole-genome sequencing at various depths -- it is
+Given the wide range in genome dataset sizes processed in practice -- e.g.
+exomes and whole-genome sequencing at various depths -- it is
 important to understand the scaling characteristics of the Hadoop and HPC processing
@@
 
-\subsection*{Comparing the efficiency of Hadoop and HPC}
-\Fixme{New heading propsal: Efficiency of calculations}
+\subsection*{Efficiency of calculations}
 
@@
 We note that the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on 'bare metal' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
@@
 
-\subsection*{Comparing the efficiency of network communication with the Hadoop and HPC approaches}
-\Fixme{New heading propsal: Efficiency of network communication}
+\subsection*{Efficiency of network communication}
 
@@
 For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:pipleline-timings}.
+$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:timings-algn}.
 The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
@@
 \hline
-Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1.0$	\\
+Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673.0\pm1.0$	\\
 Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
@@
 Hadoop however needs to preprocess reads before the actual alignment stage in order to be able to operate in MR manner resulting in what we term ``communication costs''.
-Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).\Fixme{This table is not referenced in text.
-Since we have a figure, do we really need the table?
-What extra does it add?
-}}
+Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
 \begin{center}
@@
 \end{center}
-\label{table:}
+\label{table:timings-algn}
 \end{table}%

### 2014-04-04 paper-plos.tex
wrote github consistent

@@
 Where available these features were used to improve the analysis speed.
-The exact workflow used is available from GitHub~\cite{code_repo_bash}.
+The exact workflow used is available from Github~\cite{code_repo_bash}.
 
@@
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from GitHub~\cite{code_repo_mr}).
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from Github~\cite{code_repo_mr}).
 Our implementation is described in Section~{\it MapReduce-based Data Preprocessor}.

### 2014-04-04 paper-plos.tex
updated headings for Wes comments

@@
 
-\subsection*{Accuracy of Pipelines}
+\subsection*{Accuracy of pipelines}
 Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
@@
 
-\subsection*{Performance evaluation of MR-based preprocessing stage}
+\subsection*{Efficiency of preprocessing stage}
+%\subsection*{Performance evaluation of MR-based preprocessing stage}
 
@@
 
-\subsection*{Scalability of HPC and Hadoop approaches}
+\subsection*{Scalability}
 Given the wide range in genome dataset sizes processed in practice -- e.g.

### 2014-04-15 bioinformatics/document.tex
added paper in bioinformatics template

--- 2014-04-04 paper-plos.tex
@@
-% Template for PLoS
-% Version 1.0 January 2009
-%
-% To compile to pdf, run:
-% latex plos.template
-% bibtex plos.template
-% latex plos.template
-% latex plos.template
-% dvipdf plos.template
-
-\documentclass[10pt]{article}
-
-% amsmath package, useful for mathematical formulas
-\usepackage{amsmath}
-% amssymb package, useful for mathematical symbols
-\usepackage{amssymb}
-
-% graphicx package, useful for including eps and pdf graphics
-% include graphics with the command \includegraphics
-\usepackage{graphicx}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% required by the coloured \Fixme and \Change macros
-\usepackage{color}
-
-% Use doublespacing - comment out for single spacing
-%\usepackage{setspace}
-%\doublespacing
-
-\definecolor{ChangeColor}{rgb}{0.24,0.56,0.18}
-%\definecolor{ChangeColor}{rgb}{0,0,0}
-\newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}
-
-\definecolor{FixmeColor}{rgb}{1,0,0}
-%\definecolor{FixmeColor}{rgb}{0,0,0}
-\newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}
-
-% Text layout
-\topmargin 0.0cm
-\oddsidemargin 0.5cm
-\evensidemargin 0.5cm
-\textwidth 16cm
-\textheight 21cm
-
-% Bold the 'Figure #' in the caption and separate it with a period
-% Captions will be left justified
-\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
-
-% Use the PLoS provided bibtex style
-\bibliographystyle{plos2009}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-\pagestyle{myheadings}
-%% ** EDIT HERE **
-\newcommand{\COMMENT}[1]{{\color{red} #1 }}
-\newcommand{\COM}[1]{{\color{blue} #1 }}
+\documentclass{bioinfo}
+\copyrightyear{2014}
+\pubyear{2014}
 
 \graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
-\usepackage[normalem]{ulem}
-\usepackage{amssymb}
-\usepackage[colorlinks]{hyperref}
-\usepackage{booktabs}
-\usepackage{amsmath}
-%\usepackage{multirow}% http://ctan.org/pkg/multirow
-%\usepackage{tablefootnote}
-%\usepackage{import}
 \usepackage{subcaption}
-\usepackage{caption}
-%\usepackage[format=hang,font=small,labelfont=bf]{caption}
-
-%% ** EDIT HERE **
-%% PLEASE INCLUDE ALL MACROS BELOW
-
-%% END MACROS SECTION
 
 \begin{document}
+\firstpage{1}
 
-% Title must be 150 characters or less
-\begin{flushleft}
-{\Large
-\textbf{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
-}
-% Insert Author names, affiliations and corresponding author email.
-\\
-Alexey Siretskiy$^{1 \ast}$,
-Luca Pireddu$^{2}$,
-Ola Spjuth$^{3,4}$
-\\
-{\bf 1} Department of Information Technology, Uppsala University, Uppsala, Sweden
- \\
-{\bf 2} CRS4, Polaris, Pula, Italy
- \\
-{\bf 3} Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden
- \\
-{\bf 4} Science for Life Laboratory, Uppsala University, Uppsala, Sweden
-\\
-$\ast$ E-mail: alexey.siretskiy@it.uu.se
-\end{flushleft}
+\title[Hadoop for analyzing DNA sequence data]{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
+\author[Alexey Siretskiy]{Corresponding Author\,$^1$\footnote{to whom correspondence should be addressed}, Luca Pireddu\,$^{2}$ and Ola Spjuth\,$^{3,4}$}
+\address{$^{1}$Department of Information Technology, Uppsala University, Uppsala, Sweden\\
+$^{2}$CRS4, Polaris, Pula, Italy\\
+$^{3}$Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden\\
+$^{4}$Science for Life Laboratory, Uppsala University, Uppsala, Sweden}
 
-% Please keep the abstract between 250 and 300 words
-\section*{Abstract}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
+\maketitle
+
+\begin{abstract}
+
+\section{Motivation:}
 New high-throughput technologies such as massively parallel sequencing have transformed the life sciences into a data-intensive field.
 With increasing data volumes comes the necessity to analyze data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
-Hadoop, which has emerged over the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
+
+\section{Results:}
+Hadoop is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
 Herein we introduce metrics and report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step.
-% Author Summary not valid for PLoS ONE submissions.
-\section*{Author Summary}
-New experimental technologies in biology generate vast amounts of data that require
-high-performance computing to analyze and store.
-With the increase in dataset volumes there is a need for new computational tools and technologies that can cope with the scale of data.
-In this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
-While there is an overhead in using Hadoop, we show that as sequencing data sets reach sizes of 100 gigabases Hadoop becomes an attractive option compared with traditional methods.
-We also show that Hadoop scales linearly while traditional systems become network-saturated at larger data sizes.
-As sequencing data are bound to increase in the future, Hadoop is a very interesting framework on which to base future analysis.
+\section{Availability:}
+All code and data in this study is freely available as open source in public repositories.
 
-%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%
+\section{Contact:} \href{alexey.siretskiy@it.uu.se}{alexey.siretskiy@it.uu.se}
+\end{abstract}
 
-\section*{Introduction}
+\section{Introduction}
 %The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data providing insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
+Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data providing insight into the workings of biological machinery \citep{metzker, Marx:2013fk}.
 Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes.
 Extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline requiring a high-performance computing (HPC) infrastructure to produce results.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gbases) corresponding to 2\,terabytes (TB) of raw data in 3 days~\cite{illumina}.
+As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gbases) corresponding to 2\,terabytes (TB) of raw data in 3 days (http://www.illumina.com/systems/sequencing.ilmn).
 
@@
 A common step of NGS data analysis consists of aligning short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Many of the most widespread tools for this purpose -- like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools} -- are not designed with distributed computing in mind.
+Many of the most widespread tools for this purpose -- like BWA \citep{bwa}, Bowtie \citep{Langmead:2009uq} and Samtools \citep{samtools} -- are not designed with distributed computing in mind.
 Many others do not even have the native ability to use multiple cores.
@@
 %Current practices: How to parallelize
-The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16~\cite{top500}.
+The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP) (http://openmp.org/), but this approach is naturally limited by the number of cores per node, which does not usually exceed~16 (http://www.top500.org/statistics/list/).
 For the tools which do not support OMP natively -- e.g.
-Samtools -- variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
+Samtools -- variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility \citep{Tange2011a}.
 It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.
-A common way to implement multi-node parallelization is by using Message Passing Interface~(MPI)~\cite{mpi1}.
-However, writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code, and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
+A common way to implement multi-node parallelization is by using Message Passing Interface~(MPI) (http://www.mcs.anl.gov/research/projects/mpi/).
+However, writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code, and there are only a few existing solutions available for processing sequencing data \citep{gnumap} (http://erne.sourceforge.net, http://bmi.osu.edu/hpc/software/pmap/pmap.html).
+
+\begin{table*}[!ht]
+\small
+\footnotesize
+\caption{Datasets used in the comparison. }
+\begin{center}
+\begin{tabular}{|l|l|l|}
+Dataset &	Organism &	Size in Gbases\\
+\hline
+ I		&	{\it A.thaliana}	&	1.4	\\
+ II	&	{\it A.thaliana}	&	7.0\\
+ III	&	{\it A.thaliana}	&	30.0	\\
+ IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
+ V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
+
+\end{tabular}
+\end{center}
+\label{table:datasets}
+\normalsize
+\end{table*}%
 
@@
 %Properties of Hadoop and HDFS- data localization and programming model etc
-The Map-Reduce~(MR) programming paradigm~\cite{dean.2004.mapreduce} offers a compelling alternative for running tasks in a {\it massively} parallel way.
-This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
-The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}.
-The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies of data on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
+The Map-Reduce~(MR) programming paradigm \citep{dean.2004.mapreduce} offers a compelling alternative for running tasks in a {\it massively} parallel way.
+This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes \citep{lin2010}.
+The most prevalent open source implementation of Map-Reduce is Hadoop \citep{hadoop,Hadoop:Guide}.
+The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies of data on different nodes), and automated collection of results \citep{Hadoop:Guide}.
 Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively enabling moving the computation to the data and thus reducing network traffic.
-HDFS provides a storage system where the bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
+HDFS provides a storage system where the bandwidth and size scales with the number of nodes in the cluster \citep{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
 
@@
 
-%%%%%%%%%%%
-% Methods
-%%%%%%%%%%%
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Methods}
+\begin{methods}
+\section{Methods}
 
@@
 and the other on the Hadoop platform.  Although there are many different
-approaches and software tools to perform this task in an HPC environment~\cite{Li:2013fk} there is only a limited
+approaches and software tools to perform this task in an HPC environment \citep{Li:2013fk} there is only a limited
 number of options for Hadoop. Given this limitation, we decided to keep the
@@
 two platforms.
-%(e.g., GATK~\cite{gatk}, \Fixme{Other example? Otherwise none}),
+%(e.g., GATK \citep{gatk}, \Fixme{Other example? Otherwise none}),
 
-\begin{enumerate}
+\begin{enumerate}[(1)]
 \item HPC approach
-\subitem Short-read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: Samtools~ver.~0.1.19
+\begin{itemize}[]
+\item[] Short-read alignment: Bowtie~ver.~0.12.8
+\item[] SNP calling: Samtools~ver.~0.1.19
+\end{itemize}
 \item Hadoop approach: Crossbow
-\subitem Short-read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: SOAPsnp~1.02
+\begin{itemize}[]
+\item[] Short-read alignment: Bowtie~ver.~0.12.8
+\item[] SNP calling: SOAPsnp~1.02
+\end{itemize}
+
 \end{enumerate}
 
-In the HPC pipeline, reads were aligned with Bowtie~\cite{Langmead:2009uq}, followed by sorting the aligned reads and~SNP calling with Samtools~\cite{samtools}.
+\begin{figure*}[!hb]
+
+ \begin{subfigure}[b]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{figures/fig1-sep.png}
+	\caption{}
+		\label{fig:fig1}
+        \end{subfigure}%
+ \begin{subfigure}[b]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{figures/fig2-sep.png}
+	\caption{}
+	\label{fig:fig2}
+        \end{subfigure}
+	\caption{
+	\textbf{(a)} Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
+	increasing input dataset size. Each test dataset (labelled with Roman
+	numerals) was used. The image clearly highlights the linear scaling
+	characteristics of the Hadoop-based approach.\\
+	\textbf{(b)} The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
+	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
+	a 16-core HPC node, respectively.
+	The points are fit to a quadratic least-squares curve, which makes it possible
+	to predict {\it infinite} dataset size. Datasets are labeled with Roman
+	numerals.
+		}
+\end{figure*}
+
+In the HPC pipeline, reads were aligned with Bowtie \citep{Langmead:2009uq}, followed by sorting the aligned reads and~SNP calling with Samtools \citep{samtools}.
 The Bowtie aligner natively implements~OMP -- therefore, using 8 cores on the same computer will theoretically produce a result 8 times faster than using a single core.
@@
 Where available these features were used to improve the analysis speed.
-The exact workflow used is available from Github~\cite{code_repo_bash}.
+The exact workflow used is available from Github (https://github.com/raalesir/HPC\_bash\_align).
 
-The equivalent Hadoop-based pipeline was implemented with Crossbow~\cite{Langmead:2009kx}. The input
+The equivalent Hadoop-based pipeline was implemented with Crossbow \citep{Langmead:2009kx}. The input
 data and the indexed genome reference were copied to Hadoop's storage
@@
 from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
-Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP~\cite{soapsnp} to call SNPs.
+Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP (http://soap.genomics.org.cn/soapsnp.html) to call SNPs.
 Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program).
 Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from Github~\cite{code_repo_mr}).
+To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from Github https://github.com/raalesir/mr\_python).
 Our implementation is described in Section~{\it MapReduce-based Data Preprocessor}.
@@
 
+\begin{table*}[!ht]
+\small
+\caption{%
+	Time (in minutes) employed to preprocess Datasets II, III, and IV by
+	Crossbow's native read preprocessor and our MR preprocessor on a 56-core Hadoop
+	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
+	}
+
+\begin{center}
+\begin{tabular}{l|c|c|c}
+Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
+\hline
+Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673.0\pm1.0$	\\
+Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
+Bash, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
+\end{tabular}
+\end{center}
+\label{table:preprocess}
+\normalsize
+\end{table*}%
+
 \subsection*{MapReduce-based Data Preprocessor}
@@
 download data from Amazon S3, FTP and HTTP servers over the
-Internet~\cite{Langmead:2009kx}. However, this preprocessor does not parallelize
+Internet \citep{Langmead:2009kx}. However, this preprocessor does not parallelize
 its operations. In theory one can manually introduce some degree of
@@
 versions of the Illumina sequencing platform as well as FASTQ files generated
-from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
-Hadoop-based approach imposes the requirement that the input data is stored
+from SRA files (NCBI Sequence Read Archive, http://www.ncbi.nlm.nih.gov/Traces/sra). However, the Hadoop-based approach imposes the requirement that the input data is stored
 where the Hadoop nodes can access it, which is in our opinion does not bias the comparison since
@@
 To run the HPC analysis pipeline we used a computational cluster at UPPMAX,
-which is heavily used for analysis of NGS-data~\cite{lampa}.
+which is heavily used for analysis of NGS-data \citep{lampa}.
 Data and reference genomes were stored on a parallel, shared storage system.
 On the other hand, the Hadoop test platform was deployed on a private cloud at
-UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was
-configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}.
+UPPMAX using the OpenNebula (http://opennebula.org) cloud manager. The cluster was
+configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0 (http://www.cloudera.com/content/\\cloudera/en/why-cloudera/hadoop-and-big-data.html).
 For the fully detailed hardware configuration of the two computational resources
@@
 
-%%%%%%%%%%%%
-% Results
-%%%%%%%%%%%%
-% Results and Discussion can be combined.
-\section*{Results}
+\end{methods}
 
-%As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
-%(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
-%with a batch processing system (hereafter referred to as the HPC method)
-%(Methods).
+\begin{table*}
+\small
 
+\caption{%
+	Time (in minutes) for processing different dataset sizes on the HPC and
+	Hadoop deployments with their respective pipelines. Datasets are labelled by
+	Roman numerals, (``f.r.'' stands for ``forward reads''). The large variance
+	for the HPC deployment is due to Samtools BAM sorting, which is I/O and memory
+	intensive; the data spillage to the file system makes it very susceptible to
+	the load of the entire shared HPC cluster.
+	}
+
+\begin{center}
+\begin{tabular}{l|c|c|c|c|c|c|c}
+
+Data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
+				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
+\hline
+Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
+HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\
+
+\end{tabular}
+\end{center}
+\label{table:pipleline-timings}
+\normalsize
+\end{table*}%
+
+\begin{table*}[!ht]
+\small
+\caption{Timings for Dataset I processed on different number of cores of a HPC node.}
+\begin{center}
+\begin{tabular}{c|c|c|c|ccc}
+$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c}{speed-ups} \\
+\hline
+	& alignment 	&	SNP calling	&	total  &total & alignment& SNP call\\
+\hline
+1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
+2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
+4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
+6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
+8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
+10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
+12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
+14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
+16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
+\end{tabular}
+\end{center}
+\label{table:timings-hpc}
+\normalsize
+\end{table*}
+
+\section{Results}
 We executed the two pipelines for each of the five 
@@
 verifying their ability to correctly identify the mutation $C\rightarrow T$ on
-chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.
+chromosome 4 at position $16702262$ \citep{schneeberger} in Dataset I.
 
@@
 our MapReduce approach to a different parallelization strategy based on a custom Bash script
-that uses all processing cores available on a single computing node to accelerate the operation (see
-Supplementary Material for details of the implementation). This program read its
+that uses all processing cores available on a single computing node to accelerate the operation (see Supplementary Material for details of the implementation). This program read its
 input directly from conventional HPC storage, rather than HDFS\@.  The experiment
@@
 
+ \begin{table*}[hbtp]
+\small
+
+\caption{%
+	Time (in minutes) required by Samtools to sort
+	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
+	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
+	amount of available RAM reported is for all 16 cores.
+}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+RAM, GB		&	8			&		16			&			32		& 		64				\\
+timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
+\end{tabular}
+\end{center}
+\label{table:samtools-sort}
+\normalsize
+\end{table*}%
+
 %ran on a single HPC node, utilizes multiple cores, with the limiting factor
@@
 calculation time linearly depends on the size of the
-dataset, regardless the scaling nature of the underlying program~\cite{Langmead:2009kx,Pireddu:2011vn}.
+dataset, regardless the scaling nature of the underlying program \citep{Langmead:2009kx,Pireddu:2011vn}.
 Figure~\ref{fig:fig1} shows the wall clock
@@
 calculation time as the input grows -- even after parallelizing analysis by
-chromosome as previously suggested~\cite{biostars_samtools}.
+chromosome as previously suggested (http://www.biostars.org/p/48781).
+
+\begin{figure*}[!hb]
+	\small
+	\input{./figures/fig9/fig3.tex}
+	\normalsize
+\caption{%
+	Ratios of alignment time~$T_{alignment}$ to communication costs~$T_{comm}$ for
+	the HPC and Hadoop clusters as a function of reciprocal number of nodes~$1/N$.
+	All runs used Dataset~IV as input. Two HPC scenarios are shown as ``HPC
+	SLURM'' and ``HPC random'', which correspond to standard SLURM behavior and a
+	modified one that allocates nodes from random racks.  Linear fit to $ax+b$ was
+	done with the least-squares method.  The estimated coefficients for the
+	Hadoop-based approach are $(a,b)\approx(0.95,2.07)$.  One can see two defined
+	linear regions for the HPC approach with very different tangents for linear
+	and non-linear scalings correspondingly.  For  the ``HPC SLURM'' $(a,b)\approx
+	(0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random''
+	$(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.  The constant
+	term for HPC goes below unity -- i.e.\ for a large number of nodes more time
+	will be spent on communication than the actual alignment. 
+	}
+	\label{fig:fig3}
+\end{figure*}
 
@@
 
-Since Hadoop was designed to digest huge datasets\cite{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach.
+Since Hadoop was designed to digest huge datasets\citep{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach.
 In order to compare the 'suitability' for the Hadoop and HPC platforms run on different number of cores, we construct the function showing the amount of used core-hours:
@@
 
-\subsection*{Efficiency of network communication}
-
-The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
-The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance degradation.
-To study this phenomenon, we compared HPC and Hadoop network communication costs
-as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).
- Since the alignment process is embarrassingly parallel~-- the read-pairs are
-independent of each other and thus can be aligned independently~-- it is
-possible to distribute the work over more than one computational resource, e.g. by
-splitting the initial data into chunks to processing them in parallel. In this
-approach, reducing the size of each data chunk reduces the size of the aligner
-job,~$T_{alignment}$, but at the same time more chunks have to be sent over the network increasing the communication time,~$T_{comm}$.
-
-There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
-We noted however that the Pmap package functions poorly for datasets larger than 20 Gbases, and instead implemented a highly optimized Bash script for alignment based on standard Unix utilities that makes very efficient use of the HPC
-cluster network (source code available on Github~\cite{code_repo_bash}). We then compared the network
-performance with the standard Hadoop HDFS approach. We separated alignment
-time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
-the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
-(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
-however $T_{comm}$ has a different explanation: For Hadoop, the short reads in
-FASTQ format have to be preprocessed
-to be able to run in MR-fashion, while the data locality will be automatically
-achieved during the data ingestion into HDFS. For the HPC approach,~$T_{comm}$ involves the chunks that are
-sent from the sequence delivery location to the local node scratch disks where
-the actual alignment happens, and the transfer of the aligned SAM files back to
-the delivery location over the network.
-%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
-
-%The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
-
-\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: The ratio~$T_{alignment}/T_{comm}$ indicates weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that Bowtie provides almost linear scaling between alignment time and dataset chunk size~$D$:~$T_{alignment}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and our experiments summarized in Figure~\ref{fig:fig1}.
-Since the ratio~$T_{alignment}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
-
-%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
-\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@.
-One can see two distinct linear stretches.
-One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
-The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
-%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
-and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency in the selected coordinates, which can be observed in the plot.
-The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
-Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated on the same rack, which is the default SLURM behavior.
-
-Scalability can be improved by overriding the default behavior of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
-Allocating the nodes on random racks allows for engaging more nodes without network saturation.
-For our cluster we could go up to 30-35 nodes with perfect linear scaling.
-For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:timings-algn}.
-The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
-The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
-HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
-Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behavior is a generic for Hadoop with HDFS\cite{lin2010, Hadoop:Guide}.
-
-\subsection*{Usability aspects}
-\label{subsectionIV_2}
-
-Hadoop presents a computing paradigm that is different from what most users are
-accustomed to using.  Therefore, adopting it requires an investment in learning
-how it works, how to use it, and how to understand what the programs running on
-the platform are doing.
-A popular way to define and run bioinformatics pipelines on HPC and cloud resources
-is via the Galaxy\cite{galaxy,Afgan:2010uq} platform, which provides a Web-based graphical
-user interface~(GUI) to bioinformatic programs, simplifying the experience for
-the end user.
-A similar project is Cloudgene\cite{cloudgene}, which provides a GUI to help lower the learning curve
-for Hadoop adopters and is specifically targeted at users in the life
-sciences. It consists of a light-weight and flexible Web-based solution for both
-public and private clouds. We implemented our Hadoop-pipeline in Cloudgene and
-extended the platform with functions to import data from the central file system
-at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).  For our
-particular task related to DNA sequencing, Cloudgene makes it easy to select
-data and parameters for its execution. Most of the data management work is done
-automatically and the results can be downloaded to the client machine. The
-modular structure allows modification of the source code to adapt to the
-existing computing center architecture. For example, UPPMAX users can import
-their data from the sequencing platform directly to the Hadoop cluster by
-pressing a button and entering the credentials; at the same time they can be
-sure that their sensitive data is kept private.
-
-%%%%%%%%%%%%
-% Discussion
-%%%%%%%%%%%%
-\section*{Discussion}
-
-In this report we have described two approaches for high-performance analysis of
-DNA sequencing data; one based on regular HPC using a batch system and one based
-on Hadoop. In order to compare the two platforms we developed highly optimized pipelines
-for both scenarios and found that Hadoop becomes comparable with HPC when processing
-datasets larger than 100\,Gbases.
-As this corresponds to typical human genome data with a sequencing depth of about~$30x$, we consider Hadoop not only to be a technology for the future but an attractive option for bioinformatics data analysis already today.
-While implementations on HPC are more efficient on the use of CPU time than the Hadoop-based approach, the price paid for this is increased overall runtime.
-The main benefit of Hadoop is however its almost linear scaling with the problem size, and we show that this also holds for analysis of NGS data.
-Our results further reveal that the calculations on Hadoop with the file system HDFS scales
-better than the network attached parallel storage commonly used in the HPC centers.  
-Finally, we demonstrated that an existing and extensible
-publicly available web-based GUI (Cloudgene) provides an easy way to execute
-bioinformatics analysis on Hadoop for those who are less experienced with Linux
-scripting.
-
-Hadoop has so far seen relatively low adoption in bioinformatics.
-This is likely due to that existing high-performance infrastructure in most cases are built as traditional HPC systems with batch processing queues.
-Also, the bioinformatics software flora is still not as big for Hadoop and most analysis tools are implemented as regular linux applications.
-This may very well change soon with the demonstrated scalability and wide uptake of Hadoop in other settings.
-The results in this paper show that Hadoop enables highly scalable massively parallel analysis of DNA sequence data, and as the size of data sets are expected to increase over time we envision that Hadoop will become increasingly popular also in bioinformatics to deliver results faster.
-
-%%%%%%%%%%%
-% Acknowledgments
-%%%%%%%%%%%
-% Do NOT remove this, even if you are not including acknowledgments
-\section*{Acknowledgments}
-The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 Next Generation Sequencing Data Analysis Network, SeqAhead.
-We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
-Part of L.P.'s activity was performed within the context of the Ph.D.\ program in
-Biomedical Engineering at the University of Cagliari, Italy.
-
-%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%
-%\section*{References}
-% The bibtex filename
-\bibliography{paper-plos}
-
-%%%%%%%%%%%
-% Supplementary Material
-%%%%%%%%%%%
-\section*{Supplementary Material}
-
-\subsection*{Datasets}
-
-The datasets used in the paper are publicly available at:
-
-I: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0\\/data/reads/Schneeberger.2009/Schneeberger.2009.single\_end.gz}
-
-II: \url{http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads1.fq.gz, http://1001genomes.org/data/software/shoremap/shoremap\_2.0/data/reads/Galvao.2012/Galvao.2012.reads2.fq.gz}
-
-III: \url{ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611084//SRR611084.sra, ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR611/SRR611085//SRR611085.sra}
-
-IV: artificial pair-ended dataset for {\it A.thaliana} created with the {\tt wgsim} program from the Samtools package.
-
-V: \url{http://www.ncbi.nlm.nih.gov/sra/SRX148888}
-
-\subsection*{Reference genomes}
-\begin{itemize}
-\item TAIR10 for datasets II-IV \url{ftp://ftp.arabidopsis.org/home/tair/Sequences/whole\_chromosomes/*.fas}
-\item TAIR8 for dataset I \url{ftp://ftp.arabidopsis.org/home/tair/Genes/TAIR8\_genome\_release/}
-\item H.sapiens, NCBI v37 \url{ftp://ftp.ccb.jhu.edu/pub/data/bowtie\_indexes/h\_sapiens\_37\_asm.ebwt.zip}
-\end{itemize}
-
-\subsection*{Description of computational facilities}
-
-\begin{enumerate}
-%\item  HPC:
-%The HPC analysis pipeline was run on a node from the Kalkyl~\cite{kalkyl} cluster, equipped with two quad-core processors Intel Xeon~5520 (clock frequency of 2.26\,GHz; 1\,MB L2 cache, 8\,MB L3 cache), 24\,GB of RAM and an Infiniband node-to-node network connection, and 10Gbasesit/s uplink.
-The data and reference genomes were read and written to a parallel shared storage system.
-
-\item 
-HPC:
-Multinode short-read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8-core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
-
-\item Storage: 
-Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600
-storage boxes and an additional node for metadata handling. In total, it
-provides roughly 1 PB of storage and is accessed with Lustre's own protocol. It
-supports data striping over multiple nodes and disk targets and can give a
-theoretical single file read performance of up to 80 Gbits per second.
-
-\item Our Hadoop test platform was deployed on a private cloud at UPPMAX using
-	the OpenNebula~\cite{opennebula} cloud management system. Each physical node
-	was equipped with two 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache),
-	16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet. The cluster was set up
-	with Cloudera Hadoop Distribution version~2.0.0-cdh4.5.0~\cite{cloudera}.
-	Note that the physical hardware provided less RAM than desired. Each VM node
-	has 7 cores and 14~GB of RAM (2~GB/core), which is less than recommended for
-	Hadoop. We would expect to see much better performance with twice as much memory.
-\end{enumerate}
-
-\subsection*{Crossbow preprocessing stage as a Bash script}
-
-The following Bash script mimics functionality of the Crossbow's preprocessing stage,
-and uses multi-core parallelism. The script reformats the
-\texttt{bzip2}-compressed short-reads in FASTQ format into a single text file
-where each line contains following tab-separated fields:
-\begin{itemize}
-\item read header
-\item forward read
-\item forward read qualities
-\item reverse read
-\item reverse read qualities
-\end{itemize}
-
-\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc \${file2}) 
-| paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}
-
-%%%%%%%%%%%
-% Figure Legends
-%%%%%%%%%%%
-\section*{Figure Legends}
-
-\begin{figure}[!ht]
-	\input{./figures/fig1/fig1.tex}
-	\caption{%
-		Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
-		increasing input dataset size. Each test dataset (labelled with Roman
-		numerals) was used. The image clearly highlights the linear scaling
-		characteristics of the Hadoop-based approach.
-		}
-	\label{fig:fig1}
-\end{figure}
-
-\begin{figure}[!ht]
-	\input{./figures/fig2/fig2.tex}
-	\caption{%
-	The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
-	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
-	a 16-core HPC node, respectively.
-	The points are fit to a quadratic least-squares curve, which makes it possible
-	to predict {\it infinite} dataset size. Datasets are labeled with Roman
-	numerals.
-	}
-	\label{fig:fig2}
-\end{figure}
-
-\begin{figure}[!ht]
-	\small
-	\input{./figures/fig9/fig3.tex}
-	\normalsize
-\caption{%
-	Ratios of alignment time~$T_{alignment}$ to communication costs~$T_{comm}$ for
-	the HPC and Hadoop clusters as a function of reciprocal number of nodes~$1/N$.
-	All runs used Dataset~IV as input. Two HPC scenarios are shown as ``HPC
-	SLURM'' and ``HPC random'', which correspond to standard SLURM behavior and a
-	modified one that allocates nodes from random racks.  Linear fit to $ax+b$ was
-	done with the least-squares method.  The estimated coefficients for the
-	Hadoop-based approach are $(a,b)\approx(0.95,2.07)$.  One can see two defined
-	linear regions for the HPC approach with very different tangents for linear
-	and non-linear scalings correspondingly.  For  the ``HPC SLURM'' $(a,b)\approx
-	(0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random''
-	$(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.  The constant
-	term for HPC goes below unity -- i.e.\ for a large number of nodes more time
-	will be spent on communication than the actual alignment. 
-	}
-
-%The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
-%The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
-%linearly to the reciprocal cluster size, and becomes more of less a constant, while the alignment time continues to satisfy~$T_{alignment}\propto 1/N$.
-	\label{fig:fig3}
-\end{figure}
-
-\begin{figure}[!ht]
- 	\begin{subfigure}[b]{0.4\textwidth}
-        	\includegraphics[width=\textwidth]{a.png}
-		\subcaption{}
-	\end{subfigure}
-	\hspace{0.5cm}
-	\begin{subfigure}[b]{0.6\textwidth}
-	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{}
-	\end{subfigure}
-%	\begin{subfigure}[b]{0.6\textwidth}
-%		\includegraphics[width=\textwidth]{b.png}
-%		\subcaption{Cloudgene: specifying job parameters}
-%	\end{subfigure}
-	\caption{%
-	An example of a job setup view with the graphical Hadoop front-end Cloudgene,
-	providing a smooth user experience even for novice users. \textbf{a)} Pipeline
-	selection -- in our case containing the Crossbow pipeline. \textbf{b)} The
-	UPPMAX-adapted functionality to browsing and import data from the user's home
-	folder in the shared file system.
-	}
-	\label{fig:fig4}
-\end{figure}
-
-%%%%%%%%%%%
-% Tables
-%%%%%%%%%%%
-\section*{Tables}
-
-\begin{table}[!ht]
-\small
-\footnotesize
-\caption{Datasets used in the comparison. }
-\begin{center}
-\begin{tabular}{|l|l|l|}
-Dataset &	Organism &	Size in Gbases\\
-\hline
- I		&	{\it A.thaliana}	&	1.4	\\
- II	&	{\it A.thaliana}	&	7.0\\
- III	&	{\it A.thaliana}	&	30.0	\\
- IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
- V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
-
-\end{tabular}
-\end{center}
-\label{table:datasets}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\small
-\caption{%
-	Time (in minutes) employed to preprocess Datasets II, III, and IV by
-	Crossbow's native read preprocessor and our MR preprocessor on a 56-core Hadoop
-	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
-	}
-
-\begin{center}
-\begin{tabular}{l|c|c|c}
-Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
-\hline
-Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673.0\pm1.0$	\\
-Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
-Bash, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
-\end{tabular}
-\end{center}
-\label{table:preprocess}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\small
-
-\caption{%
-	Time (in minutes) for processing different dataset sizes on the HPC and
-	Hadoop deployments with their respective pipelines. Datasets are labelled by
-	Roman numerals, (``f.r.'' stands for ``forward reads''). The large variance
-	for the HPC deployment is due to Samtools BAM sorting, which is I/O and memory
-	intensive; the data spillage to the file system makes it very susceptible to
-	the load of the entire shared HPC cluster.
-	}
-
-\begin{center}
-\begin{tabular}{l|c|c|c|c|c|c|c}
-
-Data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
-				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
-\hline
-Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
-HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\
-
-\end{tabular}
-\end{center}
-\label{table:pipleline-timings}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\small
-\caption{Timings for Dataset I processed on different number of cores of a HPC node.}
-\begin{center}
-\begin{tabular}{c|c|c|c|ccc}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c}{speed-ups} \\
-\hline
-	& alignment 	&	SNP calling	&	total  &total & alignment& SNP call\\
-\hline
-1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
-2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
-4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
-6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
-8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
-10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
-12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
-14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
-16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
-\end{tabular}
-\end{center}
-\label{table:timings-hpc}
-\normalsize
-\end{table}
-
- \begin{table}[hbtp]
-\small
-
-\caption{%
-	Time (in minutes) required by Samtools to sort
-	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
-	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
-	amount of available RAM reported is for all 16 cores.
-}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-RAM, GB		&	8			&		16			&			32		& 		64				\\
-timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
-\end{tabular}
-\end{center}
-\label{table:samtools-sort}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
+\begin{table*}
 \caption{Timings  for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
@@
 \label{table:timings-algn}
-\end{table}%
+\end{table*}%
+
+\subsection*{Efficiency of network communication}
+
+The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
+The effective bandwidth of the Hadoop network increases with the cluster size \citep{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance degradation.
+To study this phenomenon, we compared HPC and Hadoop network communication costs
+as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).
+ Since the alignment process is embarrassingly parallel~-- the read-pairs are
+independent of each other and thus can be aligned independently~-- it is
+possible to distribute the work over more than one computational resource, e.g. by
+splitting the initial data into chunks to processing them in parallel. In this
+approach, reducing the size of each data chunk reduces the size of the aligner
+job, $T_{alignment}$, but at the same time more chunks have to be sent over the network increasing the communication time,~$T_{comm}$.
+
+There are several program packages for short-read alignment with MPI support \citep{gnumap} \\(http://bmi.osu.edu/hpc/software/pmap/pmap.html) and almost linear scaling up to 32 nodes for pair-ended reads has been reported \citep{Bozdag:2010cn}.
+We noted however that the Pmap package functions poorly for datasets larger than 20 Gbases, and instead implemented a highly optimized Bash script for alignment based on standard Unix utilities that makes very efficient use of the HPC
+cluster network (source code available on Github https://github.com/raalesir/HPC\_bash\_align). We then compared the network
+performance with the standard Hadoop HDFS approach. We separated alignment
+time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
+the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
+(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
+however $T_{comm}$ has a different explanation: For Hadoop, the short reads in
+FASTQ format have to be preprocessed
+to be able to run in MR-fashion, while the data locality will be automatically
+achieved during the data ingestion into HDFS. For the HPC approach,~$T_{comm}$ involves the chunks that are
+sent from the sequence delivery location to the local node scratch disks where
+the actual alignment happens, and the transfer of the aligned SAM files back to
+the delivery location over the network.
+%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
+
+%The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
+
+\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: The ratio $T_{alignment}/T_{comm}$ indicates weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
+It is known that Bowtie provides almost linear scaling between alignment time and dataset chunk size~$D$:~$T_{alignment}\propto  D\propto 1/N$, see \citep{Langmead:2009uq}, and our experiments summarized in Figure~\ref{fig:fig1}.
+Since the ratio~$T_{alignment}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
+
+%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
+\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@.
+One can see two distinct linear stretches.
+One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
+The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
+The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
+%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
+and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency in the selected coordinates, which can be observed in the plot.
+The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
+Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
+The nodes are being allocated on the same rack, which is the default SLURM behavior.
+
+\begin{figure*}
+	\centering
+ 	\begin{subfigure}[b]{0.4\textwidth}
+        	\includegraphics[width=\textwidth]{a.png}
+		\subcaption{}
+	\end{subfigure}
+	\hspace{1.5cm}
+	\begin{subfigure}[b]{0.36\textwidth}
+	        	\includegraphics[width=\textwidth]{c.png}
+		\subcaption{}
+	\end{subfigure}
+%	\begin{subfigure}[b]{0.6\textwidth}
+%		\includegraphics[width=\textwidth]{b.png}
+%		\subcaption{Cloudgene: specifying job parameters}
+%	\end{subfigure}
+	\caption{%
+	An example of a job setup view with the graphical Hadoop front-end Cloudgene,
+	providing a smooth user experience even for novice users. \textbf{a)} Pipeline
+	selection -- in our case containing the Crossbow pipeline. \textbf{b)} The
+	UPPMAX-adapted functionality to browsing and import data from the user's home
+	folder in the shared file system.
+	}
+	\label{fig:fig4}
+\end{figure*}
+
+Scalability can be improved by overriding the default behavior of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
+Allocating the nodes on random racks allows for engaging more nodes without network saturation.
+For our cluster we could go up to 30-35 nodes with perfect linear scaling.
+For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
+$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:timings-algn}.
+The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
+The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
+HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
+Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behavior is a generic for Hadoop with HDFS\citep{lin2010, Hadoop:Guide}.
+
+\subsection*{Usability aspects}
+\label{subsectionIV_2}
+
+Hadoop presents a computing paradigm that is different from what most users are
+accustomed to using.  Therefore, adopting it requires an investment in learning
+how it works, how to use it, and how to understand what the programs running on
+the platform are doing.
+A popular way to define and run bioinformatics pipelines on HPC and cloud resources
+is via the Galaxy\citep{galaxy,Afgan:2010uq} platform, which provides a Web-based graphical
+user interface~(GUI) to bioinformatic programs, simplifying the experience for
+the end user.
+A similar project is Cloudgene\citep{cloudgene}, which provides a GUI to help lower the learning curve
+for Hadoop adopters and is specifically targeted at users in the life
+sciences. It consists of a light-weight and flexible Web-based solution for both
+public and private clouds. We implemented our Hadoop-pipeline in Cloudgene and
+extended the platform with functions to import data from the central file system
+at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).  For our
+particular task related to DNA sequencing, Cloudgene makes it easy to select
+data and parameters for its execution. Most of the data management work is done
+automatically and the results can be downloaded to the client machine. The
+modular structure allows modification of the source code to adapt to the
+existing computing center architecture. For example, UPPMAX users can import
+their data from the sequencing platform directly to the Hadoop cluster by
+pressing a button and entering the credentials; at the same time they can be
+sure that their sensitive data is kept private.
+
+\section{Discussion}
+
+In this report we have described two approaches for high-performance analysis of
+DNA sequencing data; one based on regular HPC using a batch system and one based
+on Hadoop. In order to compare the two platforms we developed highly optimized pipelines
+for both scenarios and found that Hadoop becomes comparable with HPC when processing
+datasets larger than 100\,Gbases.
+As this corresponds to typical human genome data with a sequencing depth of about~$30x$, we consider Hadoop not only to be a technology for the future but an attractive option for bioinformatics data analysis already today.
+While implementations on HPC are more efficient on the use of CPU time than the Hadoop-based approach, the price paid for this is increased overall runtime.
+The main benefit of Hadoop is however its almost linear scaling with the problem size, and we show that this also holds for analysis of NGS data.
+Our results further reveal that the calculations on Hadoop with the file system HDFS scales
+better than the network attached parallel storage commonly used in the HPC centers.  
+Finally, we demonstrated that an existing and extensible
+publicly available web-based GUI (Cloudgene) provides an easy way to execute
+bioinformatics analysis on Hadoop for those who are less experienced with Linux
+scripting.
+
+Hadoop has so far seen relatively low adoption in bioinformatics.
+This is likely due to that existing high-performance infrastructure in most cases are built as traditional HPC systems with batch processing queues.
+Also, the bioinformatics software flora is still not as big for Hadoop and most analysis tools are implemented as regular linux applications.
+This may very well change soon with the demonstrated scalability and wide uptake of Hadoop in other settings.
+The results in this paper show that Hadoop enables highly scalable massively parallel analysis of DNA sequence data, and as the size of data sets are expected to increase over time we envision that Hadoop will become increasingly popular also in bioinformatics to deliver results faster.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+%     as it may ignore the figures.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Acknowledgement}
+We thank system experts Pontus Freyhult, Tore Sundqvist, and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage.
+We also thank Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
+
+\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
+This work was supported by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) [p2013023]; the Swedish strategic research programme eSSENCE; and COST Action BM1006 Next Generation Sequencing Data Analysis Network SeqAhead.
+Part of L.P.'s activity was performed within the context of the Ph.D.\ program in Biomedical Engineering at the University of Cagliari, Italy.\\
+
+\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+%
+%\bibliographystyle{plain}
+%
+%\bibliography{Document}
+
+\bibliography{paper}
 

### 2014-04-15 bioinformatics/figures/fig2/text.tex
added paper in bioinformatics template

--- 2014-01-15 figures/fig2/text.tex
@@
 \begin{center}
-\input{fig.tex}
+\input{fig2.tex}
 \end{center}
-The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC correspondingly.
-The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.
+The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=16$ cores for Hadoop and HPC correspondingly.
+The points are fit in a curve  using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.
 \end{document}

### 2014-04-15 bioinformatics/figures/fig3/makefig1.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+pdflatex text.tex && open text.pdf

### 2014-04-16 bioinformatics/document.tex
Added supplement and updated for alexey comments

@@
 \title[Hadoop for analyzing DNA sequence data]{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
-\author[Alexey Siretskiy]{Corresponding Author\,$^1$\footnote{to whom correspondence should be addressed}, Luca Pireddu\,$^{2}$ and Ola Spjuth\,$^{3,4}$}
+\author[Alexey Siretskiy]{Alexey Siretskiy\,$^1$\footnote{to whom correspondence should be addressed}, Luca Pireddu\,$^{2}$, Tore Sundqvist\,$^1$ and Ola Spjuth\,$^{3,4}$}
 \address{$^{1}$Department of Information Technology, Uppsala University, Uppsala, Sweden\\
@@
 	to predict {\it infinite} dataset size. Datasets are labeled with Roman
-	numerals.
+	numerals; f.r. stands for "forward reads".
 		}
@@
 HiSeq sequencing data. Additional information about the source of the datasets is
-provided in the Supplementary Material section.
+provided in the Supplementary Data.
 
@@
 For the fully detailed hardware configuration of the two computational resources
-please see the Supplementary Material section.
+please see the Supplementary Data.
 
@@
 our MapReduce approach to a different parallelization strategy based on a custom Bash script
-that uses all processing cores available on a single computing node to accelerate the operation (see Supplementary Material for details of the implementation). This program read its
+that uses all processing cores available on a single computing node to accelerate the operation (see Supplementary Data for details of the implementation). This program read its
 input directly from conventional HPC storage, rather than HDFS\@.  The experiment
@@
 
-Since Hadoop was designed to digest huge datasets\citep{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach.
+Since Hadoop was designed to digest huge datasets \citep{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach.
 In order to compare the 'suitability' for the Hadoop and HPC platforms run on different number of cores, we construct the function showing the amount of used core-hours:
@@
 We note that the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on 'bare metal' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
+The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on "bare metal" of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
 
@@
 the end user.
-A similar project is Cloudgene\citep{cloudgene}, which provides a GUI to help lower the learning curve
+A similar project is Cloudgene \citep{cloudgene}, which provides a GUI to help lower the learning curve
 for Hadoop adopters and is specifically targeted at users in the life
@@
 \section*{Acknowledgement}
-We thank system experts Pontus Freyhult, Tore Sundqvist, and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage.
+We thank system experts Pontus Freyhult, and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage.
 We also thank Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
 
-\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
-This work was supported by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) [p2013023]; the Swedish strategic research programme eSSENCE; and COST Action BM1006 Next Generation Sequencing Data Analysis Network SeqAhead.
+\paragraph{Funding\textcolon} This work was supported by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) [p2013023]; the Swedish strategic research programme eSSENCE; and COST Action BM1006 Next Generation Sequencing Data Analysis Network SeqAhead.
 Part of L.P.'s activity was performed within the context of the Ph.D.\ program in Biomedical Engineering at the University of Cagliari, Italy.\\

### 2014-04-16 bioinformatics/supplement.tex
Added supplement and updated for alexey comments

--- 2014-04-04 paper-plos.tex
@@
-% Template for PLoS
-% Version 1.0 January 2009
-%
-% To compile to pdf, run:
-% latex plos.template
-% bibtex plos.template
-% latex plos.template
-% latex plos.template
-% dvipdf plos.template
+\documentclass[11pt]{article}
 
-\documentclass[10pt]{article}
-
-% amsmath package, useful for mathematical formulas
-\usepackage{amsmath}
-% amssymb package, useful for mathematical symbols
-\usepackage{amssymb}
-
-% graphicx package, useful for including eps and pdf graphics
-% include graphics with the command \includegraphics
-\usepackage{graphicx}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% required by the coloured \Fixme and \Change macros
-\usepackage{color}
-
-% Use doublespacing - comment out for single spacing
-%\usepackage{setspace}
-%\doublespacing
-
-\definecolor{ChangeColor}{rgb}{0.24,0.56,0.18}
-%\definecolor{ChangeColor}{rgb}{0,0,0}
-\newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}
-
-\definecolor{FixmeColor}{rgb}{1,0,0}
-%\definecolor{FixmeColor}{rgb}{0,0,0}
-\newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}
-
-% Text layout
-\topmargin 0.0cm
-\oddsidemargin 0.5cm
-\evensidemargin 0.5cm
-\textwidth 16cm
-\textheight 21cm
-
-% Bold the 'Figure #' in the caption and separate it with a period
-% Captions will be left justified
-\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
-
-% Use the PLoS provided bibtex style
-\bibliographystyle{plos2009}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-\pagestyle{myheadings}
-%% ** EDIT HERE **
-\newcommand{\COMMENT}[1]{{\color{red} #1 }}
-\newcommand{\COM}[1]{{\color{blue} #1 }}
-
-\graphicspath { {./figures/fig1/} {./figures/fig2/} {./figures/fig9/} {./figures/fig5/}}
-\usepackage[normalem]{ulem}
-\usepackage{amssymb}
 \usepackage[colorlinks]{hyperref}
-\usepackage{booktabs}
-\usepackage{amsmath}
-%\usepackage{multirow}% http://ctan.org/pkg/multirow
-%\usepackage{tablefootnote}
-%\usepackage{import}
-\usepackage{subcaption}
-\usepackage{caption}
-%\usepackage[format=hang,font=small,labelfont=bf]{caption}
-
-%% ** EDIT HERE **
-%% PLEASE INCLUDE ALL MACROS BELOW
-
-%% END MACROS SECTION
 
@@
 
-% Title must be 150 characters or less
-\begin{flushleft}
-{\Large
-\textbf{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
-}
-% Insert Author names, affiliations and corresponding author email.
-\\
-Alexey Siretskiy$^{1 \ast}$,
-Luca Pireddu$^{2}$,
-Ola Spjuth$^{3,4}$
-\\
-{\bf 1} Department of Information Technology, Uppsala University, Uppsala, Sweden
- \\
-{\bf 2} CRS4, Polaris, Pula, Italy
- \\
-{\bf 3} Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden
- \\
-{\bf 4} Science for Life Laboratory, Uppsala University, Uppsala, Sweden
-\\
-$\ast$ E-mail: alexey.siretskiy@it.uu.se
-\end{flushleft}
-
-% Please keep the abstract between 250 and 300 words
-\section*{Abstract}
-New high-throughput technologies such as massively parallel sequencing have transformed the life sciences into a data-intensive field.
-With increasing data volumes comes the necessity to analyze data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging.
-Hadoop, which has emerged over the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes.
-Herein we introduce metrics and report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases.
-In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts.
-From the observed scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100\,gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution.
-As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in the future of biological data analysis.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step.
-% Author Summary not valid for PLoS ONE submissions.
-\section*{Author Summary}
-New experimental technologies in biology generate vast amounts of data that require
-high-performance computing to analyze and store.
-With the increase in dataset volumes there is a need for new computational tools and technologies that can cope with the scale of data.
-In this paper we provide a detailed study comparing the performance of the Hadoop framework with traditional high-performance computing for data in biological sequencing.
-While there is an overhead in using Hadoop, we show that as sequencing data sets reach sizes of 100 gigabases Hadoop becomes an attractive option compared with traditional methods.
-We also show that Hadoop scales linearly while traditional systems become network-saturated at larger data sizes.
-As sequencing data are bound to increase in the future, Hadoop is a very interesting framework on which to base future analysis.
-
-%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%
-
-\section*{Introduction}
-%The problem: dealing with BIG data
-Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data providing insight into the workings of biological machinery~\cite{metzker, Marx:2013fk}.
-Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes.
-Extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline requiring a high-performance computing (HPC) infrastructure to produce results.
-As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 1800\,gigabases (Gbases) corresponding to 2\,terabytes (TB) of raw data in 3 days~\cite{illumina}.
-
-%Current practices: HPC systems
-A common step of NGS data analysis consists of aligning short reads to a reference sequence and then finding the genetic variations specific to the sample.
-Many of the most widespread tools for this purpose -- like BWA~\cite{bwa}, Bowtie~\cite{Langmead:2009uq} and Samtools~\cite{samtools} -- are not designed with distributed computing in mind.
-Many others do not even have the native ability to use multiple cores.
-
-%Current practices: How to parallelize
-The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16~\cite{top500}.
-For the tools which do not support OMP natively -- e.g.
-Samtools -- variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
-It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.
-A common way to implement multi-node parallelization is by using Message Passing Interface~(MPI)~\cite{mpi1}.
-However, writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronization (or load balancing) has to be woven into the software code, and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.
-
-Another common way to introduce parallelization into NGS analysis pipelines in Linux systems is to use custom scripting in a language such as Bash, Perl, or Python.
-This involves using existing utilities and cluster tools to split the data into chunks, process them on the separate nodes, and merge the results afterwards.
-This kind of solution benefits from both MPI-like and OMP parallelization and provides good performance, but the development requires substantial expertise in order to be efficient.
-Since the process is tightly coupled to the local computational cluster and network architectures, it might also be nearly impossible to re-use such developments in other settings.
-
-%Introduce Hadoop, MR, and HDFS
-%Properties of Hadoop and HDFS- data localization and programming model etc
-The Map-Reduce~(MR) programming paradigm~\cite{dean.2004.mapreduce} offers a compelling alternative for running tasks in a {\it massively} parallel way.
-This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
-The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}.
-The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies of data on different nodes), and automated collection of results~\cite{Hadoop:Guide}.
-Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively enabling moving the computation to the data and thus reducing network traffic.
-HDFS provides a storage system where the bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.
-
-%Manuscript focus: When is Hadoop currently an appealing alternative
-%Introduce what we did in the manuscript.
-In this manuscript we focus on the question of {\it when} is Hadoop an appealing alternative to the programs for DNA-seq analysis generally found in~HPC centers.
-Since the Hadoop framework poses a significant performance overhead, we seek to estimate the average data size at which point it starts to pay off from a performance perspective.
-We also study the analysis execution times for datasets of different sizes and evaluate Hadoop and the HPC approaches from a scaling perspective.
-
-%%%%%%%%%%%
-% Methods
-%%%%%%%%%%%
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Methods}
-
-\subsection*{Analysis pipelines}
-
-For our work to determine when Hadoop is an appealing option for genomic
-sequence analysis, we chose as a representative 
-use case for our experiments the task of identifying
-single-nucleotide polymorphisms (SNPs) from short-read data.  We constructed
-two pipelines that perform the same task: one that runs in an HPC environment
-and the other on the Hadoop platform.  Although there are many different
-approaches and software tools to perform this task in an HPC environment~\cite{Li:2013fk} there is only a limited
-number of options for Hadoop. Given this limitation, we decided to keep the
-analysis pipeline as simple as possible to ensure a fair comparison between the
-two platforms.
-%(e.g., GATK~\cite{gatk}, \Fixme{Other example? Otherwise none}),
-
-\begin{enumerate}
-\item HPC approach
-\subitem Short-read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: Samtools~ver.~0.1.19
-\item Hadoop approach: Crossbow
-\subitem Short-read alignment: Bowtie~ver.~0.12.8
-\subitem SNP calling: SOAPsnp~1.02
-\end{enumerate}
-
-In the HPC pipeline, reads were aligned with Bowtie~\cite{Langmead:2009uq}, followed by sorting the aligned reads and~SNP calling with Samtools~\cite{samtools}.
-The Bowtie aligner natively implements~OMP -- therefore, using 8 cores on the same computer will theoretically produce a result 8 times faster than using a single core.
-Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions.
-Where available these features were used to improve the analysis speed.
-The exact workflow used is available from Github~\cite{code_repo_bash}.
-
-The equivalent Hadoop-based pipeline was implemented with Crossbow~\cite{Langmead:2009kx}. The input
-data and the indexed genome reference were copied to Hadoop's storage
-system~(HDFS) before starting the experiments.  Specifically, the storage volume
-containing the input data was mounted with SSHFS to one of the Hadoop nodes,
-from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
-Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP~\cite{soapsnp} to call SNPs.
-Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program).
-Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments.
-To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from Github~\cite{code_repo_mr}).
-Our implementation is described in Section~{\it MapReduce-based Data Preprocessor}.
-
-\subsection*{Datasets}
-For our experiments we used three publicly available DNA-seq datasets~(I--III),
-a synthetic dataset~(IV) of {\it A.thaliana} -- the well-known model plant --
-and a dataset (V) of two {\it H.sapiens} individuals
-(Table~\ref{table:datasets}). Datasets I-III and V are composed of Illumina
-HiSeq sequencing data. Additional information about the source of the datasets is
-provided in the Supplementary Material section.
-
-All data was standardized to FASTQ format and compressed. The
-\texttt{bzip2} compression codec was selected as it provides very good compression
-and is splittable by Hadoop, meaning that the archive can be expanded in a
-parallel manner allowing Hadoop to process multiple segments simultaneously.
-Also, due to the lack of a standard FASTQ record id, which is
-required by our test pipelines to extract the read number, we also standardized
-the read ids in the datasets by generating new ids based on the SHA-1 hash
-function and labelled read mates with ``.1" and ``.2" suffixes.
-
-\subsection*{MapReduce-based Data Preprocessor}
-\label{sec:preprocessor}
-
-The Crossbow package includes a native preprocessing stage that provides great
-flexibility in delivering the data to the Hadoop cluster; it can autonomously
-download data from Amazon S3, FTP and HTTP servers over the
-Internet~\cite{Langmead:2009kx}. However, this preprocessor does not parallelize
-its operations. In theory one can manually introduce some degree of
-parallelization by manually splitting the original input read data into smaller
-files and to explicitly configuring Crossbow's preprocessor to download each one
-(this is done by listing all of them in a special \textit{manifest file}).
-
-%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
-We reimplemented the main functionality in the Crossbow preprocessor as an MR
-program to benefit from its massively parallel nature. The new MR preprocessor
-was implemented in Python using the Hadoop streaming library.
-%\Fixme{Consider deleting the following implementation detail.}
-%To route both reads of the same pair to the same Reducer, a secondary key-sorting mechanism was employed.
-The resulting program efficiently processes short reads produced by different
-versions of the Illumina sequencing platform as well as FASTQ files generated
-from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
-Hadoop-based approach imposes the requirement that the input data is stored
-where the Hadoop nodes can access it, which is in our opinion does not bias the comparison since
-sequencing platforms in the general case have direct access to storage and
-computing facilities.
-
-\subsection*{Computational resources}
-To run the HPC analysis pipeline we used a computational cluster at UPPMAX,
-which is heavily used for analysis of NGS-data~\cite{lampa}.
-Data and reference genomes were stored on a parallel, shared storage system.
-On the other hand, the Hadoop test platform was deployed on a private cloud at
-UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was
-configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}.
-For the fully detailed hardware configuration of the two computational resources
-please see the Supplementary Material section.
-
-%%%%%%%%%%%%
-% Results
-%%%%%%%%%%%%
-% Results and Discussion can be combined.
-\section*{Results}
-
-%As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
-%(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
-%with a batch processing system (hereafter referred to as the HPC method)
-%(Methods).
-
-We executed the two pipelines for each of the five 
-datasets on the HPC and Hadoop platforms, and measured the wall-clock run time for each pipeline stage.  
-All experiments were repeated at least three times, 
-and the results averaged to obtain a data point for the
-particular combination of data size and computational platform.
-
-\subsection*{Accuracy of Pipelines}
-Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
-SOAPsnp) we could not expect them to deliver perfectly
-matching SNP lists. Instead, we verified the accuracy of the test pipelines by
-verifying their ability to correctly identify the mutation $C\rightarrow T$ on
-chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.
-
-\subsection*{Performance evaluation of MR-based preprocessing stage}
-
-To quantify the processing time saved by adopting the MR approach for
-preprocessing the read data, we ran an
-experiment comparing the new preprocessor described in
-Section~{\it MapReduce-based Data Preprocessor} to the original preprocessor furnished with
-Crossbow.  To further validate our strategy, we also compared the performance of
-our MapReduce approach to a different parallelization strategy based on a custom Bash script
-that uses all processing cores available on a single computing node to accelerate the operation (see
-Supplementary Material for details of the implementation). This program read its
-input directly from conventional HPC storage, rather than HDFS\@.  The experiment
-was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
-with 56 cores (7 slave nodes) while the Bash approach ran on a single node
-equipped with 16 CPU cores.
-
-%\footnote{other options like splittable LZO are also possible}
-
-Table~\ref{table:preprocess} shows the results of the experiment, and we see that all 
-of of the parallel approaches are much faster than the original tool, for
-all dataset sizes.  Also, we noticed that the difference between the Bash and Hadoop
-approaches is minimal when running on Dataset II\@.  Indeed, with the relatively
-small dataset size the overhead incurred by the Hadoop framework likely
-overwhelms the useful compute time used for each available CPU core. However,
-with increasing input size the superior scalability of MR becomes apparent, with
-the largest dataset requiring less than half the time of the Bash
-approach to be processed. At the same time, it is important to also notice that
-the Hadoop approach employed approx.\ 1.6 times more CPU time compared to the
-Bash approach.
-
-%ran on a single HPC node, utilizes multiple cores, with the limiting factor
-%being \sout{the hard disk IO performance.}\COM{More correctly to say, that the
-%pipe is not filled completely. i.e. the degree of achieved parallelization does
-%not utilize all 16 cores. Perhaps not the best bash commands etc.} , which
-%effectively leaves with 5-7 cores of 16. \COMMENT{awkward}
-
-\subsection*{Scalability of HPC and Hadoop approaches}
-Given the wide range in genome dataset sizes processed in practice -- e.g.
-exomes and whole-genome sequencing at various depths -- it is
-important to understand the scaling characteristics of the Hadoop and HPC processing
-platforms with respect to input size. For each platform, we
-measured the time required to execute their respective variant calling
-pipeline for each of the test datasets (I-V). The results are
-shown in Table~\ref{table:pipleline-timings}.
-
-The measurements highlight that one of the advantages of using the Hadoop framework is
-its almost linear scalability with respect to input size -- i.e.\ the
-calculation time linearly depends on the size of the
-dataset, regardless the scaling nature of the underlying program~\cite{Langmead:2009kx,Pireddu:2011vn}.
-Figure~\ref{fig:fig1} shows the wall clock
-running time as a function of the dataset size for the 56-core Hadoop cluster and datasets
-I-IV.
-%To stress the linear nature of scaling, both sets of points were fit using the
-%least squares method.
-
-To investigate the scalability characteristics of the HPC approach we ran the
-HPC pipeline using Dataset~I as input and collected the running times as a
-function of the number of cores used (Table~\ref{table:timings-hpc}). The results show that the
-alignment process with Bowtie scales fairly well, but overall the pipeline
-suffers because of the SNP calling step with Samtools. Indeed, this operation
-is the bottleneck of the workflow as it scales
-worse than Bowtie and consumes a progressively larger portion of the total
-calculation time as the input grows -- even after parallelizing analysis by
-chromosome as previously suggested~\cite{biostars_samtools}.
-
-%\footnote{\url{http://www.biostars.org/p/48781} \COM{here was a footnote
-%referring to biostars site. Ok, if the footnotes are not welcome, lets then
-%have it as a reference. That code is not trivial. Maybe we can leave the ref?}
-%\footnote{The amount of available memory, specified with the \texttt{-m}
-%option, is of big importance for the Samtools performance. During its sorting
-%phase the lack of RAM results in spilling to the HDD. For example for an half
-%of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and
-%15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our
-%tasks we used the default Samtool's settings.}.
-
-\subsection*{Efficiency of calculations}
-
-Since Hadoop was designed to digest huge datasets\cite{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach.
-In order to compare the 'suitability' for the Hadoop and HPC platforms run on different number of cores, we construct the function showing the amount of used core-hours:
-$F=T_{p}\times p,$ where~$T_{p}$ is the calculation time on~$p$ cores.
-
-Using the data from Table~\ref{table:pipleline-timings}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}).
-Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset.
-We note that the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases.
-The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on 'bare metal' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
-
-The simplest curve to fit the points nicely is a parabolic function, importantly not a constant. 
-Since the Hadoop approach for the datasets reveals the linear scaling (Figure~\ref{fig:fig1}), a deviation of the ratio from the constant shows the 
-lack of scalability of the HPC approach.
-%Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
-%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e.
-being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}.
-\COMMENT{Why jump back and forth in the discussion of figures 1 and 2?}
-We suspect that the main reason for such a behavior is that for large datasets
-(Dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory-
-and time-demanding than the alignment with Bowtie.  Since Samtools memory
-requirements scale with input size, the application swaps
-data to disk for large input data creating multiple (up to hundreds) temporary files while it sorts
-the BAM file. These files can end up on shared network-accessed
-file systems and thus generate large amounts of network traffic.
-Table~\ref{table:samtools-sort} shows the time required by Samtools to sort dataset~IV with
-different amount of RAM available on a single HPC node. Though there is a large
-deviation when using small amounts of RAM -- we believe due to other users' load
-on the shared cluster network -- the trend clearly indicates that a relatively
-large amount of RAM per node is needed in order to use Samtools effectively. In
-contrast, our tests Hadoop performed the SNP calling on much more economic
-hardware in linear time with just 12GB of RAM per virtual node.
-
-%However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
-%infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
-
-\subsection*{Efficiency of network communication}
-
-The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages.
-The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance degradation.
-To study this phenomenon, we compared HPC and Hadoop network communication costs
-as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).
- Since the alignment process is embarrassingly parallel~-- the read-pairs are
-independent of each other and thus can be aligned independently~-- it is
-possible to distribute the work over more than one computational resource, e.g. by
-splitting the initial data into chunks to processing them in parallel. In this
-approach, reducing the size of each data chunk reduces the size of the aligner
-job,~$T_{alignment}$, but at the same time more chunks have to be sent over the network increasing the communication time,~$T_{comm}$.
-
-There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
-We noted however that the Pmap package functions poorly for datasets larger than 20 Gbases, and instead implemented a highly optimized Bash script for alignment based on standard Unix utilities that makes very efficient use of the HPC
-cluster network (source code available on Github~\cite{code_repo_bash}). We then compared the network
-performance with the standard Hadoop HDFS approach. We separated alignment
-time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
-the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
-(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
-however $T_{comm}$ has a different explanation: For Hadoop, the short reads in
-FASTQ format have to be preprocessed
-to be able to run in MR-fashion, while the data locality will be automatically
-achieved during the data ingestion into HDFS. For the HPC approach,~$T_{comm}$ involves the chunks that are
-sent from the sequence delivery location to the local node scratch disks where
-the actual alignment happens, and the transfer of the aligned SAM files back to
-the delivery location over the network.
-%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}
-
-%The HPC approach is presented in two versions that are based on a bit different strategies of the resource allocation.
-
-\textit{Hadoop results (filled circles in Figure~\ref{fig:fig3})}: The ratio~$T_{alignment}/T_{comm}$ indicates weak dependency in a wide range of number of nodes~$N$: from 4 up to 40.
-It is known that Bowtie provides almost linear scaling between alignment time and dataset chunk size~$D$:~$T_{alignment}\propto  D\propto 1/N$, see~\cite{Langmead:2009uq}, and our experiments summarized in Figure~\ref{fig:fig1}.
-Since the ratio~$T_{alignment}/T_{comm}$ is approximately constant, one can conclude that $T_{comm}\propto 1/N$, meaning that the more nodes involved, the faster the communication is in the preprocessing stage.
-
-%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
-\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@.
-One can see two distinct linear stretches.
-One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60.
-The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed.
-The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
-%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.}
-and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency in the selected coordinates, which can be observed in the plot.
-The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec.
-Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch.
-The nodes are being allocated on the same rack, which is the default SLURM behavior.
-
-Scalability can be improved by overriding the default behavior of SLURM and allocating the nodes not from the same rack, but randomly from all available racks (``HPC random'', open squares in Figure~\ref{fig:fig3}).
-Allocating the nodes on random racks allows for engaging more nodes without network saturation.
-For our cluster we could go up to 30-35 nodes with perfect linear scaling.
-For the most resources used (58 nodes) the deviation from a linear speedup is~$\approx 7\%$ i.e.
-$5.50$ minutes against the ideal $5.14$, see Table~\ref{table:timings-algn}.
-The threshold number of nodes in this strategy~($\approx35$) is because of the uplink cable with the throughput of~50Gbit/sec being saturated.
-The proposed HPC strategies aimed at getting the maximum performance from the storage resources show that while even properly adjusted and tuned, the HPC approaches suffer from the network saturation at higher number of nodes.
-HDFS on the other hand maintains data locality, reducing communication and data transfer and hence lead to better scalability.
-Our Hadoop cluster has no more free nodes to continue to investigate the scaling as in plot at Figure~\ref{fig:fig3}, but we do not expect any significant deviations, since the observed behavior is a generic for Hadoop with HDFS\cite{lin2010, Hadoop:Guide}.
-
-\subsection*{Usability aspects}
-\label{subsectionIV_2}
-
-Hadoop presents a computing paradigm that is different from what most users are
-accustomed to using.  Therefore, adopting it requires an investment in learning
-how it works, how to use it, and how to understand what the programs running on
-the platform are doing.
-A popular way to define and run bioinformatics pipelines on HPC and cloud resources
-is via the Galaxy\cite{galaxy,Afgan:2010uq} platform, which provides a Web-based graphical
-user interface~(GUI) to bioinformatic programs, simplifying the experience for
-the end user.
-A similar project is Cloudgene\cite{cloudgene}, which provides a GUI to help lower the learning curve
-for Hadoop adopters and is specifically targeted at users in the life
-sciences. It consists of a light-weight and flexible Web-based solution for both
-public and private clouds. We implemented our Hadoop-pipeline in Cloudgene and
-extended the platform with functions to import data from the central file system
-at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).  For our
-particular task related to DNA sequencing, Cloudgene makes it easy to select
-data and parameters for its execution. Most of the data management work is done
-automatically and the results can be downloaded to the client machine. The
-modular structure allows modification of the source code to adapt to the
-existing computing center architecture. For example, UPPMAX users can import
-their data from the sequencing platform directly to the Hadoop cluster by
-pressing a button and entering the credentials; at the same time they can be
-sure that their sensitive data is kept private.
-
-%%%%%%%%%%%%
-% Discussion
-%%%%%%%%%%%%
-\section*{Discussion}
-
-In this report we have described two approaches for high-performance analysis of
-DNA sequencing data; one based on regular HPC using a batch system and one based
-on Hadoop. In order to compare the two platforms we developed highly optimized pipelines
-for both scenarios and found that Hadoop becomes comparable with HPC when processing
-datasets larger than 100\,Gbases.
-As this corresponds to typical human genome data with a sequencing depth of about~$30x$, we consider Hadoop not only to be a technology for the future but an attractive option for bioinformatics data analysis already today.
-While implementations on HPC are more efficient on the use of CPU time than the Hadoop-based approach, the price paid for this is increased overall runtime.
-The main benefit of Hadoop is however its almost linear scaling with the problem size, and we show that this also holds for analysis of NGS data.
-Our results further reveal that the calculations on Hadoop with the file system HDFS scales
-better than the network attached parallel storage commonly used in the HPC centers.  
-Finally, we demonstrated that an existing and extensible
-publicly available web-based GUI (Cloudgene) provides an easy way to execute
-bioinformatics analysis on Hadoop for those who are less experienced with Linux
-scripting.
-
-Hadoop has so far seen relatively low adoption in bioinformatics.
-This is likely due to that existing high-performance infrastructure in most cases are built as traditional HPC systems with batch processing queues.
-Also, the bioinformatics software flora is still not as big for Hadoop and most analysis tools are implemented as regular linux applications.
-This may very well change soon with the demonstrated scalability and wide uptake of Hadoop in other settings.
-The results in this paper show that Hadoop enables highly scalable massively parallel analysis of DNA sequence data, and as the size of data sets are expected to increase over time we envision that Hadoop will become increasingly popular also in bioinformatics to deliver results faster.
-
-%%%%%%%%%%%
-% Acknowledgments
-%%%%%%%%%%%
-% Do NOT remove this, even if you are not including acknowledgments
-\section*{Acknowledgments}
-The computations were performed on resources provided by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) under project p2013023.
-This work was supported by the Swedish strategic research programme eSSENCE and COST Action BM1006 Next Generation Sequencing Data Analysis Network, SeqAhead.
-We thank system experts Pontus Freyhult and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage, Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.
-Part of L.P.'s activity was performed within the context of the Ph.D.\ program in
-Biomedical Engineering at the University of Cagliari, Italy.
-
-%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%
-%\section*{References}
-% The bibtex filename
-\bibliography{paper-plos}
-
-%%%%%%%%%%%
-% Supplementary Material
-%%%%%%%%%%%
-\section*{Supplementary Material}
+\title{Supplemental information}
+\author{}
+\date{}
+\maketitle
 
@@
 HPC:
-Multinode short-read alignment was performed on the Milou cluster\cite{milouCluster}, equipped with dual 8-core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
+Multinode short-read alignment was performed on the Milou cluster (\url{http://www.uppmax.uu.se/the-milou-cluster}), equipped with dual 8-core Intel Xeon E5-2660, (2.2 GHz, 2\,MB L2 cache, 20\,MB L3 cache), 128\,GB of RAM, Infiniband node-to-node network connection, and 10Gbit/s uplink.
 
 \item Storage: 
-Gulo\cite{gulo} is a custom built Lustre 2.4 system using 8 HP nodes with MDS600
+Gulo (\url{http://www.uppmax.uu.se/gulo}) is a custom built Lustre 2.4 system using 8 HP nodes with MDS600
 storage boxes and an additional node for metadata handling. In total, it
@@
 \item Our Hadoop test platform was deployed on a private cloud at UPPMAX using
-	the OpenNebula~\cite{opennebula} cloud management system. Each physical node
+	the OpenNebula (\url{http://opennebula.org}) cloud management system. Each physical node
 	was equipped with two 4-core Intel Xeon 5420 (2.50\,GHz; 12~MB L2 cache),
 	16~GB RAM, one 1~TB SATA disk and Gigabit Ethernet. The cluster was set up
-	with Cloudera Hadoop Distribution version~2.0.0-cdh4.5.0~\cite{cloudera}.
+	with Cloudera Hadoop Distribution version~2.0.0-cdh4.5.0 (\url{http://www.cloudera.com/content/cloudera/en/why-cloudera/hadoop-and-big-data.html}).
 	Note that the physical hardware provided less RAM than desired. Each VM node
@@
 
-%%%%%%%%%%%
-% Figure Legends
-%%%%%%%%%%%
-\section*{Figure Legends}
-
-\begin{figure}[!ht]
-	\input{./figures/fig1/fig1.tex}
-	\caption{%
-		Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
-		increasing input dataset size. Each test dataset (labelled with Roman
-		numerals) was used. The image clearly highlights the linear scaling
-		characteristics of the Hadoop-based approach.
-		}
-	\label{fig:fig1}
-\end{figure}
-
-\begin{figure}[!ht]
-	\input{./figures/fig2/fig2.tex}
-	\caption{%
-	The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
-	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
-	a 16-core HPC node, respectively.
-	The points are fit to a quadratic least-squares curve, which makes it possible
-	to predict {\it infinite} dataset size. Datasets are labeled with Roman
-	numerals.
-	}
-	\label{fig:fig2}
-\end{figure}
-
-\begin{figure}[!ht]
-	\small
-	\input{./figures/fig9/fig3.tex}
-	\normalsize
-\caption{%
-	Ratios of alignment time~$T_{alignment}$ to communication costs~$T_{comm}$ for
-	the HPC and Hadoop clusters as a function of reciprocal number of nodes~$1/N$.
-	All runs used Dataset~IV as input. Two HPC scenarios are shown as ``HPC
-	SLURM'' and ``HPC random'', which correspond to standard SLURM behavior and a
-	modified one that allocates nodes from random racks.  Linear fit to $ax+b$ was
-	done with the least-squares method.  The estimated coefficients for the
-	Hadoop-based approach are $(a,b)\approx(0.95,2.07)$.  One can see two defined
-	linear regions for the HPC approach with very different tangents for linear
-	and non-linear scalings correspondingly.  For  the ``HPC SLURM'' $(a,b)\approx
-	(0.96, 3.49)$ and $(a,b)\approx (32.99, 0.60)$, and for  the ``HPC random''
-	$(a,b)\approx (0.88, 3.72)$ and $(a,b)\approx (98.24, 0.53)$.  The constant
-	term for HPC goes below unity -- i.e.\ for a large number of nodes more time
-	will be spent on communication than the actual alignment. 
-	}
-
-%The first one resembles to that of Hadoop, and corresponds to the case then the communication in the cluster is not a bottleneck. 
-%The second one shows the regime corresponding to the case of saturation of the network, resulting in traffic jams: data distibution time~$T_{comm}$ stops to scale 
-%linearly to the reciprocal cluster size, and becomes more of less a constant, while the alignment time continues to satisfy~$T_{alignment}\propto 1/N$.
-	\label{fig:fig3}
-\end{figure}
-
-\begin{figure}[!ht]
- 	\begin{subfigure}[b]{0.4\textwidth}
-        	\includegraphics[width=\textwidth]{a.png}
-		\subcaption{}
-	\end{subfigure}
-	\hspace{0.5cm}
-	\begin{subfigure}[b]{0.6\textwidth}
-	        	\includegraphics[width=\textwidth]{c.png}
-		\subcaption{}
-	\end{subfigure}
-%	\begin{subfigure}[b]{0.6\textwidth}
-%		\includegraphics[width=\textwidth]{b.png}
-%		\subcaption{Cloudgene: specifying job parameters}
-%	\end{subfigure}
-	\caption{%
-	An example of a job setup view with the graphical Hadoop front-end Cloudgene,
-	providing a smooth user experience even for novice users. \textbf{a)} Pipeline
-	selection -- in our case containing the Crossbow pipeline. \textbf{b)} The
-	UPPMAX-adapted functionality to browsing and import data from the user's home
-	folder in the shared file system.
-	}
-	\label{fig:fig4}
-\end{figure}
-
-%%%%%%%%%%%
-% Tables
-%%%%%%%%%%%
-\section*{Tables}
-
-\begin{table}[!ht]
-\small
-\footnotesize
-\caption{Datasets used in the comparison. }
-\begin{center}
-\begin{tabular}{|l|l|l|}
-Dataset &	Organism &	Size in Gbases\\
-\hline
- I		&	{\it A.thaliana}	&	1.4	\\
- II	&	{\it A.thaliana}	&	7.0\\
- III	&	{\it A.thaliana}	&	30.0	\\
- IV	&{\it A.thaliana}, the artificial dataset created using Samtools package	&	100.0	\\
- V	&	{\it H.sapiens}, two individuals (GM12750 and GM12004), sample SRR499924		&	250.0\\
-
-\end{tabular}
-\end{center}
-\label{table:datasets}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\small
-\caption{%
-	Time (in minutes) employed to preprocess Datasets II, III, and IV by
-	Crossbow's native read preprocessor and our MR preprocessor on a 56-core Hadoop
-	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
-	}
-
-\begin{center}
-\begin{tabular}{l|c|c|c}
-Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
-\hline
-Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673.0\pm1.0$	\\
-Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
-Bash, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
-\end{tabular}
-\end{center}
-\label{table:preprocess}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\small
-
-\caption{%
-	Time (in minutes) for processing different dataset sizes on the HPC and
-	Hadoop deployments with their respective pipelines. Datasets are labelled by
-	Roman numerals, (``f.r.'' stands for ``forward reads''). The large variance
-	for the HPC deployment is due to Samtools BAM sorting, which is I/O and memory
-	intensive; the data spillage to the file system makes it very susceptible to
-	the load of the entire shared HPC cluster.
-	}
-
-\begin{center}
-\begin{tabular}{l|c|c|c|c|c|c|c}
-
-Data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
-				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
-\hline
-Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
-HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\
-
-\end{tabular}
-\end{center}
-\label{table:pipleline-timings}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\small
-\caption{Timings for Dataset I processed on different number of cores of a HPC node.}
-\begin{center}
-\begin{tabular}{c|c|c|c|ccc}
-$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c}{speed-ups} \\
-\hline
-	& alignment 	&	SNP calling	&	total  &total & alignment& SNP call\\
-\hline
-1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
-2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
-4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
-6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
-8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
-10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
-12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
-14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
-16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
-\end{tabular}
-\end{center}
-\label{table:timings-hpc}
-\normalsize
-\end{table}
-
- \begin{table}[hbtp]
-\small
-
-\caption{%
-	Time (in minutes) required by Samtools to sort
-	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
-	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
-	amount of available RAM reported is for all 16 cores.
-}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-RAM, GB		&	8			&		16			&			32		& 		64				\\
-timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
-\end{tabular}
-\end{center}
-\label{table:samtools-sort}
-\normalsize
-\end{table}%
-
-\begin{table}[!ht]
-\caption{Timings  for alignment and the ratio~$T_{alignment}/T_{comm}$ for HPC and Hadoop clusters for Dataset IV.
-For the ``HPC random'' approach, data chunks have to be copied to the local scratch disks first and the alignments (SAM files) copied back while Hadoop keeps all the data inside HDFS and hence does not need data staging.
-Hadoop however needs to preprocess reads before the actual alignment stage in order to be able to operate in MR manner resulting in what we term ``communication costs''.
-Note that each HPC node has 16 cores, while each Hadoop node has 7 (one core is dedicated to run the virtual machine).}
-\begin{center}
-\begin{tabular}{c|c|c|c|c|c}
- \multicolumn{3}{c|}{Hadoop} & \multicolumn{3}{c}{HPC random} \\
- \hline
-
-Number of nodes	&Alignment time,	&$\frac{T_{alignment}}{T_{comm}}$	&Number of nodes	&Alignment time,	&$\frac{T_{alignment}}{T_{comm}}$\\
-(cores)					&minutes		&							&(cores)			&minutes&\\
-\hline
-4(28)	&293.5	&2.33	&4(64)	&74.4	&3.89\\
-6(42)	&189.8	&2.19	&10(160)	&32.4	&3.76\\
-8(56)	&136.0	&2.23	&14(224)	&22.7	&3.77\\
-16(112)	&70.3	&1.96	&18(288)	&17.9	&3.78\\
-32(224)	&39.3	&2.20	&22(352)	&14.5	&3.79\\
-40(280)	&32.5	&2.15	&26(416)	&12.3	&3.77\\
-			&&&30(480)	&10.7	&3.73\\
-			&&&34(544)	&9.5	&3.45\\
-			&&&38(608)	&8.5	&3.16\\
-			&&&42(672)	&7.6	&2.96\\
-			&&&46(736)	&7.0	&2.55\\
-			&&&50(800)	&6.4	&2.65\\
-			&&&54(864)	&5.9	&2.34\\
-			&&&58(928)	&5.5	&2.12\\
-
-\end{tabular}
-\end{center}
-\label{table:timings-algn}
-\end{table}%
-
 \end{document}

### 2014-05-28 rebuttal.txt
Create rebuttal file

----------------------------------------
Reviewer: 1

Comments to the Author
The authors designed an experiment to test the performance of Hadoop-based aligner, crossbow, for short sequence alignment. The wall-time, scalability and computing efficiency metrics were reported. Based on the results from their specific Hadoop configuration the authors concluded that Hadoop based solution becomes more appealing when data sets get larger (>100GB).

General Comments:

Despite that this work is targeting a very interesting problem, in my opinion it is rather preliminary. Short read alignment problem is just one of many existing bioinformatics problems, do the readers expect the conclusions similarly apply to other problems? The results were obtained from a single Hadoop implementation, this work may benefit from testing results from several distinct Hadoop implementations.  Also Hadoop does have its own limitations, and this was not adequately discussed.

Reviewer: 2

Comments to the Author
The authors present evidence as to when a Hadoop-based approach becomes more effective than (that is, becomes "an appealing alternative" to) current standard HPC parallelization. They do so in the context of in a common and important bioinformatics problem: alignment of NGS short reads to a reference genome, and then searching for SNPs.

They performed their testing using open source software, as well as using custom programs that they have made available on Github, and whose locations are given.

I believe the authors have conducted valid tests, and have made a quite credible case for their conclusions. Of course, the tests were conducted on different hardware, using different software, for only one type of workflow. However, the limitations are clearly stated, so readers can draw their own conclusions.

Also, in general, I find the paper to be well-written and up to journal standards. However, I would like to see some clarifications in the paper in these aspects:

1) The authors conduct two tests. The first compares the Hadoop approach across a cluster to multi-core HPC on one node, and the mapping and SNP dicovery results are reported out in terms of wall clock time and a comparison of core-hours required on the different data sets. The second compares the Hadoop approach against a parallelized Bash script running on multiple nodes, for only alignment (not adding SNP searches), for examining efficiency of network communication and how such relates to scalability.

The reader discovers the switch in hardware and workflow only in details of each section. I consider this a drawback to readability. It would be good to have a short paragraph at the start saying: two types of test, here's why, here's what we will get from each test.

2) The Hadoop hardware the authors had available had only 2 GB RAM per core. The authors note that this is a limitation on the effectiveness of the Hadoop approach - they expect "much better performance with twice of much memory". Well - yes, indeed. Unfortunately, I can only find this note in the Supplement. This information needs to be also stated clearly in the main article. Readers should be aware that in a properly provisioned Hadoop cluster, the break-even input data size might be even smaller (even more in Hadoop's favor).

3) In the second test (on efficiency of network communication), it is not quite clear to me as to of what T-communication time T-comm consists. I think the authors mean that T-comm in the Hadoop approach is the time for preprocessing the data, while in the standard HPC approach (using their optimized Bash script) T-comm is the time for sending the data chunks to the local node scratch disk, and then sending the aligned SAM files back to the delivery location. But - I am not sure, and a reader should not have to guess. This needs to be made explicit. Also: I can find no info  as to *how* T-comm is measured. I would like hear how the authors separate out the T-comm time from T-alignment time.

4) Re figure 1(b): I understand that the authors are graphing the ratio of core-hours needed in Hadoop over core-hours needed in HPC, versus the inverse size of the input data set. So in order to show the advantage of Hadoop use, the ratio should be as small as possible, and we see that as the inverse data set size approaches zero in the figure. The figure is informative, once you work through it. However, it took me a couple minutes to wrap my head around the figure, and so I want to ask: is there a simpler way to present this, in terms of, say, an x-axis that simply uses increasing data set size rather than the more complicated inverse data set size? I know the authors were doing least squares fitting to a curve, but just want them to think about it.

5) p. 3 - In the "accuracy of pipelines" the authors state their criterion for accuracy. However, they never *explicitly* say that the criterion has been met - the reader has to infer it. Can we get a sentence simply stating: "Both pipelines were able to make the correct identification."

6) p.4 - in the first sentence on the page, we get "To further validate our strategy, we also compared ...". I think this is somewhat misleading. What is about to be described is not "furthering" the main comparison - it *is* the main comparison. The only process done previous is preprocessing set in the MR (Hadoop) approach, which tells you nothing in isolation. So - I think this should be reworded.
----------------------------------------
